---
title: "Join interesting figures from microbiome analyses"
author: "Paula Rodriguez-Garcia"
date: "`r Sys.time()`"
output:
  html_document:
    output_dir: "html"
params: 
  markdown_id: "190"
  methane_group: "Methane_group"


---

# Setup

## Knit options
```{r setup}
knitr::opts_chunk$set(warning = FALSE, message = FALSE,
                       root.dir = rprojroot::find_rstudio_root_file())
```

## Libraries
```{r}
library(tibble)
library(dplyr)
library(magrittr)
library(tidyr)
library(readxl)
library(glue)
library(purrr)
library(stringr)
library(ggpubr)
library(openxlsx)
library(flextable)
library(officer)
library(gtsummary)
```


## Study paramters
```{r}
if (file.exists("../../study_parameters.R")) {
  source("../../study_parameters.R")
} else {
  source("study_parameters.R")
}

paulaR::show_rmd_params(params)
```


## Directories
```{r}
```

## Functions
```{r}
ft_theme <-  function(ft,
                             font_size = 10,
                             header_fill = "gray85") {

  ft |>
    font(fontname = "Times New Roman", part = "all") |>
    fontsize(size = font_size, part = "all") |>
    
    color(color = "black", part = "all") |>
    
    bg(bg = "white", part = "body") |>
    bg(bg = header_fill, part = "header") |>
    
    bold(part = "header") |>
    color(color = "black", part = "header") |>

    border_remove() |>

    hline_top(
      border = fp_border(color = "black", width = 1.2),
      part = "header"
    ) |>

    hline(
      i = 1,
      border = fp_border(color = "black", width = 0.8),
      part = "header"
    ) |>

    hline_bottom(
      border = fp_border(color = "black", width = 1.2),
      part = "body"
    ) |>

    autofit()
}
```


# Read tables
```{r}
bsl_char_raw <- readRDS(glue("{rds_dir}/111_baseline_table_means.rds"))

ph_gc_raw <- readRDS(glue("{rds_dir}/123_ph_methane_group_wilcoxon.rds"))
tt_gc_raw <- readRDS(glue("{rds_dir}/123_tt_methane_group_wilcoxon.rds"))

gut_gc_raw <- readRDS(glue("{rds_dir}/121_gut_factors_methane_group_tt_linear_models.rds"))

alpha_gc_raw <- readRDS(glue("{rds_dir}/141_alpha_div_methane_tt_linear_models.rds"))
alpha_cor_raw <- readRDS(glue("{rds_dir}/141_alpha_div_correlation_linear_models.rds"))

ub_gc_raw <- readRDS(glue("{rds_dir}/142_genus_counts_ub_methane_group_tt_linear_models.rds"))

scfa_gc_raw <- readRDS(glue("{rds_dir}/171_dry_long_scfa_methane_group_comparison.rds"))
scfa_cor_raw <- readRDS(glue("{rds_dir}/171_dry_long_scfa_correlation.rds"))

diet_gc_raw <- readRDS(glue("{rds_dir}/161_means_diet_methane_group_comparison.rds"))
```

# Clean tables

## 1) BSL characteristics
```{r}
bsl_char <- bsl_char_raw %>%
  as_flex_table() %>% 
  ft_theme() %>% 
  autofit()

bsl_char
```

## 2) pH group comp (wilcox)
```{r}
ph_gc <- ph_gc_raw %>% 
 mutate(across(where(is.numeric), round, 3)) %>% 
  mutate(hmp_mean_sd = str_c(
    mean_HMP, " (", sd_HMP, ")",
    sep = "")) %>% 
    mutate(lmp_mean_sd = str_c(
    mean_LMP, " (", sd_LMP, ")",
    sep = "")) %>% 
  mutate(" *" = case_when(
    p<0.05 & FDR<0.1 ~ "●",
    p<0.05  ~ "*",
    TRUE ~ "")) %>% 
  select("Segment" = "pH segment",
         "pH HMP, mean (SD)" = hmp_mean_sd,
         "pH LMP, mean (SD)" = lmp_mean_sd,
         "p" = "p",
         "q" = "FDR",
         " *") %>% 
  flextable() %>% 
  ft_theme() %>% 
  autofit()

ph_gc
```

## 3) TT group comp (wilcox)
```{r}
tt_gc <- tt_gc_raw %>% 
 mutate(across(where(is.numeric), round, 3)) %>% 
  mutate(hmp_mean_sd = str_c(
    mean_HMP, " (", sd_HMP, ")",
    sep = "")) %>% 
    mutate(lmp_mean_sd = str_c(
    mean_LMP, " (", sd_LMP, ")",
    sep = "")) %>% 
  mutate(" *" = case_when(
    p<0.05 & FDR<0.1 ~ "●",
    p<0.05  ~ "*",
    TRUE ~ "")) %>% 
  select(Variable,
         "Hours HMP, mean (SD)" = hmp_mean_sd,
         "Hours LMP, mean (SD)" = lmp_mean_sd,
         "p" = "p",
         "q" = "FDR",
         " *") %>% 
  flextable() %>% 
  ft_theme() %>% 
  autofit()

tt_gc
```

## 4) Gut group comp (mixed lm) 
```{r}
gut_gc <-
gut_gc_raw %>% 
  filter(adjusted_by %in% c("Subject_ID", "", "WGTT", "WGTT, Subject_ID")) %>% 
  mutate(response = factor(response, levels = gut_gc_raw$response %>% unique())) %>% 
  mutate(adjusted_by = factor(adjusted_by, levels = c(
    "",  "Subject_ID", "WGTT", "WGTT, Subject_ID"
  ))) %>% 
  arrange(adjusted_by) %>% 
  mutate(Test = case_when(
    adjusted_by == "" ~ "CS LM, unadjusted",
    adjusted_by == "Subject_ID" ~ "Long. LMM, adjusted for subject",
    adjusted_by == "WGTT" ~ "CS LM, adjusted for WGTT",
    adjusted_by == "WGTT, Subject_ID" ~ "Long. LMM, adjusted for WGTT and subject"
  )) %>% 
  mutate(" *" = case_when(
    p<0.05 & FDR<0.1 ~ "●",
    p<0.05  ~ "*",
    TRUE ~ "")) %>% 
  select("Variable" = response,
         "Test",
         "STD β" = estimate, # here the raw estimate is already STD
         "p" = p,
         "q" = FDR,
         " *") %>% 
   mutate(across(where(is.numeric), round, 3)) %>% 
  flextable() %>% 
  ft_theme() %>% 
  # to separate the fdr adjustment
  #hline(i = 11, border = fp_border(color = "black", width = 1)) %>% 
  autofit() 

gut_gc
```

## 5)  Alpha group comp
```{r}
# here Y was also scaled, thus estimate is directly beta std
alpha_gc <-
alpha_gc_raw %>% 
   filter(adjusted_by %in% c("(1|Subject_ID)", "WGTT + (1|Subject_ID)")) %>% 
  mutate(response = str_to_sentence(response)) %>% 
  mutate(adjusted_by = factor(adjusted_by, levels = c(
    "(1|Subject_ID)", "WGTT + (1|Subject_ID)"
  ))) %>% 
  arrange(adjusted_by) %>% 
  mutate(Test = case_when(
    adjusted_by == "(1|Subject_ID)" ~ "Long. LMM, adjusted for subject",
    adjusted_by == "WGTT + (1|Subject_ID)" ~ "Long. LMM, adjusted for WGTT and subject"
  )) %>%
  mutate(" *" = case_when(
    p<0.05 & FDR<0.1 ~ "●",
    p<0.05  ~ "*",
    TRUE ~ "")) %>% 
  select("Variable" = response,
         "Test",
         "STD β" = beta_estimate, # here the raw estimate is already STD
         "p" = p,
         "q" = FDR,
         " *") %>% 
   mutate(#across(where(is.numeric), round, 3),
          p = round(p, 4),
          `q` = round(`q`, 4),
          `STD β` = round(`STD β`, 3)
          ) %>% 
  flextable() %>% 
  ft_theme() %>% 
  # to separate the fdr adjustment
 hline(i = 2, border = fp_border(color = "black", width = 1)) %>% 
  autofit() 

alpha_gc
```

## 6) Alpha corr
```{r}
# here Y was also scaled, thus estimate is directly beta std
alpha_cor <-
alpha_cor_raw %>% 
   filter(adjusted_by %in% c("", "(1|Subject_ID)")) %>% 
  mutate(response = str_to_sentence(response)) %>% 
  mutate(predictor = case_when(
    predictor == "fasting_CH4" ~ "Fasting breath methane",
    predictor == "fasting_H2" ~ "Fasting breath hydrogen",
    predictor == "smithii_inv_ct" ~ "M. smithii",
    predictor == "archaea_inv_ct" ~ "Archaea",
    predictor == "bacterial_load_wet" ~ "Microbial load",
    predictor == "stool_energy" ~ "Energy density",
    predictor == "ph_proximal_colon" ~ "Proximal colonic pH",
  )) %>% 
  mutate(predictor = factor(predictor, levels = c(
    "Fasting breath methane",
    "Fasting breath hydrogen",
    "M. smithii",
    "Archaea",
    "Microbial load",
    "Energy density",
    "Proximal colonic pH"
  ))) %>% 
  mutate(adjusted_by = factor(adjusted_by, levels = c(
   "",  "(1|Subject_ID)"
  ))) %>% 
  arrange(predictor) %>% 
  mutate(Test = case_when(
    adjusted_by == "" ~ "CS LM, unadjusted",
    adjusted_by == "(1|Subject_ID)" ~ "Long. LMM, adjusted for subject",
  )) %>% 
  # We need to do FDR again
  group_by(Test) %>% 
  mutate(FDR = p.adjust(p, "BH")) %>% 
  ungroup() %>%
  mutate(" *" = case_when(
    p<0.05 & FDR<0.1 ~ "●",
    p<0.05  ~ "*",
    TRUE ~ "")) %>% 
  select("Predictor" = predictor,
         "Response" = response,
         "Test",
         "STD β" = estimate, # remember that here the raw estimate is already STD
         "p" = p,
         "q" = FDR,
         " *") %>% 
   mutate(#across(where(is.numeric), round, 3),
          p = round(p, 4),
          `q` = round(`q`, 4),
          `STD β` = round(`STD β`, 3)
          ) %>% 
  flextable() %>% 
  ft_theme() %>% 
  autofit() 

alpha_cor
```

## 7) Microbiome group comp
```{r}
ub_gc <- 
ub_gc_raw %>% 
  filter(adjusted_by %in% c("(1|Subject_ID)", "WGTT + (1|Subject_ID)")) %>% 
  mutate(adjusted_by = factor(adjusted_by, levels = c(
    "(1|Subject_ID)", "WGTT + (1|Subject_ID)"
  ))) %>% 
  arrange(adjusted_by) %>% 
  mutate(Test = case_when(
    adjusted_by == "(1|Subject_ID)" ~ "Long. LMM, adjusted for subject",
    adjusted_by == "WGTT + (1|Subject_ID)" ~ "Long. LMM, adjusted for WGTT and subject"
  )) %>% 
  mutate(" *" = case_when(
    p<0.05 & FDR<0.1 ~ "●",
    p<0.05  ~ "*",
    TRUE ~ "")) %>% 
  select("Variable" = response,
         "Test",
         "STD β" = estimate_std,
         "p" = p,
         "q" = FDR, 
         " *") %>% 
   mutate(across(where(is.numeric), round, 3),
          #P = round(P, 4),
          #`FDR-adjusted q` = round(`FDR-adjusted q`, 4),
          #`STD β` = round(`STD β`, 3)
          ) %>% 
  flextable() %>% 
  ft_theme() %>% 
  # to separate the fdr adjustment
 hline(i = 84, border = fp_border(color = "black", width = 1)) %>% 
  autofit() 

ub_gc
```

## 8) Faecal SCFA group comp
```{r}
f_gc <- 
scfa_gc_raw %>% 
  filter(type_response == "fecal") %>% 
  filter(adjusted_by %in% c(", (1|Subject_ID)", "WGTT, (1|Subject_ID)")) %>% 
  mutate(adjusted_by = factor(adjusted_by, levels = c(
    ", (1|Subject_ID)", "WGTT, (1|Subject_ID)"
  ))) %>%
  arrange(adjusted_by) %>% 
  mutate(Test = case_when(
    adjusted_by == ", (1|Subject_ID)" ~ "Long. LMM, adjusted for subject",
    adjusted_by == "WGTT, (1|Subject_ID)" ~ "Long. LMM, adjusted for WGTT and subject"
  ))  %>% 
  mutate(response = str_remove(response, "f_")) %>% 
  mutate(" *" = case_when(
    p<0.05 & FDR<0.1 ~ "●",
    p<0.05  ~ "*",
    TRUE ~ "")) %>% 
  select("Variable" = response,
         "Test",
         "STD β" = estimate_std,
         "p" = p,
         "q" = FDR, 
         " *") %>% 
   mutate(across(where(is.numeric), round, 3),
          #P = round(P, 4),
          #`FDR-adjusted q` = round(`FDR-adjusted q`, 4),
          #`STD β` = round(`STD β`, 3)
          ) %>% 
  flextable() %>% 
  ft_theme() %>% 
  # to separate the fdr adjustment
 hline(i = 8, border = fp_border(color = "black", width = 1)) %>% 
  autofit() 

f_gc
```

## 9) Plasma SCFA group comp
```{r}
p_gc <- 
scfa_gc_raw %>% 
  filter(type_response == "plasma") %>% 
  filter(adjusted_by %in% c(", (1|Subject_ID)", "WGTT, (1|Subject_ID)")) %>% 
  mutate(adjusted_by = factor(adjusted_by, levels = c(
    ", (1|Subject_ID)", "WGTT, (1|Subject_ID)"
  ))) %>%
  arrange(adjusted_by) %>% 
  mutate(Test = case_when(
    adjusted_by == ", (1|Subject_ID)" ~ "Long. LMM, adjusted for subject",
    adjusted_by == "WGTT, (1|Subject_ID)" ~ "Long. LMM, adjusted for WGTT and subject"
  ))  %>% 
  mutate(response = str_remove(response, "p_")) %>% 
  mutate(" *" = case_when(
    p<0.05 & FDR<0.1 ~ "●",
    p<0.05  ~ "*",
    TRUE ~ "")) %>% 
  select("Variable" = response,
         "Test",
         "STD β" = estimate_std,
         "p" = p,
         "q" = FDR, 
         " *") %>% 
   mutate(across(where(is.numeric), round, 3),
          #P = round(P, 4),
          #`FDR-adjusted q` = round(`FDR-adjusted q`, 4),
          #`STD β` = round(`STD β`, 3)
          ) %>% 
  flextable() %>% 
  ft_theme() %>% 
  # to separate the fdr adjustment
 hline(i = 7, border = fp_border(color = "black", width = 1)) %>% 
  autofit() 

p_gc
```

## 10) Plasma SCFA correlation
```{r}
p_cor <- 
scfa_cor_raw %>% 
  filter(type_response == "plasma",
         type_predictor %in% c("breath", "methanogens", "pH")) %>% 
  mutate(type_predictor = forcats::fct_collapse(
      type_predictor,
      "breath_methanogens" = c("breath", "methanogens")
    )
  )  %>% 
  filter(adjusted_by %in% c("", "(1|Subject_ID)")) %>%
  mutate(adjusted_by = factor(adjusted_by, levels = c(
    "", "(1|Subject_ID)"
  ))) %>%
  arrange(type_predictor) %>% 
  mutate(Test = case_when(
    adjusted_by == "" ~ "CS LM, unadjusted",
    adjusted_by == "(1|Subject_ID)" ~ "Long. LMM, adjusted for subject",
  ))   %>% 
  mutate(predictor = case_when(
    predictor == "fasting_CH4" ~ "Fasting breath methane",
    predictor == "fasting_H2" ~ "Fasting breath hydrogen",
    predictor == "smithii_inv_ct" ~ "M. smithii",
    predictor == "archaea_inv_ct" ~ "Archaea",
    predictor == "ph_proximal_colon" ~ "Proximal colonic pH",
    predictor == "ph_distal_colon" ~ "Distal colonic pH",
    predictor == "ph_sigmoid_colon" ~ "Sigmoid colonic pH",
    TRUE ~ NA
  )) %>% 
  filter(!is.na(predictor)) %>% 
  mutate(predictor = factor(predictor, levels = c(
    "Fasting breath methane",
    "Fasting breath hydrogen",
    "M. smithii",
    "Archaea",
    "Proximal colonic pH",
    "Distal colonic pH",
    "Sigmoid colonic pH"
  ))) %>%
  mutate(response = str_remove(response, "p_")) %>% 
  # redo fdr after filtering
  group_by(type_predictor) %>% 
  mutate(FDR = p.adjust(p, "BH")) %>% 
  ungroup() %>% 
  mutate(" *" = case_when(
    p<0.05 & FDR<0.1 ~ "●",
    p<0.05  ~ "*",
    TRUE ~ "")) %>% 
  select("Predictor" = predictor,
         "Response" = response,
         "Test",
         "STD β" = estimate_std,
         "p" = p,
         "q" = FDR, 
         " *") %>% 
   mutate(across(where(is.numeric), round, 3),
          #P = round(P, 4),
          #`FDR-adjusted q` = round(`FDR-adjusted q`, 4),
          #`STD β` = round(`STD β`, 3)
          ) %>% 
  flextable() %>% 
  ft_theme() %>% 
  # to separate the fdr adjustment
 hline(i = 28, border = fp_border(color = "black", width = 1)) %>% 
  autofit() 

p_cor
```

## 11) Faecal SCFA correlation
```{r}
f_cor <- 
scfa_cor_raw %>% 
  filter(type_response == "fecal",
         type_predictor %in% c("breath", "methanogens", "pH")) %>% 
  mutate(type_predictor = forcats::fct_collapse(
      type_predictor,
      "breath_methanogens" = c("breath", "methanogens")
    )
  )  %>% 
  filter(adjusted_by %in% c("", "(1|Subject_ID)")) %>%
  mutate(adjusted_by = factor(adjusted_by, levels = c(
    "", "(1|Subject_ID)"
  ))) %>%
  arrange(type_predictor) %>% 
  mutate(Test = case_when(
    adjusted_by == "" ~ "CS LM, unadjusted",
    adjusted_by == "(1|Subject_ID)" ~ "Long. LMM, adjusted for subject",
  ))   %>% 
  mutate(predictor = case_when(
    predictor == "fasting_CH4" ~ "Fasting breath methane",
    predictor == "fasting_H2" ~ "Fasting breath hydrogen",
    predictor == "smithii_inv_ct" ~ "M. smithii",
    predictor == "archaea_inv_ct" ~ "Archaea",
    predictor == "ph_proximal_colon" ~ "Proximal colonic pH",
    predictor == "ph_distal_colon" ~ "Distal colonic pH",
    predictor == "ph_sigmoid_colon" ~ "Sigmoid colonic pH",
    TRUE ~ NA
  )) %>% 
  filter(!is.na(predictor)) %>% 
  mutate(predictor = factor(predictor, levels = c(
    "Fasting breath methane",
    "Fasting breath hydrogen",
    "M. smithii",
    "Archaea",
    "Proximal colonic pH",
    "Distal colonic pH",
    "Sigmoid colonic pH"
  ))) %>%
  mutate(response = str_remove(response, "f_")) %>% 
  # redo fdr after filtering
  group_by(type_predictor) %>% 
  mutate(FDR = p.adjust(p, "BH")) %>% 
  ungroup() %>% 
  mutate(" *" = case_when(
    p<0.05 & FDR<0.1 ~ "●",
    p<0.05  ~ "*",
    TRUE ~ "")) %>% 
  select("Predictor" = predictor,
         "Response" = response,
         "Test",
         "STD β" = estimate_std,
         "p" = p,
         "q" = FDR, 
         " *") %>% 
   mutate(across(where(is.numeric), round, 3),
          #P = round(P, 4),
          #`FDR-adjusted q` = round(`FDR-adjusted q`, 4),
          #`STD β` = round(`STD β`, 3)
          ) %>% 
  flextable() %>% 
  ft_theme() %>% 
  # to separate the fdr adjustment
 hline(i = 32, border = fp_border(color = "black", width = 1)) %>% 
  autofit() 

f_cor
```

## 12) Diet group comp
```{r}
diet_gc <-
diet_gc_raw %>% 
  filter(response %in% c(
    "total_kcal",
    "carbohydrate",
    "fat",
    "protein",
    "fish_per_1000_kcal",
    "meat_per_1000_kcal",
    "fibre_per_1000_kcal",
    "nuts_per_1000_kcal",
    "whole_grain_per_1000_kcal",
    "fruit_vegetables_per_1000_kcal"
  )) %>% 
  mutate(response = factor(response, levels = c(
    "total_kcal",
    "carbohydrate",
    "fat",
    "protein",
    "fish_per_1000_kcal",
    "meat_per_1000_kcal",
    "fibre_per_1000_kcal",
    "nuts_per_1000_kcal",
    "whole_grain_per_1000_kcal",
    "fruit_vegetables_per_1000_kcal"
  ))) %>% 
  arrange(response) %>% 
  mutate(Variable = case_when(
    response == "total_kcal" ~ "Total energy (kcal)",
    response == "carbohydrate" ~ "Carbohydrate (g)",
    response == "fat" ~ "Fat (g)",
    response == "protein" ~ "Protein (g)",
    response == "fish_per_1000_kcal" ~ "Fish (g per 1000 kcal)",
    response == "meat_per_1000_kcal" ~ "Meat (g per 1000 kcal)",
    response == "fibre_per_1000_kcal" ~ "Fibre (g per 1000 kcal)",
    response == "nuts_per_1000_kcal" ~ "Nuts (g per 1000 kcal)",
    response == "whole_grain_per_1000_kcal" ~ "Whole-grain (g per 1000 kcal)",
    response == "fruit_vegetables_per_1000_kcal" ~ "Fruit and vegetables (g per 1000 kcal)"
  )) %>% 
  mutate(across(where(is.numeric), round, 3)) %>% 
  mutate(group = case_when(
    str_detect(response, "fish") ~ "fish_meat",
    str_detect(response, "meat") ~ "fish_meat",
    str_detect(response, "_per_1000_kcal") ~ "fibre_groups",
    response == "total_kcal" ~ "total_energy",
    TRUE ~ "macronutrients"
  )) %>% 
  ## redo FDR
  group_by(group) %>% 
  mutate(FDR = round(p.adjust(p, "BH"), 3)) %>% 
  ungroup %>% 
  mutate(" *" = case_when(
    p<0.05 & FDR<0.1 ~ "●",
    p<0.05  ~ "*",
    TRUE ~ "")) %>% 
  mutate(hmp_mean_sd = str_c(
    round(mean_HMP, 1), " (", round(sd_HMP, 1), ")",
    sep = "")) %>% 
    mutate(lmp_mean_sd = str_c(
    round(mean_LMP, 1), " (", round(sd_LMP, 1), ")",
    sep = "")) %>% 
  mutate(Test = "Wilcoxon rank-sum test") %>% 
  select(Variable,
         "HMP, mean (SD)" = hmp_mean_sd,
         "LMP, mean (SD)" = lmp_mean_sd,
         # Test,
         "p" = p,
         "q" = FDR,
         " *"
         ) %>% 
  flextable() %>% 
  ft_theme() %>% 
  # to separate the fdr adjustment
 hline(i = c(1,4,6), border = fp_border(color = "black", width = 1)) %>% 
  autofit() 

diet_gc
```


# Save results
```{r}
ft_list <- list(bsl_char, ph_gc, tt_gc, gut_gc,
                alpha_gc, alpha_cor, ub_gc,
                f_gc, p_gc, p_cor, f_cor,
                diet_gc)

is_landscape <- c(FALSE, FALSE, FALSE, FALSE,
                  FALSE, FALSE, TRUE,
                  FALSE, FALSE, FALSE, FALSE,
                  FALSE)

doc <- read_docx()
walk2(ft_list,
      seq_along(ft_list),
      function(ft, i) {
        
  # add table     
  doc <<- doc %>%
    body_add_par(paste("Supplementary Table", i),
                 style = "Normal") %>%
     body_add_par(" ", style = "Normal") %>% 
    body_add_flextable(ft)
  
  
  # decide page layout 
  if (i == 1){
    doc <<- body_end_section_portrait(doc)
  } else if (is_landscape[i]) {
       doc <<- body_end_section_landscape(doc)
   } else if (!is_landscape[i]) {
       doc <<- body_end_section_portrait(doc)
}

  
})

print(doc,
      glue("{word_dir}/{params$markdown_id}_Supplementary_Tables.docx"))
```


# Finalise
```{r}
sessionInfo()
```

