---
title: "Methane, methane groups, and methanogens"
author: "Paula Rodriguez-Garcia"
date: "`r Sys.time()`"
output:
  html_document:
    output_dir: "html"
params: 
  markdown_id: "110" # "110_MTT2" means methane TT 2group
  methane_group: "Methane_group" # "Methane_TT_2group"
  cor_method: "spearman"
---


# Setup

## Knit options
```{r setup}
knitr::opts_chunk$set(warning = FALSE, message = FALSE,
                       root.dir = rprojroot::find_rstudio_root_file())
```

## Libraries
```{r}
library(tibble)
library(dplyr)
library(magrittr)
library(tidyr)
library(glue)
library(patchwork)
library(ggplot2)
library(ggpubr)
library(gridExtra)
```

## Study paramters
```{r}
if (file.exists("../../study_parameters.R")) {
  source("../../study_parameters.R")
} else {
  source("study_parameters.R")
}

paulaR::show_rmd_params(params)
```


## Directories
```{r}
```

## Functions
```{r}
count_how_many_zero_per_column <- function (data){
    sapply(data, function(x) sum(x == 0, na.rm = T))
}
```


# Read data
## Meta
```{r}
meta_means_raw <- readRDS(meta_means_filename)
meta_all_raw   <- readRDS(meta_all_filename)
```


# Prepare data
```{r}
meta <- meta_means_raw %>% 
  filter(!is.na(!!sym(params$methane_group)))

# extract longitudinal breath, archaea, and smithii data
breath_data_all <- meta_all_raw %>% 
  filter(!is.na(!!sym(params$methane_group))) %>% 
  select(Subject_ID, fasting_CH4, Day) %>% 
  filter(!is.na(fasting_CH4)) %>% 
  pivot_wider(id_cols = "Subject_ID",
              names_from = "Day",
              values_from = fasting_CH4,
              names_prefix = "Day_") %>% 
  left_join(meta %>% select(Subject_ID, !!sym(params$methane_group)),
            by = "Subject_ID")

archaea_data_all <- meta_all_raw %>% 
  filter(!is.na(!!sym(params$methane_group))) %>% 
  select(Subject_ID, archaea_inv_ct, Day) %>% 
  arrange(Subject_ID, Day) %>% 
  filter(!is.na(archaea_inv_ct)) %>% 
  group_by(Subject_ID) %>%
  arrange(Day, .by_group = TRUE) %>%
  mutate(measurement = row_number()) %>%
  ungroup() %>%
  pivot_wider(id_cols = "Subject_ID",
              names_from = "measurement",
              values_from = archaea_inv_ct,
              names_prefix = "Measurement_") %>% 
  left_join(meta %>% select(Subject_ID, !!sym(params$methane_group)),
            by = "Subject_ID")

smithii_data_all <- meta_all_raw %>% 
  filter(!is.na(!!sym(params$methane_group))) %>% 
  select(Subject_ID, smithii_inv_ct, Day) %>% 
  arrange(Subject_ID, Day) %>% 
  filter(!is.na(smithii_inv_ct)) %>% 
  group_by(Subject_ID) %>%
  arrange(Day, .by_group = TRUE) %>%
  mutate(measurement = row_number()) %>%
  ungroup() %>%
  pivot_wider(id_cols = "Subject_ID",
              names_from = "measurement",
              values_from = smithii_inv_ct,
              names_prefix = "Measurement_") %>% 
  left_join(meta %>% select(Subject_ID, !!sym(params$methane_group)),
            by = "Subject_ID")
```

# Analyse data
## Chisq detected smithii
```{r}
cont_table_smi <- meta %>% 
  mutate(is_smithii_detected = ifelse(is.na(smithii_inv_ct), "No", "Yes")) %>% 
  select(params$methane_group, is_smithii_detected) %>% 
  table()

test_smi <- chisq.test(cont_table_smi)

p_label_smi <- ifelse(test_smi$p.value < 0.001, "p < 0.001", glue("p = {test_smi$p.value}"))


cont_df_smi  <- cont_table_smi %>% as_tibble()

# used for the plot later
cont_plot_smi <- tableGrob(
  tibble(
    !!p_label_smi := c("Detected", "Not detected"),
    "HMP" = c(cont_df_smi %>% filter(!!sym(params$methane_group) == "HMP", is_smithii_detected == "Yes" ) %>% pull(n),
              cont_df_smi %>% filter(!!sym(params$methane_group) == "HMP", is_smithii_detected == "No" ) %>% pull(n)),
    "LMP" = c(cont_df_smi %>% filter(!!sym(params$methane_group) == "LMP", is_smithii_detected == "Yes" ) %>% pull(n),
          cont_df_smi %>% filter(!!sym(params$methane_group) == "LMP", is_smithii_detected == "No" ) %>% pull(n))
  ), 
  rows = NULL,
    theme = ttheme_default(
    core = list(fg_params = list(cex = 1.1),  # text size inside table
                bg_params = list(fill = "white", col = "black")),  # white background, black border
    colhead = list(fg_params = list(cex = 1.1),  # header text size
                   bg_params = list(fill = "white", col = "black"))
  )
)


```




# Figures

## Aesthetics
```{r}
fontsize <- 14
methane_uni_color <- "#9EAFBA"
```

## 1) CH4 day 2 and CH4 day 9
```{r}
x <- "Day_2"
y <- "Day_9"

xlim <- c(0, paulaR::ceiling_signif(max(breath_data_all[[x]], breath_data_all[[y]], na.rm = T), 1))
ylim <- xlim

test <- breath_data_all %>%
  summarise(
    r = cor(!!sym(x), !!sym(y), method = params$cor_method),
    p = cor.test(!!sym(x), !!sym(y), method = params$cor_method, exact = FALSE)$p.value
  ) %>%
  mutate(p_lab = ifelse(
    p < 0.001,
    "p < 0.001",
    glue("p = {round(p, 3)}")
    ),
    r_lab = glue("R = {round(r, 3)}"),
    label = glue("{r_lab}, {p_lab}"),
    y_pos = max(breath_data_all[[y]], na.rm = TRUE) * 1.15,
    x_pos = min(breath_data_all[[x]], na.rm = TRUE) * 1.05
   )

ch4_day2_day9_correlation <-
  breath_data_all %>% 
    ggplot(aes(x = !!sym(x),
               y = !!sym(y))
           ) +
    geom_point(
              fill = methane_uni_color, 
               shape = 21,
               stroke = 1.2,
               color = "black",
               alpha = 0.6,
               size = 3) +
    geom_smooth(method = "lm", color = "black", se = F) +
    annotate("text",
             x = test$x_pos, y = test$y_pos,
             label = test$label,
             size = fontsize/3,
             hjust = 0) +
    scale_x_continuous(limits = xlim,
                       breaks = seq(0, xlim[2], 20)
                       ) +
    scale_y_continuous(limits = ylim,
                       breaks = seq(0, ylim[2], 20)
                       ) +
    labs(title = "",
         y = "Fasting breath methane, day 9 (ppm)",
         x = "Fasting breath methane, day 2 (ppm)",
    ) +
    paulaR::white_theme(font_size = fontsize) +
    theme(legend.position = "bottom")

ch4_day2_day9_correlation
```

## 2) Breath methane and groups
```{r}
x <- params$methane_group
y <- "fasting_CH4"
ylim <- c(0, paulaR::ceiling_signif(max(meta[[y]], na.rm = T), 1))

model_data <- meta %>% 
     select(!!sym(x), !!sym(y)) %>% 
     na.omit()

subject_counts <- model_data %>% 
  group_by(group = !!sym(x)) %>% 
  count()

labels <- setNames(
  paste0(subject_counts$group, "\n(N=", subject_counts$n, ")"),
  subject_counts$group
)

test <- compare_means(as.formula(paste(y, "~", x)),
                      model_data,
                      "wilcox.test",
                      exact = FALSE) %>%
  mutate(label = ifelse(
    p < 0.001,
    "p < 0.001",
    glue("p = {round(p, 3)}")
    ),
    y.position = max(model_data[[y]], na.rm = TRUE) * 1.05
    )

ch4_groups_boxplot <- model_data %>% 
     ggplot(aes(x = !!sym(x), 
                y = !!sym(y),
                fill = !!sym(x))) + 
     geom_boxplot(outlier.shape = NA,
                  color = "black",
                  show.legend = FALSE) +
     geom_jitter(aes(fill = !!sym(x)), 
                 color = "black",
                 shape = 21,
                 size = 3,
                 stroke = 1.2,
                 height = 0,
                 alpha = 0.6,
                 show.legend = TRUE) +
     scale_fill_manual(values = study_colors$methane) +
    stat_pvalue_manual(
        test,
        label = "label",
        tip.length = 0.01,
        size = fontsize/3
      ) +
  scale_x_discrete(labels = labels) +
     labs(title = "",
          y = "Fasting breath methane (ppm)",
          x = "",
          ) +
    coord_cartesian(ylim = ylim) +
    scale_y_continuous(expand = expansion(mult = c(0.05, 0.15))) +
    paulaR::white_theme(font_size = fontsize) +
    theme(legend.position = "none")

ch4_groups_boxplot
```

## 3) Breath methane and Archaea
```{r}
y <- "archaea_inv_ct"
x <- "fasting_CH4"


ylim <- c(paulaR::trunc_signif(min(meta[[y]], na.rm = T), -2)+0.005,
          paulaR::ceiling_signif(max(meta[[y]], na.rm = T), -2))


xlim <- c(0, paulaR::ceiling_signif(max(meta[[x]], na.rm = T), 1)+10)

test <- meta %>%
  summarise(
    r = cor(!!sym(x), !!sym(y), method = params$cor_method),
    p = cor.test(!!sym(x), !!sym(y), method = params$cor_method, exact = FALSE)$p.value
  ) %>%
  mutate(p_lab = ifelse(
    p < 0.001,
    "p < 0.001",
    glue("p = {round(p, 3)}")
    ),
    r_lab = glue("R = {round(r, 3)}"),
    label = glue("{r_lab}, {p_lab}"),
    y_pos = min(meta[[y]], na.rm = TRUE) ,
    x_pos = min(meta[[x]], na.rm = TRUE) + 10
   )

arc_ch4_correlation <-
  meta %>% 
  select(!!sym(x), !!sym(y), !!sym(params$methane_group)) %>% 
  na.omit() %>% 
    ggplot(aes(x = !!sym(x),
               y = !!sym(y))
           ) +
    geom_point(
      fill = methane_uni_color,
               shape = 21,
               stroke = 1.2,
               color = "black",
               alpha = 0.6,
               size = 3) +
    geom_smooth(method = "lm", color = "black", se = F) +
    annotate("text",
             x = test$x_pos, y = test$y_pos,
             label = test$label,
             size = fontsize/3,
             hjust = 0) +
    scale_x_continuous(limits = xlim,
                       breaks = seq(xlim[1], xlim[2], 20)
                       ) +
    scale_y_continuous(limits = ylim,
                       breaks = seq(ylim[1]+0.005, ylim[2], 0.01)
                       ) +
    labs(title = "",
         x = "Fasting breath methane (ppm)",
         y = "Archaea qPCR signal (1/Ct)",
         fill = ""
    ) +
    paulaR::white_theme(font_size = fontsize) +
    theme(legend.position = "bottom")

arc_ch4_correlation
```

## 4) Archaea and groups
```{r}
x <- params$methane_group
y <- "archaea_inv_ct"


ylim <- c(0, 
          paulaR::ceiling_signif(max(meta[[y]], na.rm = T), -2)+0.005)

model_data <- meta %>% 
     select(!!sym(x), !!sym(y)) %>% 
     na.omit()

subject_counts <- model_data %>% 
  group_by(group = !!sym(x)) %>% 
  count()

labels <- setNames(
  paste0(subject_counts$group, "\n(N=", subject_counts$n, ")"),
  subject_counts$group
)


test <- compare_means(as.formula(paste(y, "~", x)),
                      model_data,
                      "wilcox.test",
                      exact = FALSE) %>%
  mutate(label = ifelse(
    p < 0.001,
    "p < 0.001",
    glue("p = {round(p, 3)}")
    ),
    y.position = max(model_data[[y]], na.rm = TRUE) * 1.05
    )

arc_groups_boxplot <- 
  model_data %>% 
     select(!!sym(x), !!sym(y)) %>% 
     filter(!is.na(!!sym(x)), !is.na(!!sym(y))) %>% 
     ggplot(aes(x = !!sym(x), 
                y = !!sym(y),
                fill = !!sym(x))) + 
     geom_boxplot(outlier.shape = NA,
                  color = "black",
                  show.legend = FALSE) +
     geom_jitter(aes(fill = !!sym(x)), 
                 color = "black",
                 shape = 21,
                 size = 3,
                 height = 0,
                 stroke = 1.2,
                 alpha = 0.6,
                  show.legend = TRUE) +
     scale_fill_manual(values = study_colors$methane) +
    stat_pvalue_manual(
        test,
        label = "label",
        tip.length = 0.01,
        size = fontsize/3
      ) +
  scale_x_discrete(labels = labels) +
     labs(title = "",
          y = "Archaea qPCR signal (1/Ct)",
          x = "",
          ) +
    coord_cartesian(ylim = ylim, 
                    ) +
    scale_y_continuous(expand = expansion(mult = c(0.05, 0.1))) +
    paulaR::white_theme(font_size = fontsize) +
    theme(legend.position = "none")

arc_groups_boxplot
```

## 5) Breath methane and M. smithii
```{r}
y <- "smithii_inv_ct"
x <- "fasting_CH4"

ylim <- c(paulaR::trunc_signif(min(meta[[y]], na.rm = T), -2)-0.005, 
          paulaR::ceiling_signif(max(meta[[y]], na.rm = T), -2)+0.01)

xlim <- c(0, paulaR::ceiling_signif(max(meta[[x]], na.rm = T), 1)+10)


test <- meta %>%
  select(!!sym(x), !!sym(y)) %>% 
  na.omit() %>% 
  summarise(
    r = cor(!!sym(x), !!sym(y), method = params$cor_method),
    p = cor.test(!!sym(x), !!sym(y), method = params$cor_method, exact = FALSE)$p.value
  ) %>%
  mutate(p_lab = ifelse(
    p < 0.001,
    "p < 0.001",
    glue("p = {round(p, 3)}")
    ),
    r_lab = glue("R = {round(r, 3)}"),
    label = glue("{r_lab}, {p_lab}"),
    y_pos = min(meta[[y]], na.rm = TRUE) ,
    x_pos = min(meta[[x]], na.rm = TRUE) + 10
   )

smi_ch4_correlation <-
  meta %>%
  select(!!sym(x), !!sym(y), !!sym(params$methane_group)) %>% 
  na.omit() %>% 
    ggplot(aes(x = !!sym(x),
               y = !!sym(y))
           ) +
    geom_point(fill = methane_uni_color,
                shape = 21,
               stroke = 1.2,
               color = "black",
               alpha = 0.6,
               size = 3) +
    geom_smooth(method = "lm", color = "black", se = F) +
    annotate("text",
             x = test$x_pos, y = test$y_pos,
             label = test$label,
             size = fontsize/3,
             hjust = 0) +
    scale_x_continuous(limits = xlim,
                       breaks = seq(xlim[1], xlim[2], 20)
                       ) +
    scale_y_continuous(limits = ylim,
                  breaks = seq(ylim[1]+0.005, ylim[2], 0.01)
                   ) +
    labs(title = "",
         x = "Fasting breath methane (ppm)",
         y = "M. smithii qPCR signal (1/Ct)",
         fill = ""
    ) +
    paulaR::white_theme(font_size = fontsize) +
    theme(legend.position = "bottom")

smi_ch4_correlation
```

## 6) M. smithii and groups
```{r}
x <- params$methane_group
y <- "smithii_inv_ct"
ylim <- c(0, paulaR::ceiling_signif(max(meta[[y]], na.rm = T), -2)+0.015)

model_data <- meta %>% 
     select(!!sym(x), !!sym(y)) %>% 
     na.omit()

subject_counts <- model_data %>% 
  group_by(group = !!sym(x)) %>% 
  count()

labels <- setNames(
  paste0(subject_counts$group, "\n(N=", subject_counts$n, ")"),
  subject_counts$group
)


test <- compare_means(as.formula(paste(y, "~", x)),
                      model_data,
                      "wilcox.test",
                      exact = FALSE) %>%
  mutate(label = ifelse(
    p < 0.001,
    "p < 0.001",
    glue("p = {round(p, 3)}")
    ),
    y.position = max(model_data[[y]], na.rm = TRUE) * 1.05
    )

smi_groups_boxplot <- 
  model_data %>% 
     select(!!sym(x), !!sym(y)) %>% 
     filter(!is.na(!!sym(x)), !is.na(!!sym(y))) %>% 
     ggplot(aes(x = !!sym(x), 
                y = !!sym(y),
                fill = !!sym(x))) + 
     geom_boxplot(outlier.shape = NA,
                  color = "black") +
     geom_jitter(aes(fill = !!sym(x)), 
                 color = "black",
                 shape = 21,
                 size = 3,
                 height = 0,
                 stroke = 1.2,
                 alpha = 0.6) +
     scale_fill_manual(values = study_colors$methane) +
    stat_pvalue_manual(
        test,
        label = "label",
        tip.length = 0.01,
        size = fontsize/3
      ) +
  scale_x_discrete(labels = labels) +
     labs(title = "",
          y = "M. smithii qPCR signal (1/Ct)",
          x = "",
          ) +
    coord_cartesian(ylim = ylim) +
    scale_y_continuous(expand = expansion(mult = c(0.05, 0.1))) +
    paulaR::white_theme(font_size = fontsize) +
    theme(legend.position = "none")  +
    annotation_custom(
      grob = cont_plot_smi,
      xmin = 1, xmax = 2, 
      ymin = 0.001, ymax = 0.025
    )

smi_groups_boxplot
```


## 7) Breath methane and hydrogen
```{r}
x <- "fasting_H2"
y <- "fasting_CH4"

xlim <- c(paulaR::trunc_signif(min(meta[[x]], na.rm = F), 0),
          paulaR::ceiling_signif(max(meta[[x]], na.rm = T), 1))
ylim <- c(0, paulaR::ceiling_signif(max(meta[[y]], na.rm = T), 1)+10)

test <- meta %>%
  summarise(
    r = cor(!!sym(x), !!sym(y), method = params$cor_method),
    p = cor.test(!!sym(x), !!sym(y), method = params$cor_method, exact = FALSE)$p.value
  ) %>%
  mutate(p_lab = ifelse(
    p < 0.001,
    "p < 0.001",
    glue("p = {round(p, 3)}")
    ),
    r_lab = glue("R = {round(r, 3)}"),
    label = glue("{r_lab}, {p_lab}"),
    y_pos = max(meta[[y]], na.rm = TRUE) * 1.15,
    x_pos = min(meta[[x]], na.rm = TRUE) * 1.05
   )

h2_ch4_correlation <-
  meta %>% 
    ggplot(aes(x = !!sym(x),
               y = !!sym(y))
           ) +
    geom_point(aes(fill = !!sym(params$methane_group)), 
               shape = 21,
               stroke = 1.2,
               color = "black",
               alpha = 0.6,
               size = 3) +
    scale_fill_manual(values = study_colors$methane) +
    geom_smooth(method = "lm", color = "black", se = F) +
    annotate("text",
             x = test$x_pos, y = test$y_pos,
             label = test$label,
             size = fontsize/3,
             hjust = 0) +
    scale_x_continuous(limits = xlim,
                       breaks = seq(xlim[1], xlim[2], 20)
                       ) +
    scale_y_continuous(limits = ylim,
                       breaks = seq(0, ylim[2], 20)
                       ) +
    labs(title = "",
         y = "Fasting breath methane (ppm)",
         x = "Fasting breath hydrogen (ppm)",
         fill = ""
    ) +
    paulaR::white_theme(font_size = fontsize) +
    theme(legend.position = "bottom") 

h2_ch4_correlation
```

## 8) M. smithii and Archaea
```{r}
x <- "archaea_inv_ct"
y <- "smithii_inv_ct"


xlim <- c(paulaR::trunc_signif(min(meta[[x]], na.rm = T), -2)+0.005,
          paulaR::ceiling_signif(max(meta[[x]], na.rm = T), -2))
ylim <- c(paulaR::trunc_signif(min(meta[[y]], na.rm = T), -2)-0.005,
          paulaR::ceiling_signif(max(meta[[y]], na.rm = T), -2)+0.01)

test <- meta %>%
    select(!!sym(x), !!sym(y)) %>% 
  na.omit() %>% 
  summarise(
    r = cor(!!sym(x), !!sym(y), method = params$cor_method),
    p = cor.test(!!sym(x), !!sym(y), method = params$cor_method, exact = FALSE)$p.value
  ) %>%
  mutate(p_lab = ifelse(
    p < 0.001,
    "p < 0.001",
    glue("p = {round(p, 3)}")
    ),
    r_lab = glue("R = {round(r, 3)}"),
    label = glue("{r_lab}, {p_lab}"),
    y_pos = max(meta[[y]], na.rm = TRUE),
    x_pos = min(meta[[x]], na.rm = TRUE) * 1.05
   )

arc_smi_correlation <-
  meta %>% 
  select(!!sym(x), !!sym(y), !!sym(params$methane_group)) %>% 
  na.omit() %>% 
    ggplot(aes(x = !!sym(x),
               y = !!sym(y))
           ) +
    geom_point(aes(fill = !!sym(params$methane_group)), 
               shape = 21,
               stroke = 1.2,
               color = "black",
               alpha = 0.6,
               size = 3) +
    scale_fill_manual(values = study_colors$methane) +
    geom_smooth(method = "lm", color = "black", se = F) +
    annotate("text",
             x = test$x_pos, y = test$y_pos,
             label = test$label,
             size = fontsize/3,
             hjust = 0) +
    scale_x_continuous(limits = xlim,
                       breaks = seq(xlim[1]+0.005, xlim[2], 0.01)
                       ) +
    scale_y_continuous(limits = ylim,
                       breaks = seq(ylim[1]+0.005, ylim[2], 0.01)
                       ) +
    labs(title = "",
         y = "Detected M. smithii qPCR signal (1/Ct)",
         x = "Archaea qPCR signal (1/Ct)",
         fill = ""
    ) +
    paulaR::white_theme(font_size = fontsize) +
    theme(legend.position = "bottom")

arc_smi_correlation
```

## 9) Archaea measurements 1 and 2
```{r}
x <- "Measurement_1"
y <- "Measurement_2"


xlim <- c(paulaR::trunc_signif(min(archaea_data_all[[x]], 
                                     archaea_data_all[[y]],
                                     na.rm = T), -2)+0.005,
          paulaR::ceiling_signif(max(archaea_data_all[[x]], 
                                     archaea_data_all[[y]],
                                     na.rm = T), -2)
          )

ylim <- xlim

test <- archaea_data_all %>%
  summarise(
    r = cor(!!sym(x), !!sym(y), method = params$cor_method),
    p = cor.test(!!sym(x), !!sym(y), method = params$cor_method, exact = FALSE)$p.value
  ) %>%
  mutate(p_lab = ifelse(
    p < 0.001,
    "p < 0.001",
    glue("p = {round(p, 3)}")
    ),
    r_lab = glue("R = {round(r, 3)}"),
    label = glue("{r_lab}, {p_lab}"),
    y_pos = max(ylim[2], na.rm = TRUE) * 0.95,
    x_pos = min(xlim[1], na.rm = TRUE) * 1.05
   )

arc_m1_m2_correlation <-
  archaea_data_all %>% 
    ggplot(aes(x = !!sym(x),
               y = !!sym(y))
           ) +
    geom_point(
      aes(fill = !!sym(params$methane_group)), 
               shape = 21,
               stroke = 1.2,
               color = "black",
               alpha = 0.6,
               size = 3) +
    scale_fill_manual(values = study_colors$methane) +
    geom_smooth(method = "lm", color = "black", se = F) +
    annotate("text",
             x = test$x_pos, y = test$y_pos,
             label = test$label,
             size = fontsize/3,
             hjust = 0) +
    scale_x_continuous(limits = xlim,
                       breaks = seq(xlim[1]+0.005, xlim[2], 0.01)
                       ) +
    scale_y_continuous(limits = ylim,
                       breaks = seq(ylim[1]+0.005, ylim[2], 0.01)
                       ) +
    labs(title = "",
         y = "Archaea, measurement 2 qPCR signal (1/Ct)",
         x = "Archaea, measurement 1 qPCR signal (1/Ct)",
         fill = ""
    ) +
    paulaR::white_theme(font_size = fontsize) +
    theme(legend.position = "bottom")

arc_m1_m2_correlation
```

## 10) M. smithii measurements 1 and 2
```{r}
x <- "Measurement_1"
y <- "Measurement_2"

xlim <- c(paulaR::trunc_signif(min(smithii_data_all[[x]], 
                                     smithii_data_all[[y]],
                                     na.rm = T), -2)+0.005,
          paulaR::ceiling_signif(max(smithii_data_all[[x]], 
                                     smithii_data_all[[y]],
                                     na.rm = T), -2)
          )
ylim <- xlim

test <- smithii_data_all %>%
  select(!!sym(x), !!sym(y), !!sym(params$methane_group)) %>% 
  na.omit() %>% 
  summarise(
    r = cor(!!sym(x), !!sym(y), method = params$cor_method),
    p = cor.test(!!sym(x), !!sym(y), method = params$cor_method, exact = FALSE)$p.value
  ) %>%
  mutate(p_lab = ifelse(
    p < 0.001,
    "p < 0.001",
    glue("p = {round(p, 3)}")
    ),
    r_lab = glue("R = {round(r, 3)}"),
    label = glue("{r_lab}, {p_lab}"),
    y_pos = max(ylim[2], na.rm = TRUE) * 0.98,
    x_pos = min(xlim[1], na.rm = TRUE) * 1.05
   )

smi_m1_m2_correlation <-
  smithii_data_all %>% 
    select(!!sym(x), !!sym(y), !!sym(params$methane_group)) %>% 
  na.omit() %>% 
    ggplot(aes(x = !!sym(x),
               y = !!sym(y))
           ) +
    geom_point(
      aes(fill = !!sym(params$methane_group)), 
               shape = 21,
               stroke = 1.2,
               color = "black",
               alpha = 0.6,
               size = 3) +
    scale_fill_manual(values = study_colors$methane) +
    geom_smooth(method = "lm", color = "black", se = F) +
    annotate("text",
             x = test$x_pos, y = test$y_pos,
             label = test$label,
             size = fontsize/3,
             hjust = 0) +
    scale_x_continuous(limits = xlim,
                       breaks = seq(xlim[1]+0.005, xlim[2], 0.01)
                       ) +
    scale_y_continuous(limits = ylim,
                       breaks = seq(ylim[1]+0.005, ylim[2], 0.01)
                       ) +
    labs(title = "",
         y = "Detected M. smithii, measurement 2 qPCR signal (1/Ct)",
         x = "Detected M. smithii, measurement 1 qPCR signal (1/Ct)",
         fill = ""
    ) +
    paulaR::white_theme(font_size = fontsize) +
    theme(legend.position = "bottom")

smi_m1_m2_correlation
```

## 11) Breath methane and WGTTT
```{r}
# this plot is included in supplementary when params$methane_group = "Methane_TT_2group"
x <- "WGTT"
y <- "fasting_CH4"

xlim <- c(paulaR::trunc_signif(min(meta[[x]], na.rm = T), 0),
          paulaR::ceiling_signif(max(meta[[x]], na.rm = T), 0))
ylim <- c(0, paulaR::ceiling_signif(max(meta[[y]], na.rm = T), 1)+10)

test <- meta %>%
  select(!!sym(x), !!sym(y)) %>% 
  na.omit() %>% 
  summarise(
    r = cor(!!sym(x), !!sym(y), method = params$cor_method),
    p = cor.test(!!sym(x), !!sym(y), method = params$cor_method, exact = FALSE)$p.value
  ) %>%
  mutate(p_lab = ifelse(
    p < 0.001,
    "p < 0.001",
    glue("p = {round(p, 3)}")
    ),
    r_lab = glue("R = {round(r, 3)}"),
    label = glue("{r_lab}, {p_lab}"),
    y_pos = max(meta[[y]], na.rm = TRUE) * 1.15,
    x_pos = min(meta[[x]], na.rm = TRUE) * 1.05
   )

wgtt_ch4_correlation <-
  meta %>% 
    select(!!sym(x), !!sym(y), !!sym(params$methane_group)) %>% 
  na.omit() %>% 
    ggplot(aes(x = !!sym(x),
               y = !!sym(y))
           ) +
    geom_point(aes(fill = !!sym(params$methane_group)), 
               shape = 21,
               stroke = 1.2,
               color = "black",
               alpha = 0.6,
               size = 3) +
    scale_fill_manual(values = study_colors$methane) +
    geom_smooth(method = "lm", color = "black", se = F) +
    annotate("text",
             x = test$x_pos, y = test$y_pos,
             label = test$label,
             size = fontsize/3,
             hjust = 0) +
    scale_x_continuous(limits = xlim,
                       breaks = seq(xlim[1], xlim[2], 10)
                       ) +
    scale_y_continuous(limits = ylim,
                       breaks = seq(0, ylim[2], 20)
                       ) +
    labs(title = "",
         y = "Fasting breath methane (ppm)",
         x = "Whole-gut transit time (hours)",
         fill = ""
    ) +
    paulaR::white_theme(font_size = fontsize) +
    theme(legend.position = "bottom")

wgtt_ch4_correlation
```

## 12) Histrogram breath methane
```{r}
x <- "fasting_CH4"

xlim <- c(0, paulaR::ceiling_signif(max(meta[[x]], na.rm = T), 1)+10)
brks <- c(0,1,2,3,4,5,6,7,8,9,10,20,30,40,50,60,70,80)


ch4_hist <- meta %>%
  select(!!sym(x), !!sym(params$methane_group)) %>% 
  na.omit() %>% 
    ggplot(aes(x = !!sym(x))
           ) +
  geom_histogram(
    fill = methane_uni_color,
    breaks = brks,
    closed = "left"
    ) +

  geom_vline(xintercept = 10, linetype = "dashed") +
  scale_x_continuous(limits = xlim,
                     breaks = seq(xlim[1], xlim[2], 20)) +

  scale_y_continuous(limits = c(0, 30),
                     breaks = seq(0, 30, 10)) + 
    labs(title = "",
         x = "Fasting breath methane (ppm)",
         y = "Number of subjects"
    ) +
    paulaR::white_theme(font_size = fontsize) +
    theme(legend.position = "bottom")


ch4_hist
```

## 13) Histogram breath methane, day 2 and day 9
```{r}
x <- "Day_2"

ch4_hist_d2 <- breath_data_all %>%
  select(!!sym(x), !!sym(params$methane_group)) %>% 
  na.omit() %>% 
    ggplot(aes(x = !!sym(x))
           ) +
  geom_histogram(
      fill = methane_uni_color,
      breaks = c(0, 1, 5, 10, 20, 30, 40, 50, 60, 70),
      closed = "left"
    ) +
   geom_vline(xintercept = 10, linetype = "dashed") +
  scale_x_continuous(limits = xlim,
                     breaks = seq(xlim[1], xlim[2], 20)) +
  scale_y_continuous(limits = c(0, 30),
                     breaks = seq(0, 30, 10)) + 
    labs(title = "",
         x = "Fasting breath methane on day 2 (ppm)",
         y = "Number of subjects"
    ) +
    paulaR::white_theme(font_size = fontsize) +
    theme(legend.position = "bottom")


x <- "Day_9"

ch4_hist_d9 <- breath_data_all %>%
  select(!!sym(x), !!sym(params$methane_group)) %>% 
  na.omit() %>% 
    ggplot(aes(x = !!sym(x))
           ) +
  geom_histogram(
      fill = methane_uni_color,
      breaks = c(0, 1, 5, 10, 20, 30, 40, 50, 60, 70),
      closed = "left"
    ) +
   geom_vline(xintercept = 10, linetype = "dashed") +
  scale_x_continuous(limits = xlim,
                     breaks = seq(xlim[1], xlim[2], 20)) +
  scale_y_continuous(limits = c(0, 30),
                     breaks = seq(0, 30, 10)) + 
    labs(title = "",
         x = "Fasting breath methane on day 9 (ppm)",
         y = "Number of subjects"
    ) +
    paulaR::white_theme(font_size = fontsize) +
    theme(legend.position = "bottom")
```

## 14) Histogram Archaea
```{r}
x <- "archaea_inv_ct"


xlim <- c(paulaR::trunc_signif(min(meta[[x]], na.rm = T), -2)-0.001,
          paulaR::ceiling_signif(max(meta[[x]], na.rm = T), -2)+0.001)

arc_hist <- meta %>%
  select(!!sym(x), !!sym(params$methane_group)) %>% 
  na.omit() %>% 
    ggplot(aes(x = !!sym(x))
           ) +
  geom_histogram(
    fill = methane_uni_color,
   breaks = seq(xlim[1]+0.001, xlim[2]-0.001, 0.005),
   closed = "left"
    ) +
  scale_x_continuous(limits = xlim,
                     breaks = seq(xlim[1]+0.001, xlim[2]-0.001, 0.01)
                     ) +
  scale_y_continuous(limits = c(0, 40), 
                     breaks = seq(0, 40, 10)) + 
    labs(title = "",
         x = "Archaea qPCR signal (1/Ct)",
         y = "Number of subjects"
    ) +
    paulaR::white_theme(font_size = fontsize) +
    theme(legend.position = "bottom")


arc_hist
```


## 15) Histogram smithii
```{r}
x <- "smithii_inv_ct"

xlim <- c(-0.001,
          paulaR::ceiling_signif(max(meta[[x]], na.rm = T), -2)+0.01)


smi_hist <- meta %>%
  select(!!sym(x), !!sym(params$methane_group)) %>% 
  mutate(!!sym(x) := ifelse(is.na(!!sym(x)), 0, !!sym(x))) %>% 
  ggplot(aes(x = !!sym(x))
           ) +
  geom_histogram(
    fill = methane_uni_color,
    breaks = seq(xlim[1]+0.001, xlim[2], 0.005),
    closed = "left"
    ) +
  scale_x_continuous(limits = xlim,
                     breaks = seq(xlim[1]+0.001, xlim[2], 0.02),
                     labels = c("0\n(not detected)", 0.02, 0.04, 0.06)
                     ) +
  scale_y_continuous(limits = c(0, 32),
                     breaks = seq(0, 32, 10)) +
    labs(title = "",
         x = "M. smithii qPCR signal (1/Ct)",
         y = "Number of subjects"
    ) +
    paulaR::white_theme(font_size = fontsize) +
    theme(legend.position = "bottom")


smi_hist
```

## 16) Archaea qpcr log10 and inv ct
```{r}
x <- "archaea_inv_ct"
y <- "archaea_qpcr_log10"

xlim <- c(0.025, 0.06)
ylim <- c(5,11)

meta %>% 
    select(!!sym(x), !!sym(y), !!sym(params$methane_group)) %>% 
  na.omit() %>% 
    ggplot(aes(x = !!sym(x),
               y = !!sym(y))
           ) +
    geom_point(aes(fill = !!sym(params$methane_group)), 
               shape = 21,
               stroke = 1.2,
               color = "black",
               alpha = 0.6,
               size = 3) +
    scale_fill_manual(values = study_colors$methane) +
    geom_smooth(method = "lm", color = "black", se = F) +
    scale_x_continuous(limits = xlim,
                        breaks = seq(xlim[1]+0.005, xlim[2], 0.01)
                       ) +
    scale_y_continuous(limits = ylim,
                        breaks = seq(ylim[1], ylim[2], 1)
                       ) +
    labs(title = "",
         y = "log10(copies/g)",
         x = "1/Ct",
         fill = ""
    ) +
    paulaR::white_theme(font_size = fontsize) +
    theme(legend.position = "bottom")
```

## 17) Hydrogen and groups
```{r}
x <- params$methane_group
y <- "fasting_H2"

ylim <- c(paulaR::trunc_signif(min(meta[[y]], na.rm = F), 0),
          paulaR::ceiling_signif(max(meta[[y]], na.rm = T), 1))

model_data <- meta %>% 
     select(!!sym(x), !!sym(y)) %>% 
     na.omit()

subject_counts <- model_data %>% 
  group_by(group = !!sym(x)) %>% 
  count()

labels <- setNames(
  paste0(subject_counts$group, "\n(N=", subject_counts$n, ")"),
  subject_counts$group
)


test <- compare_means(as.formula(paste(y, "~", x)),
                      model_data,
                      "wilcox.test",
                      exact = FALSE) %>%
  mutate(label = ifelse(
    p < 0.001,
    "p < 0.001",
    glue("p = {round(p, 3)}")
    ),
    y.position = max(model_data[[y]], na.rm = TRUE) * 1.05
    )

h2_groups_boxplot <- 
  model_data %>% 
     select(!!sym(x), !!sym(y)) %>% 
     filter(!is.na(!!sym(x)), !is.na(!!sym(y))) %>% 
     ggplot(aes(x = !!sym(x), 
                y = !!sym(y),
                fill = !!sym(x))) + 
     geom_boxplot(outlier.shape = NA,
                  color = "black",
                  show.legend = FALSE) +
     geom_jitter(aes(fill = !!sym(x)), 
                 color = "black",
                 shape = 21,
                 size = 3,
                 height = 0,
                 stroke = 1.2,
                 alpha = 0.6,
                  show.legend = TRUE) +
     scale_fill_manual(values = study_colors$methane) +
    stat_pvalue_manual(
        test,
        label = "label",
        tip.length = 0.01,
        size = fontsize/3
      ) +
  scale_x_discrete(labels = labels) +
     labs(title = "",
          y = "Fasting breath hydrogen (ppm)",
          x = "",
          ) +
    coord_cartesian(ylim = ylim, 
                    ) +
    scale_y_continuous(limits = ylim,
                       breaks = seq(ylim[1], ylim[2], 20)) +
    paulaR::white_theme(font_size = fontsize) +
    theme(legend.position = "none")

h2_groups_boxplot
```


# Create figure 1
```{r}
fig_1 <- 
( # A
  (ch4_hist +
      theme(legend.position = "none", 
            plot.margin = margin(6, 0, 20, 6))) +

    # B
  (arc_ch4_correlation + 
      theme(plot.margin = margin(6, 0, 20, 6)
                                )) +

     # C
   (smi_ch4_correlation  +
      theme(legend.position = "none",
            plot.margin = margin(6, 0, 20, 6))) +
    
         # D
   (ch4_groups_boxplot  +
      geom_hline(yintercept = 10, linetype = "dashed") +
      theme(legend.position = "none",
            plot.margin = margin(6, 0, 6, 6)))  +

    # E
  (arc_groups_boxplot  +
           labs(fill = "") + 
      theme(legend.position = "bottom",
            legend.justification = "center",
            plot.margin = margin(6, 0, 6, 6))) +
     # F
   (smi_groups_boxplot  + 
      theme(legend.position = "none", 
            plot.margin = margin(6, 0, 6, 6)))
   
) +
  plot_layout(ncol = 3, byrow = TRUE) +
  plot_annotation(tag_levels = list(c("B", "C", "D", "E", "F", "G")))


fig_1
```


# Save figures
## Independent figures
```{r}
format <- "png" # png # pdf

ch4_day2_day9_correlation
ggsave(glue("{fig_dir}/{params$markdown_id}_CH4_Day2_Day9_correlation.{format}"),
       bg = "white",
       width = 4,
       height = 3.5,
       dpi = 300)

ch4_groups_boxplot
ggsave(glue("{fig_dir}/{params$markdown_id}_CH4_groups_boxplot.{format}"),
       bg = "white",
       width = 4,
       height = 4,
       dpi = 300)

arc_ch4_correlation
ggsave(glue("{fig_dir}/{params$markdown_id}_Archaea_CH4_correlation.{format}"),
       bg = "white",
       width = 4,
       height = 4,
       dpi = 300)


arc_groups_boxplot
ggsave(glue("{fig_dir}/{params$markdown_id}_Archaea_groups_boxplot.{format}"),
       bg = "white",
       width = 4,
       height = 4,
       dpi = 300)


smi_ch4_correlation
ggsave(glue("{fig_dir}/{params$markdown_id}_M_smithii_CH4_correlation.{format}"),
       bg = "white",
       width = 4,
       height = 4,
       dpi = 300)


smi_groups_boxplot +
  labs(caption = "Samples only shown if M. smithii detected.") +
  theme(plot.caption = element_text(size = 10))
ggsave(glue("{fig_dir}/{params$markdown_id}_M_smithii_groups_boxplot.{format}"),
       bg = "white",
       width = 4,
       height = 4,
       dpi = 300)


h2_ch4_correlation
ggsave(glue("{fig_dir}/{params$markdown_id}_CH4_H2_correlation.{format}"),
       bg = "white",
       width = 4,
       height = 4,
       dpi = 300)


arc_smi_correlation
ggsave(glue("{fig_dir}/{params$markdown_id}_Archaea_M_smithii_correlation.{format}"),
       bg = "white",
       width = 4,
       height = 4,
       dpi = 300)


arc_m1_m2_correlation
ggsave(glue("{fig_dir}/{params$markdown_id}_Archaea_M1_M2_correlation.{format}"),
       bg = "white",
       width = 4,
       height = 4,
       dpi = 300)


smi_m1_m2_correlation
ggsave(glue("{fig_dir}/{params$markdown_id}_M_smithii_M1_M2_correlation.{format}"),
       bg = "white",
       width = 4,
       height = 4,
       dpi = 300)

wgtt_ch4_correlation
ggsave(glue("{fig_dir}/{params$markdown_id}_CH4_WGTT_correlation.{format}"),
       bg = "white",
       width = 4,
       height = 4,
       dpi = 300)


ch4_hist
ggsave(glue("{fig_dir}/{params$markdown_id}_CH4_histogram.{format}"),
       bg = "white",
       width = 4,
       height = 4,
       dpi = 300)

ch4_hist
ggsave(glue("{fig_dir}/{params$markdown_id}_CH4_histogram_threshold.{format}"),
       bg = "white",
       width = 4,
       height = 4,
       dpi = 300)


ch4_hist_d2
ggsave(glue("{fig_dir}/{params$markdown_id}_CH4_histogram_day2.{format}"),
       bg = "white",
       width = 4,
       height = 4,
       dpi = 300)
ch4_hist_d9 
ggsave(glue("{fig_dir}/{params$markdown_id}_CH4_histogram_day9.{format}"),
       bg = "white",
       width = 4,
       height = 4,
       dpi = 300)
```


## Fig 1 joined
```{r}
format <- "png" # png # pdf


fig_1
ggsave(glue("{fig_dir}/{params$markdown_id}_Figure_1_panel_B_F.{format}"),
       bg = "white",
       width = 10,
       height = 8,
       dpi = 300)
```

## Supl figures
```{r}
arc_hist + smi_hist + plot_annotation(tag_levels = "A")

ggsave(glue("{fig_dir}/{params$markdown_id}_Archaea_Msmithii_histograms.{format}"),
       bg = "white",
       width = 8,
       height = 4,
       dpi = 300)


h2_ch4_correlation + h2_groups_boxplot + plot_annotation(tag_levels = "A")

ggsave(glue("{fig_dir}/{params$markdown_id}_H2_CH4_corr_boxplot.{format}"),
       bg = "white",
       width = 8,
       height = 4,
       dpi = 300)


saveRDS(wgtt_ch4_correlation,
        glue("{rds_dir}/{params$markdown_id}_WGTT_CH4_correlation.rds"))
```



# Finalise
```{r}
sessionInfo()
```

