---
title: "Clean microbiome data"
author: "Paula Rodriguez-Garcia"
date: "`r Sys.time()`"
output:
  html_document:
    output_dir: "html"
params: 
  markdown_id: "021"
---

Microbiome data will be cleaned/processed from this file:
- merged.phyloseq_object_rdp_v18.RData
This file was generated at DTU following their own pipeline (https://doi.org/10.11583/DTU.22657339), which includes: demultiplexing, DADA2, and taxonomic annotation


The microbiome post-processing conducted in this script consists of these steps:
1) Filter samples: remove non-biological samples, samples with < 10000 total reads
2) Filter ASVs: remove mitochondria/chloroplasts and singletons/doubletons
3a) Save the filtered read counts
3b) Turn counts to rel. abundances
3c) Rarefy read counts
3d) Turn the rarefied counts to relative abundances
4) For the 4 datasets from step 3: aggregate ASVs to genus and family levels
5) Calculate alpha diversity from the rarefied relative abundances, at ASV level
6) For the 8 datasets from steps 3-4: calculate subject-means (*)


The resulting output files will be:

- 1) asv_counts.rds
- 2) asv_relabu.rds
- 3) asv_rar_counts.rds
- 4) asv_rar_relabu.rds

- 5) genus_counts.rds
- 6) genus_relabu.rds
- 7) genus_rar_counts.rds
- 8) genus_rar_relabu.rds

- 9)  asv_counts_means.rds
- 10) asv_relabu_means.rds
- 11) asv_rar_counts_means.rds
- 12) asv_rar_relabu_means.rds

- 13) genus_counts_means.rds
- 14) genus_relabu_means.rds
- 15) genus_rar_counts_means.rds
- 16) genus_rar_relabu_means.rds

- 17) family_counts_means.rds
- 18) family_relabu_means.rds
- 19) family_rar_counts_means.rds
- 20) family_rar_relabu_means.rds

- 21) alpha_diversity.rds
- 22) alpha_diversity_means.rds

- 23) taxonomic_annotation.rds





# Setup
## Knit options
```{r setup}
knitr::opts_chunk$set(warning = FALSE, message = FALSE,
                      root.dir = rprojroot::find_rstudio_root_file())
```

## Libraries
```{r}
library(tibble)
library(dplyr)
library(magrittr)
library(glue)
library(stringr)
library(phyloseq)
library(vegan)
```

## Study paramters
```{r}
if (file.exists("../../study_parameters.R")) {
  source("../../study_parameters.R")
} else {
  source("study_parameters.R")
}

paulaR::show_rmd_params(params)
```

## Directories
```{r}
```

## Functions
```{r}
lowest_assigned_rank <- function(row, ranks) {
  vals <- row[ranks]
  vals <- vals[!is.na(vals) & vals != ""]
  if (length(vals) == 0) return(NA_character_)
  tail(vals, 1)
}
```

## Seed
```{r}
set.seed(631)
```


# Read data

## Raw phyloseq object from DTU
```{r}
path_to_raw_data <- # here goes path to raw data
load(glue("{path_to_raw_data}/merged.phyloseq_object_rdp_v18.RData"))

data_raw <- phy@otu_table %>% as.data.frame() %>% as_tibble(rownames = "Sequencing_ID")
tax_raw  <- phy@tax_table %>% as.data.frame() %>% as_tibble(rownames = "ASV")
```


## Mapping file (maps sequencing_IDs and)
```{r}
mapping <- readRDS(glue("{rds_dir}/003_mapping_sequencing_info.rds"))
```

# Clean data
## Substitute the sequencing IDs by the sample IDs
```{r}
data <- data_raw %>% 
  left_join(mapping %>% select(Sequencing_ID, Sample_ID, Subject_ID), by = "Sequencing_ID") %>% 
  select(Sample_ID, Subject_ID, Sequencing_ID, everything())

data_raw %>% dim() # 547 x 1549 (1 ID column)
data     %>% dim() # 547 x 1551 (3 ID columns)
```

## Create a taxonomic annotation file
```{r}
ranks <- tax_raw %>% select(-ASV) %>% as.data.frame() %>% colnames()

taxonomic_annotation <- tax_raw %>%  
  rowwise() %>%
  mutate(
    closest_tax = case_when(
      !is.na(Genus) & !is.na(Species) ~
        paste(Genus, Species),
      
      !is.na(Genus) & is.na(Species) ~
        paste(Genus, "sp."),
      
      is.na(Genus) ~
        paste("Unclassified", lowest_assigned_rank(pick(all_of(ranks)), ranks))
      
    ), .after = "ASV"
  ) %>%
  ungroup() %>% 
  mutate(genus_or_closest = case_when(
    is.na(Genus) ~ closest_tax,
    TRUE ~ Genus
  ),
  .after = "Genus") %>% 
  mutate(family_or_closest = case_when(
    is.na(Family) ~ closest_tax,
    TRUE ~ Family
  ),
  .after = "Family"
  )
```




# Process data
## 1) Filter samples: remove non-biological samples, samples with < 10000 total reads
```{r}
# Identify non-biological samples
data %>% nrow() # 547 samples

data %>% 
  filter(str_detect(Sequencing_ID, "Mock") |
           str_detect(Sequencing_ID, "Water") | 
           str_detect(Sequencing_ID, "Tag") ) %>% 
  nrow() ## 77 non-biological samples

non_bio_samples <- data %>% 
  filter(is.na(Subject_ID)) %>% 
  pull(Sequencing_ID)

non_bio_samples %>% length() # 77


# Identify samples < 10000 total reads
samples_low_reads <- mapping %>%
  select(Sequencing_ID, Subject_ID, Reads) %>% 
  filter(Reads < 10000, !is.na(Subject_ID)) %>% 
  pull(Sequencing_ID)

samples_low_reads %>% length() # 99

# Filter samples
data_step_1 <- data %>% 
    filter(! Sequencing_ID %in% non_bio_samples,
           ! Sequencing_ID %in% samples_low_reads)


data_step_1 %>% dim() # 371 x 1551 (3 ID columns)
```

## 2) Filter ASVs: remove mitochondria/chloroplasts and singletons/doubletons
```{r}
# Identify mitochondria/chloroplast ASVs
mit_chlor_asv <- tax_raw %>% 
  filter(str_detect(Class, "Chloroplast") | str_detect(Family, "mitochondria") | str_detect(Phylum, "Cyanobacteria/Chloroplast")) %>%
  pull(ASV)

mit_chlor_asv %>% length() # 6

# Identify singletons/doubletons
single_doubletons_asv <- data_step_1 %>% 
  select(contains("ASV")) %>% 
  mutate(across(everything(), ~ifelse(. > 0, TRUE, FALSE))) %>% 
  colSums() %>% 
  enframe(name = "ASV", value = "samples_detected") %>% 
  filter(samples_detected <= 2) %>% 
  pull(ASV)

single_doubletons_asv %>% length() # 1341


# In case we consider only removing singletons
singletons_asv <- data_step_1 %>%
  select(contains("ASV")) %>%
  mutate(across(everything(), ~ifelse(. > 0, TRUE, FALSE))) %>%
  colSums() %>%
  enframe(name = "ASV", value = "samples_detected") %>%
  filter(samples_detected <= 1) %>%
  pull(ASV)

singletons_asv %>% length() # 280


data_step_2 <- data_step_1 %>%
    select(everything(),
           -any_of(mit_chlor_asv),
           # -any_of(single_doubletons_asv),
           -any_of(singletons_asv),
           )

data_step_2 %>% dim() 
# 371 x 210 (3 ID columns) if doubletons filtered out,
# or 371 x 1266 if singletons filtered out
```

## 3) Rarefaction and rel abu
```{r}
# Non-rarefied counts
asv_counts <- data_step_2 %>% 
  select(Sample_ID, everything(), -Subject_ID, -Sequencing_ID)

asv_counts %>% dim() # 371 x 1264


# Non-rarefied relative abundances
asv_relabu <- asv_counts %>% 
    mutate(
    across(
      -Sample_ID,
      ~ .x / rowSums(across(-Sample_ID)) * 100
    )
  )

asv_relabu %>% dim() # 371 x 1264
asv_relabu %>% select(-Sample_ID) %>% rowSums() %>% unique() # should be 100


# Rarefied counts
asv_rar_counts <- vegan::rrarefy(
  asv_counts %>% 
    select(-Sample_ID) %>%
    as.data.frame(),
  # minimum total number of reads across samples (after ASV filtering)
  sample = asv_counts %>%
    select(-Sample_ID) %>% 
    rowSums() %>% 
    min()
  ) %>%
  as_tibble() %>%
  bind_cols(Sample_ID = asv_counts$Sample_ID) %>% 
  relocate(Sample_ID, .before = 1)

asv_rar_counts %>% dim() # 371 x 1264
asv_rar_counts %>% select(-Sample_ID) %>% rowSums() %>% unique() # if one single number above 10000 is shown, it worked



# Rarefied relative abundances
asv_rar_relabu <- asv_rar_counts %>% 
    mutate(
    across(
      -Sample_ID,
      ~ .x / rowSums(across(-Sample_ID)) * 100
    )
  )

asv_rar_relabu %>% dim() # 371 x 1264
asv_rar_relabu %>% select(-Sample_ID) %>% rowSums() %>% unique() # should be 100
```

## 4) Aggregate at higher taxonomic levels
### 4.1) gGenus level
```{r}
### for those ASVs unclassified as genus level, label them at "Unclassified + lowest-taxonomic-rank-assigned"


## 1) Genus counts
genus_counts <- 
  asv_counts %>% 
  # Transpose data
  column_to_rownames("Sample_ID") %>% 
  as.matrix() %>% 
  t() %>% 
  as.data.frame() %>%
  as_tibble(rownames = "ASV") %>% 
  left_join(taxonomic_annotation %>%
              select(ASV, genus_or_closest),
            by = "ASV") %>%
  select(ASV, genus_or_closest, everything()) %>% 
  # Aggregate counts
  select(-ASV) %>% 
  group_by(genus_or_closest) %>% 
  summarise(across(everything(), sum, na.rm = TRUE)) %>% 
  # Transpose data back
  column_to_rownames("genus_or_closest") %>%
  as.data.frame() %>%
  t() %>% 
  as_tibble(rownames = "Sample_ID") %>% 
  select(Sample_ID, !contains("Unclassified"), contains("Unclassified"))

genus_counts %>% dim() # 371 x 205
genus_counts %>% colnames() 
genus_counts %>% select(contains("Unclassified")) %>% colnames() %>% length() # 45 unclassified taxa
204-45
# 204 taxa: 159 genera, 45 unclassified taxa

## 2) Genus rel abu
genus_relabu <- genus_counts %>% 
    mutate(
    across(
      -Sample_ID,
      ~ .x / rowSums(across(-Sample_ID)) * 100
    )
  )

genus_relabu %>% dim() # 371 x 205
genus_relabu %>% select(-Sample_ID) %>% rowSums() %>% unique() # should be 100


## 3) Genus rarefied counts
genus_rar_counts <- asv_rar_counts %>% 
  # Transpose data
  column_to_rownames("Sample_ID") %>% 
  as.matrix() %>% 
  t() %>% 
  as.data.frame() %>%
  as_tibble(rownames = "ASV") %>% 
  left_join(taxonomic_annotation %>%
              select(ASV, genus_or_closest),
            by = "ASV") %>%
  select(ASV, genus_or_closest, everything()) %>% 
  # Aggregate counts
  select(-ASV) %>% 
  group_by(genus_or_closest) %>% 
  summarise(across(everything(), sum, na.rm = TRUE)) %>% 
  # Transpose data back
  column_to_rownames("genus_or_closest") %>%
  as.data.frame() %>%
  t() %>% 
  as_tibble(rownames = "Sample_ID") %>% 
  select(Sample_ID, !contains("Unclassified"), contains("Unclassified"))

genus_rar_counts %>% dim() # 371 x 205
genus_rar_counts %>% colnames() 
genus_rar_counts %>% select(contains("Unclassified")) %>% colnames() %>% length() # 45 unclassified taxa
204-45
# 204 taxa: 159 genera, 45 unclassified taxa
genus_rar_counts %>% select(-Sample_ID) %>% rowSums() %>% unique() # should be the same number > 10000 as above (10079)


## 4) Genus rarefied rel abu
genus_rar_relabu <- genus_rar_counts %>% 
    mutate(
    across(
      -Sample_ID,
      ~ .x / rowSums(across(-Sample_ID)) * 100
    )
  )

genus_rar_relabu %>% dim() # 371 x 205
genus_rar_relabu %>% select(-Sample_ID) %>% rowSums() %>% unique() # should be 100
```

## 4.2) Family level
```{r}
### for those ASVs unclassified as genus level, label them at "Unclassified + lowest-taxonomic-rank-assigned"

## 1) family counts
family_counts <- 
  asv_counts %>% 
  # Transpose data
  column_to_rownames("Sample_ID") %>% 
  as.matrix() %>% 
  t() %>% 
  as.data.frame() %>%
  as_tibble(rownames = "ASV") %>% 
  left_join(taxonomic_annotation %>%
              select(ASV, family_or_closest),
            by = "ASV") %>%
  select(ASV, family_or_closest, everything()) %>% 
  # Aggregate counts
  select(-ASV) %>% 
  group_by(family_or_closest) %>% 
  summarise(across(everything(), sum, na.rm = TRUE)) %>% 
  # Transpose data back
  column_to_rownames("family_or_closest") %>%
  as.data.frame() %>%
  t() %>% 
  as_tibble(rownames = "Sample_ID") %>% 
  select(Sample_ID, !contains("Unclassified"), contains("Unclassified"))

family_counts %>% dim() # 371 x 75
family_counts %>% colnames() 
family_counts %>% select(contains("Unclassified")) %>% colnames() %>% length() # 18 unclassified taxa
75-18
# 75 taxa: 57 families, 18 unclassified taxa

## 2) family rel abu
family_relabu <- family_counts %>% 
    mutate(
    across(
      -Sample_ID,
      ~ .x / rowSums(across(-Sample_ID)) * 100
    )
  )

family_relabu %>% dim() # 371 x 75
family_relabu %>% select(-Sample_ID) %>% rowSums() %>% unique() # should be 100


## 3) family rarefied counts
family_rar_counts <- asv_rar_counts %>% 
  # Transpose data
  column_to_rownames("Sample_ID") %>% 
  as.matrix() %>% 
  t() %>% 
  as.data.frame() %>%
  as_tibble(rownames = "ASV") %>% 
  left_join(taxonomic_annotation %>%
              select(ASV, family_or_closest),
            by = "ASV") %>%
  select(ASV, family_or_closest, everything()) %>% 
  # Aggregate counts
  select(-ASV) %>% 
  group_by(family_or_closest) %>% 
  summarise(across(everything(), sum, na.rm = TRUE)) %>% 
  # Transpose data back
  column_to_rownames("family_or_closest") %>%
  as.data.frame() %>%
  t() %>% 
  as_tibble(rownames = "Sample_ID") %>% 
  select(Sample_ID, !contains("Unclassified"), contains("Unclassified"))

family_rar_counts %>% dim() # 371 x 75
family_rar_counts %>% colnames() 
family_rar_counts %>% select(contains("Unclassified")) %>% colnames() %>% length() # 18 unclassified taxa
75-18
# 75 taxa: 57 families, 18 unclassified taxa
family_rar_counts %>% select(-Sample_ID) %>% rowSums() %>% unique() # should be the same number > 10000 as above (10079)


## 4) family rarefied rel abu
family_rar_relabu <- family_rar_counts %>% 
    mutate(
    across(
      -Sample_ID,
      ~ .x / rowSums(across(-Sample_ID)) * 100
    )
  )

family_rar_relabu %>% dim() # 371 x 75
family_rar_relabu %>% select(-Sample_ID) %>% rowSums() %>% unique() # should be 100
```

## 5) Calculate alpha diversity
Note: this is done from the rarefied relative abundances, at ASV level
```{r}
alpha_div <- tibble(
  Sample_ID = asv_rar_counts$Sample_ID,
  richness  = rowSums(asv_rar_counts %>% select(-Sample_ID) > 0),
  shannon   = vegan::diversity(
    asv_rar_counts %>% select(-Sample_ID) %>% as.data.frame(),
    "shannon")
  )

alpha_div %>% dim()
```

## 6) Calculate subject-means
```{r}
### ASV files
asv_counts_means <- asv_counts %>% 
  mutate(Subject_ID = str_split_i(Sample_ID, "-", 1), .keep = "unused") %>% 
  group_by(Subject_ID) %>% 
  summarize(across(where(is.numeric), ~mean(., na.rm = TRUE))) %>% 
  arrange(Subject_ID) %>% 
  mutate(across(everything(), ~ifelse(. == "NaN", NA, .)))

asv_relabu_means <- asv_relabu %>% 
  mutate(Subject_ID = str_split_i(Sample_ID, "-", 1), .keep = "unused") %>% 
  group_by(Subject_ID) %>% 
  summarize(across(where(is.numeric), ~mean(., na.rm = TRUE))) %>% 
  arrange(Subject_ID) %>% 
  mutate(across(everything(), ~ifelse(. == "NaN", NA, .)))
asv_relabu_means %>% select(-Subject_ID) %>% rowSums() %>% unique() # should be 100

asv_rar_counts_means <- asv_rar_counts %>% 
  mutate(Subject_ID = str_split_i(Sample_ID, "-", 1), .keep = "unused") %>% 
  group_by(Subject_ID) %>% 
  summarize(across(where(is.numeric), ~mean(., na.rm = TRUE))) %>% 
  arrange(Subject_ID) %>% 
  mutate(across(everything(), ~ifelse(. == "NaN", NA, .)))

asv_rar_relabu_means <- asv_rar_relabu %>% 
  mutate(Subject_ID = str_split_i(Sample_ID, "-", 1), .keep = "unused") %>% 
  group_by(Subject_ID) %>% 
  summarize(across(where(is.numeric), ~mean(., na.rm = TRUE))) %>% 
  arrange(Subject_ID) %>% 
  mutate(across(everything(), ~ifelse(. == "NaN", NA, .)))
asv_rar_relabu_means %>% select(-Subject_ID) %>% rowSums() %>% unique() # should be 100

### Genus files
genus_counts_means <- genus_counts %>% 
  mutate(Subject_ID = str_split_i(Sample_ID, "-", 1), .keep = "unused") %>% 
  group_by(Subject_ID) %>% 
  summarize(across(where(is.numeric), ~mean(., na.rm = TRUE))) %>% 
  arrange(Subject_ID) %>% 
  mutate(across(everything(), ~ifelse(. == "NaN", NA, .)))

genus_relabu_means <- genus_relabu %>% 
  mutate(Subject_ID = str_split_i(Sample_ID, "-", 1), .keep = "unused") %>% 
  group_by(Subject_ID) %>% 
  summarize(across(where(is.numeric), ~mean(., na.rm = TRUE))) %>% 
  arrange(Subject_ID) %>% 
  mutate(across(everything(), ~ifelse(. == "NaN", NA, .)))
genus_relabu_means %>% select(-Subject_ID) %>% rowSums() %>% unique() # should be 100

genus_rar_counts_means <- genus_rar_counts %>% 
  mutate(Subject_ID = str_split_i(Sample_ID, "-", 1), .keep = "unused") %>% 
  group_by(Subject_ID) %>% 
  summarize(across(where(is.numeric), ~mean(., na.rm = TRUE))) %>% 
  arrange(Subject_ID) %>% 
  mutate(across(everything(), ~ifelse(. == "NaN", NA, .)))

genus_rar_relabu_means <- genus_rar_relabu %>% 
  mutate(Subject_ID = str_split_i(Sample_ID, "-", 1), .keep = "unused") %>% 
  group_by(Subject_ID) %>% 
  summarize(across(where(is.numeric), ~mean(., na.rm = TRUE))) %>% 
  arrange(Subject_ID) %>% 
  mutate(across(everything(), ~ifelse(. == "NaN", NA, .)))
genus_rar_relabu_means %>% select(-Subject_ID) %>% rowSums() %>% unique() # should be 100



### Family files
family_counts_means <- family_counts %>% 
  mutate(Subject_ID = str_split_i(Sample_ID, "-", 1), .keep = "unused") %>% 
  group_by(Subject_ID) %>% 
  summarize(across(where(is.numeric), ~mean(., na.rm = TRUE))) %>% 
  arrange(Subject_ID) %>% 
  mutate(across(everything(), ~ifelse(. == "NaN", NA, .)))

family_relabu_means <- family_relabu %>% 
  mutate(Subject_ID = str_split_i(Sample_ID, "-", 1), .keep = "unused") %>% 
  group_by(Subject_ID) %>% 
  summarize(across(where(is.numeric), ~mean(., na.rm = TRUE))) %>% 
  arrange(Subject_ID) %>% 
  mutate(across(everything(), ~ifelse(. == "NaN", NA, .)))
family_relabu_means %>% select(-Subject_ID) %>% rowSums() %>% unique() # should be 100

family_rar_counts_means <- family_rar_counts %>% 
  mutate(Subject_ID = str_split_i(Sample_ID, "-", 1), .keep = "unused") %>% 
  group_by(Subject_ID) %>% 
  summarize(across(where(is.numeric), ~mean(., na.rm = TRUE))) %>% 
  arrange(Subject_ID) %>% 
  mutate(across(everything(), ~ifelse(. == "NaN", NA, .)))

family_rar_relabu_means <- family_rar_relabu %>% 
  mutate(Subject_ID = str_split_i(Sample_ID, "-", 1), .keep = "unused") %>% 
  group_by(Subject_ID) %>% 
  summarize(across(where(is.numeric), ~mean(., na.rm = TRUE))) %>% 
  arrange(Subject_ID) %>% 
  mutate(across(everything(), ~ifelse(. == "NaN", NA, .)))
family_rar_relabu_means %>% select(-Subject_ID) %>% rowSums() %>% unique() # should be 100




### Alpha diversity
alpha_div_means <- alpha_div %>% 
  mutate(Subject_ID = str_split_i(Sample_ID, "-", 1), .keep = "unused") %>% 
  group_by(Subject_ID) %>% 
  summarize(across(where(is.numeric), ~mean(., na.rm = TRUE))) %>% 
  arrange(Subject_ID) %>% 
  mutate(across(everything(), ~ifelse(. == "NaN", NA, .)))
```


# Save clean/processed data
```{r}
saveRDS(asv_counts,
        glue("{rds_dir}/{params$markdown_id}_asv_counts.rds"))
saveRDS(asv_relabu,
        glue("{rds_dir}/{params$markdown_id}_asv_relabu.rds"))
saveRDS(asv_rar_counts,
        glue("{rds_dir}/{params$markdown_id}_asv_rar_counts.rds"))
saveRDS(asv_rar_relabu,
        glue("{rds_dir}/{params$markdown_id}_asv_rar_relabu.rds"))


saveRDS(genus_counts,
        glue("{rds_dir}/{params$markdown_id}_genus_counts.rds"))
saveRDS(genus_relabu,
        glue("{rds_dir}/{params$markdown_id}_genus_relabu.rds"))
saveRDS(genus_rar_counts,
        glue("{rds_dir}/{params$markdown_id}_genus_rar_counts.rds"))
saveRDS(genus_rar_relabu,
        glue("{rds_dir}/{params$markdown_id}_genus_rar_relabu.rds"))


saveRDS(family_counts,
        glue("{rds_dir}/{params$markdown_id}_family_counts.rds"))
saveRDS(family_relabu,
        glue("{rds_dir}/{params$markdown_id}_family_relabu.rds"))
saveRDS(family_rar_counts,
        glue("{rds_dir}/{params$markdown_id}_family_rar_counts.rds"))
saveRDS(family_rar_relabu,
        glue("{rds_dir}/{params$markdown_id}_family_rar_relabu.rds"))



saveRDS(asv_counts_means,
        glue("{rds_dir}/{params$markdown_id}_asv_counts_means.rds"))
saveRDS(asv_relabu_means,
        glue("{rds_dir}/{params$markdown_id}_asv_relabu_means.rds"))
saveRDS(asv_rar_counts_means,
        glue("{rds_dir}/{params$markdown_id}_asv_rar_counts_means.rds"))
saveRDS(asv_rar_relabu_means,
        glue("{rds_dir}/{params$markdown_id}_asv_rar_relabu_means.rds"))


saveRDS(genus_counts_means,
        glue("{rds_dir}/{params$markdown_id}_genus_counts_means.rds"))
saveRDS(genus_relabu_means,
        glue("{rds_dir}/{params$markdown_id}_genus_relabu_means.rds"))
saveRDS(genus_rar_counts_means,
        glue("{rds_dir}/{params$markdown_id}_genus_rar_counts_means.rds"))
saveRDS(genus_rar_relabu_means,
        glue("{rds_dir}/{params$markdown_id}_genus_rar_relabu_means.rds"))



saveRDS(family_counts_means,
        glue("{rds_dir}/{params$markdown_id}_family_counts_means.rds"))
saveRDS(family_relabu_means,
        glue("{rds_dir}/{params$markdown_id}_family_relabu_means.rds"))
saveRDS(family_rar_counts_means,
        glue("{rds_dir}/{params$markdown_id}_family_rar_counts_means.rds"))
saveRDS(family_rar_relabu_means,
        glue("{rds_dir}/{params$markdown_id}_family_rar_relabu_means.rds"))



saveRDS(alpha_div,
        glue("{rds_dir}/{params$markdown_id}_alpha_diversity.rds"))
saveRDS(alpha_div_means,
        glue("{rds_dir}/{params$markdown_id}_alpha_diversity_means.rds"))




saveRDS(taxonomic_annotation,
        glue("{rds_dir}/{params$markdown_id}_taxonomic_annotation.rds"))
```


# Finalise
```{r}
sessionInfo()
```

