---
title: "Diet data analysis"
author: "Paula Rodriguez-Garcia"
date: "`r Sys.time()`"
output:
  html_document:
    output_dir: "html"
params: 
  markdown_id: "161" # can be: "161" or "161_female"
  # if "161_female", uncomment line 90, chunk 7
  methane_group: "Methane_group"
  cor_method: "spearman"
  threshold_p: 0.05
  threshold_fdr: 0.1

---


# Setup

## Knit options
```{r setup}
knitr::opts_chunk$set(warning = FALSE, message = FALSE,
                       root.dir = rprojroot::find_rstudio_root_file())
```

## Libraries
```{r}
library(tidyverse)
library(glue)
library(openxlsx)
library(ggpubr)
library(patchwork)
```

## Study paramters
```{r}
if (file.exists("../../study_parameters.R")) {
  source("../../study_parameters.R")
} else {
  source("study_parameters.R")
}

paulaR::show_rmd_params(params)
```


## Directories
```{r}
```

## Functions
```{r}
```



# Read data
## Meta
```{r}
meta_raw <- readRDS(meta_all_filename)
meta_means_raw <- readRDS(meta_means_filename)
```


# Prepare data

## Adjust diet variables per total kcal ingested and calculate macronutrient ratios
```{r}
diet_vars_all <- c("total_kcal",
                   "carbohydrate", "fat", "protein",
                   "fibre", 
                    "fruit", "vegetables", "whole_grain", "nuts"
                   )

diet_vars_to_adjust <- c("carbohydrate", "fat", "protein",
                   "fibre",
                   "fruit", "vegetables", "whole_grain", "nuts"
                   )

diet_vars_adjusted <- paste0(diet_vars_to_adjust, "_per_1000_kcal")


meta <- 
  meta_raw %>%
  ### !!!! FEMALE FILTER!! keep commented out
  # filter(sex == "female") %>% 
  #####
    select(Sample_ID, Subject_ID,
           !!sym(params$methane_group), 
           all_of(diet_vars_all),
           sex) %>% 
    mutate(across(all_of(diet_vars_to_adjust),
                  ~ .x / total_kcal * 1000,
                   .names = "{.col}_per_1000_kcal")
    ) %>% 
  # other vars
  mutate(fruit_vegetables = fruit + vegetables,
         fruit_vegetables_per_1000_kcal = fruit_per_1000_kcal + vegetables_per_1000_kcal)


meta_means <- meta %>% 
  # recalculate means as so many variables were added
  group_by(Subject_ID) %>% 
  summarize(across(where(is.numeric), ~mean(., na.rm = TRUE))) %>% 
  arrange(Subject_ID) %>% 
  mutate(across(everything(), ~ifelse(. == "NaN", NA, .))) %>% 
  left_join(meta_means_raw %>% select(Subject_ID, !!sym(params$methane_group), sex, ph_proximal_colon),
            by = "Subject_ID")
```

## Create the variable names
```{r}
diet_vars_imp <- c("total_kcal",
                   
                   diet_vars_to_adjust[!grepl("fruit|vegetables",
                                              diet_vars_to_adjust)],
                   "fruit_vegetables",
                   
                   diet_vars_adjusted[!grepl("fruit|vegetables",
                                              diet_vars_adjusted)],
                   "fruit_vegetables_per_1000_kcal")

diet_varnames_imp <- list(
                     "Total energy intake (kcal/day)",
                     
                     "Carbohydrate (g/day)",
                     "Fat (g/day)",
                     "Protein (g/day)",
                     "Fibre (g/day)",
                     "Whole grain (g/day)",
                     "Nuts (g/day)",
                     "Fruit & vegetables (g/day)",
                     
                     "Carbohydrate (g/1000 kcal)",
                     "Fat (g/1000 kcal)",
                     "Protein (g/1000 kcal)",
                     "Fibre (g/1000 kcal)",
                     "Whole grain (g/1000 kcal)",
                     "Nuts (g/1000 kcal)",
                     "Fruit & vegetables (g/1000 kcal)"
                     )

diet_var_mapping <- diet_varnames_imp
names(diet_var_mapping) <- diet_vars_imp

```

## Common aesthetics
```{r}
fontsize <- 14
```

## Order of the variables
```{r}
diet_vars_ordered <- c(
  "total_kcal",
  "carbohydrate",
  "carbohydrate_per_1000_kcal",
  "fat",
  "fat_per_1000_kcal",
  "protein",
  "protein_per_1000_kcal",
  "fibre",
  "fibre_per_1000_kcal",
  "fruit_vegetables",
  "fruit_vegetables_per_1000_kcal",
  "whole_grain", 
  "whole_grain_per_1000_kcal",
  "nuts", 
  "nuts_per_1000_kcal"
)
```


# Methane group comparison

## Wilcox test - subject-level means
```{r}
gc_res_means <-
  map(diet_vars_ordered, function(y){
  
    x <- params$methane_group
    
  
    test <- compare_means(
      as.formula(paste0("`", y, "` ~ `", x, "`")),
      meta_means,
      "wilcox.test",
      exact = FALSE) %>%
      mutate(
        response = y,
        mean_HMP = mean(meta_means[[y]][meta_means[[x]] == "HMP"],
                        na.rm = TRUE),
        sd_HMP   = sd(meta_means[[y]][meta_means[[x]] == "HMP"],
                      na.rm = TRUE),
        mean_LMP = mean(meta_means[[y]][meta_means[[x]] == "LMP"],
                        na.rm = TRUE),
        sd_LMP   = sd(meta_means[[y]][meta_means[[x]] == "LMP"],
                      na.rm = TRUE),
       label = case_when(
        p < 0.001 ~ "p < 0.001",
        p > 0.1   ~ "p > 0.1",
        TRUE ~ glue("p = {round(p, 3)}")
        ), .before = 1)

       
}) %>% bind_rows() %>% 
  mutate(FDR = p.adjust(p, "BH"), .after = "p")

gc_res_means
```

## Boxplots - means
```{r}
diet_boxplots_means <-
  map(diet_vars_ordered, function(y){
  
    x <- params$methane_group
    
    ylab  <- diet_var_mapping[[y]] %>% str_extract("(?<=\\().*(?=\\))")

    title <- diet_var_mapping[[y]]
    print(title)
    
    ylim  <- c(0, 
               paulaR::ceiling_signif(max(meta_means[[y]], na.rm = T), 0))
    
    if (ylim[2]- ylim[1] > 2000){
      breaks <- seq(ylim[1], ylim[2], 1000)
    } else if (ylim[2]- ylim[1] > 250){
      breaks <- seq(ylim[1], ylim[2], 100)
    } else  if (ylim[2]- ylim[1] > 100) {
      breaks <- seq(ylim[1], ylim[2], 50)
    } else if (ylim[2]- ylim[1] > 20) {
      breaks <- seq(ylim[1], ylim[2], 20)
    } else if (ylim[2]- ylim[1] > 10) {
      breaks <- seq(ylim[1], ylim[2], 5)
    } else if (ylim[2]- ylim[1] > 5) {
      breaks <- seq(ylim[1], ylim[2], 2)
    } else if (ylim[2]- ylim[1] > 2) {
      breaks <- seq(ylim[1], ylim[2], 1)
    } else {
      breaks <- seq(ylim[1], ylim[2], 0.5)
    }
      
      

    ppos <- mean(ylim[2], max(meta_means[[y]], na.rm = TRUE) )
    
    test <- compare_means(as.formula(paste0("`", y, "` ~ `", x, "`")),
                            meta_means,
                            "wilcox.test",
                            exact = FALSE) %>%
      mutate(label = case_when(
              p < 0.001 ~ "p < 0.001",
              p > 0.1   ~ "p > 0.1",
              TRUE ~ glue("p = {round(p, 3)}")
              ),
             y.position = ppos
             )

       
   meta_means %>% 
     select(!!sym(x), !!sym(y)) %>% 
     filter(!is.na(!!sym(x)), !is.na(!!sym(y))) %>% 
     ggplot(aes(x = !!sym(x), 
                y = !!sym(y),
                fill = !!sym(x))) +
     geom_boxplot(outlier.shape = NA,
                  color = "black") +
     geom_jitter(aes(fill = !!sym(x)), 
                 color = "black",
                 shape = 21,
                 size = 1.5,
                 height = 0,
                 stroke = 1.2,
                 alpha = 0.6) +
     scale_fill_manual(values = study_colors$methane) +
      stat_pvalue_manual(
        test,
        label = "label",
        tip.length = 0.01,
        size = fontsize/3
      ) +
     labs(title = title,
          x = "", 
          y = ylab, 
          ) +
    scale_y_continuous(limits = ylim,
                       breaks = breaks,
                       expand = expansion(mult = c(0.05, 0.15))) +
    paulaR::white_theme(font_size = fontsize) +
    theme(legend.position = "none",
          plot.title = ggtext::element_textbox_simple(
            fill = "gray80",        
            color = "black",       
            box.color = "black",    
            size = fontsize - 2.5,
            linetype = 1,
           linewidth = 0.75,         
            padding = margin(4,4,4,4) 
          ),
          axis.title.y = element_text(size = fontsize - 2))
       
}) %>% wrap_plots(ncol = 5) & 
  theme(plot.margin = margin(5, 10, 0, 10),
        panel.background = element_rect(fill = "white"),
        panel.border = element_rect(color = "black",
                                    fill = NA, 
                                    linewidth = 1.5)
        )

diet_boxplots_means
```

## Individual boxplots - means

### A) Energy intake
```{r}
fontsize <- 14
x <- params$methane_group
y <- "total_kcal"
ylab  <- diet_var_mapping[[y]] %>% str_extract("(?<=\\().*(?=\\))")
title <- diet_var_mapping[[y]] %>% str_split_i(" \\(", 1)
ylim  <- c(0, paulaR::ceiling_signif(max(meta_means[[y]], na.rm = T), 2))
breaks <- seq(ylim[1], ylim[2], 1000)
ppos <- mean(ylim[2], max(meta_means[[y]], na.rm = TRUE) )
test <- compare_means(as.formula(paste0("`", y, "` ~ `", x, "`")),
                        meta_means,
                        "wilcox.test",
                        exact = FALSE) %>%
  mutate(label = case_when(
          p < 0.001 ~ "p < 0.001",
          p > 0.1   ~ "p > 0.1",
          TRUE ~ glue("p = {round(p, 3)}")
          ),
         y.position = ppos
         )

       
energy_intake_boxplot <- meta_means %>% 
   select(!!sym(x), !!sym(y)) %>% 
   filter(!is.na(!!sym(x)), !is.na(!!sym(y))) %>% 
   ggplot(aes(x = !!sym(x), 
              y = !!sym(y),
              fill = !!sym(x))) +
   geom_boxplot(outlier.shape = NA,
                color = "black") +
   geom_jitter(aes(fill = !!sym(x)), 
               color = "black",
               shape = 21,
               size = 1.75,
               height = 0,
               stroke = 1.2,
               alpha = 0.6) +
   scale_fill_manual(values = study_colors$methane) +
    stat_pvalue_manual(
      test,
      label = "label",
      tip.length = 0.01,
      size = 4
    ) +
   labs(title = title,
        x = "", 
        y = ylab, 
        ) +
  scale_y_continuous(limits = ylim,
                     breaks = breaks,
                     expand = expansion(mult = c(0.05, 0.15))) +
  paulaR::white_theme(font_size = fontsize) +
  theme(legend.position = "none",
        plot.title = ggtext::element_textbox_simple(
          fill = "gray80",        
          color = "black",       
          box.color = "black",    
          size = fontsize - 2.5,
          linetype = 1,
         linewidth = 0.75,         
          padding = margin(4,4,4,4) 
        ),
        axis.title.y = element_text(size = fontsize - 2))

energy_intake_boxplot
```

### B) Macronutrients
```{r}
x <- params$methane_group

macronut_boxplots <- map(c("carbohydrate", "fat", "protein"), function(y){

  ylab  <- diet_var_mapping[[y]] %>% str_extract("(?<=\\().*(?=\\))")
  title <- diet_var_mapping[[y]] %>% str_split_i(" \\(", 1)
  
  if (y == "carbohydrate"){
    ylim  <- c(0, paulaR::ceiling_signif(max(meta_means[[y]], na.rm = T), 1)+10)
      breaks <- seq(ylim[1], ylim[2], 100)
  } else {
      ylim  <- c(0, paulaR::ceiling_signif(max(meta_means[[y]], na.rm = T), 1)+30)
      breaks <- seq(ylim[1], ylim[2], 50)
  }

  
  
  ppos <- mean(ylim[2], max(meta_means[[y]], na.rm = TRUE) )
  test <- compare_means(as.formula(paste0("`", y, "` ~ `", x, "`")),
                          meta_means,
                          "wilcox.test",
                          exact = FALSE) %>%
    mutate(label = case_when(
            p < 0.001 ~ "p < 0.001",
            p > 0.1   ~ "p > 0.1",
            TRUE ~ glue("p = {round(p, 3)}")
            ),
           y.position = ppos
           )
  
         
  meta_means %>% 
     select(!!sym(x), !!sym(y)) %>% 
     filter(!is.na(!!sym(x)), !is.na(!!sym(y))) %>% 
     ggplot(aes(x = !!sym(x), 
                y = !!sym(y),
                fill = !!sym(x))) +
     geom_boxplot(outlier.shape = NA,
                  color = "black") +
     geom_jitter(aes(fill = !!sym(x)), 
                 color = "black",
                 shape = 21,
                 size = 1.75,
                 height = 0,
                 stroke = 1.2,
                 alpha = 0.6) +
     scale_fill_manual(values = study_colors$methane) +
      stat_pvalue_manual(
        test,
        label = "label",
        tip.length = 0.01,
        size = 4
      ) +
     labs(title = title,
          x = "", 
          y = ylab, 
          ) +
    scale_y_continuous(limits = ylim,
                       breaks = breaks,
                       expand = expansion(mult = c(0.05, 0.15))) +
    paulaR::white_theme(font_size = fontsize) +
    theme(legend.position = "none",
          plot.title = ggtext::element_textbox_simple(
            fill = "gray80",        
            color = "black",       
            box.color = "black",    
            size = fontsize - 4,
            linetype = 1,
           linewidth = 0.75,         
            padding = margin(4,4,4,4) 
          ),
          axis.title.y = element_text(size = fontsize - 2))
}) %>% wrap_plots(ncol = 3) + plot_layout(axis = "collect_y")

macronut_boxplots
```

### C) Fibre-related boxplots
```{r}
x <- params$methane_group

fibre_related_boxplots <- map(c(
  "fibre_per_1000_kcal",
  "whole_grain_per_1000_kcal",
  "nuts_per_1000_kcal",
  "fruit_vegetables_per_1000_kcal"),
  function(y){

  ylab  <- diet_var_mapping[[y]] %>% str_extract("(?<=\\().*(?=\\))")
  title <- diet_var_mapping[[y]] %>% str_split_i(" \\(", 1)
  
  if (y == "fruit_vegetables_per_1000_kcal"){
    ylim  <- c(0, paulaR::ceiling_signif(max(meta_means[[y]], na.rm = T), 2)+10)
    breaks <- seq(ylim[1], ylim[2], 200)
  } else if (y == "fibre_per_1000_kcal") {
    ylim  <- c(0, paulaR::ceiling_signif(max(meta_means[[y]], na.rm = T), 1))
    breaks <- seq(ylim[1], ylim[2], 10)
  } else if (y == "nuts_per_1000_kcal") {
    ylim  <- c(0, paulaR::ceiling_signif(max(meta_means[[y]], na.rm = T)+2, 1))
    breaks <- seq(ylim[1], ylim[2], 10)
  } else {
    ylim  <- c(0, paulaR::ceiling_signif(max(meta_means[[y]], na.rm = T), 0)+2)
    breaks <- seq(ylim[1], ylim[2], 25)
  }

  
  
  ppos <- mean(ylim[2], max(meta_means[[y]], na.rm = TRUE) )
  test <- compare_means(as.formula(paste0("`", y, "` ~ `", x, "`")),
                          meta_means,
                          "wilcox.test",
                          exact = FALSE) %>%
    mutate(label = case_when(
            p < 0.001 ~ "p < 0.001",
            p > 0.1   ~ "p > 0.1",
            TRUE ~ glue("p = {round(p, 3)}")
            ),
           y.position = ppos
           )
  
         
  meta_means %>% 
     select(!!sym(x), !!sym(y)) %>% 
     filter(!is.na(!!sym(x)), !is.na(!!sym(y))) %>% 
     ggplot(aes(x = !!sym(x), 
                y = !!sym(y),
                fill = !!sym(x))) +
     geom_boxplot(outlier.shape = NA,
                  color = "black") +
     geom_jitter(aes(fill = !!sym(x)), 
                 color = "black",
                 shape = 21,
                 size = 1.75,
                 height = 0,
                 stroke = 1.2,
                 alpha = 0.6) +
     scale_fill_manual(values = study_colors$methane) +
      stat_pvalue_manual(
        test,
        label = "label",
        tip.length = 0.01,
        size = 4
      ) +
     labs(title = title,
          x = "", 
          y = ylab, 
          ) +
    scale_y_continuous(limits = ylim,
                       breaks = breaks,
                       expand = expansion(mult = c(0.05, 0.15))) +
    paulaR::white_theme(font_size = fontsize) +
    theme(legend.position = "none",
          plot.title = ggtext::element_textbox_simple(
            fill = "gray80",        
            color = "black",       
            box.color = "black",    
            size = fontsize - 4,
            linetype = 1,
           linewidth = 0.75,         
            padding = margin(4,4,4,4) 
          ),
          axis.title.y = element_text(size = fontsize - 2))
}) %>% wrap_plots(ncol = 4) + plot_layout(axis = "collect_y")
fibre_related_boxplots
```




# Potential confounders
## Sex
```{r}
if (!str_detect(params$markdown_id, "female")){

  vars <- c("total_kcal", "carbohydrate", "fat", "protein", 
            "fibre_per_1000_kcal",  
            "whole_grain_per_1000_kcal",
            "fruit_vegetables_per_1000_kcal")
  
  sex_res <- map(vars, function(y){
    f <- as.formula(paste(y," ~ ", params$methane_group, " + sex"))
    model <- lm(f, data = meta_means)
    p <- coefficients(summary(model))[2, "Pr(>|t|)"]
    tibble(response = y,
           adjusted_for = "sex",
           p = p)
  }) %>% bind_rows()
  
  y <- "protein"
  f <- as.formula(paste(y," ~ ", params$methane_group, " + sex"))
  model <- lm(f, data = meta_means)
  summary(model)
}
```

## pH and diet
```{r}
vars <- c("fibre_per_1000_kcal",  
          "whole_grain_per_1000_kcal",
          "nuts_per_1000_kcal",
          "fruit_vegetables_per_1000_kcal")

# group comparison
ph_fibre_gc_res <- map(vars, function(confounder){
  
  f <- as.formula(paste("ph_proximal_colon ~ ", params$methane_group, "+",  confounder))
  
  model <- lm(f, data = meta_means)
  
  p_x <- coefficients(summary(model))[2, "Pr(>|t|)"]
  p_conf <- coefficients(summary(model))[3, "Pr(>|t|)"]
  
  tibble(y = "ph_proximal_colon",
         x = params$methane_group,
         confounder = confounder,
         p_methane_group = p_x,
         p_confounder = p_conf)
}) %>% bind_rows()


# correlations
ph_fibre_cor_lm_res <- map(vars, function(x){
  
  f <- as.formula(paste("ph_proximal_colon ~ ", x))
  
  model <- lm(f, data = meta_means)
  
  p <- coefficients(summary(model))[2, "Pr(>|t|)"]
  
  tibble(y = "ph_proximal_colon",
         x = x,
         p = p)
  
}) %>% bind_rows()

ph_fibre_cor_kendall_res <- map(vars, function(x){
  
  test <- cor.test(
    meta_means$ph_proximal_colon,
    meta_means[[x]],
    method = "kendall",
    use = "complete.obs"
  )
  
  r <- test$estimate
  p <- test$p.value
  
  tibble(y = "ph_proximal_colon",
         x = x,
         r = r,
         p = p)
  
}) %>% bind_rows()

# interpret results
ph_fibre_gc_res 
# all P-values after adjusting are closer to 0.05 (non adjusted model = 0.017) but still below 0.05
ph_fibre_cor_kendall_res
# ph proximal colonic pH is not significantly corr to fibre, fruit, nuts except for whole grain (-0.21, P = 0.043)
```


# Some numbers for the paper
```{r}
# MACRONUTRIENTS --------------
gc_res_means %>% 
  filter(response %in% c("total_kcal", "protein", "fat", "carbohydrate")) %>% 
  # correct only for macronutrient comp.
  mutate(FDR = p.adjust(p, "BH")) %>% 
  select(response, p, FDR)

meta_means %>% 
  select(!!sym(params$methane_group), protein, total_kcal) %>% 
  group_by(!!sym(params$methane_group)) %>% 
  summarise(kcal_mean = mean(total_kcal, na.rm = TRUE),
            kcal_sd = sd(total_kcal, na.rm = TRUE),
            prot_mean = mean(protein, na.rm = TRUE),
            prot_sd = sd(protein, na.rm = TRUE)
            )

if (!str_detect(params$markdown_id, "female")){
  sex_res %>% 
    filter(response %in% c("total_kcal", "protein", "fat", "carbohydrate")) %>% 
    # correct only for macronutrient comp.
    mutate(FDR = p.adjust(p, "BH")) %>% 
    select(response, p, FDR)
} 

# FIBRE-RELATED FOODS ----------------
gc_res_means %>% 
  filter(response %in% c("fibre_per_1000_kcal", 
                         "whole_grain_per_1000_kcal",
                         "nuts_per_1000_kcal",
                         "fruit_vegetables_per_1000_kcal")) %>% 
  # correct only for macronutrient comp.
  mutate(FDR = p.adjust(p, "BH")) %>% 
  select(response, p, FDR)
```




# Save results
## Diet data extended
```{r}
saveRDS(meta %>% select(Sample_ID, Subject_ID, all_of(diet_vars_ordered)),
        glue("{rds_dir}/{params$markdown_id}_diet_data_extended.rds"))
saveRDS(meta_means %>% select(Subject_ID, all_of(diet_vars_ordered)),
        glue("{rds_dir}/{params$markdown_id}_diet_data_extended_means.rds"))
```

## Comparison results
```{r}
write.xlsx(gc_res_means,
             glue("{tab_dir}/{params$markdown_id}_means_diet_methane_group_comparison.xlsx"))
saveRDS(gc_res_means,
             glue("{rds_dir}/{params$markdown_id}_means_diet_methane_group_comparison.rds"))
```

## Figures
```{r}
format <- "png" # png # pdf


diet_boxplots_means
ggsave(glue("{fig_dir}/{params$markdown_id}_diet_boxplots_means.{format}"),
       bg = "white",
       width = 10,
       height = 8,
       dpi = 300)


energy_intake_boxplot
ggsave(glue("{fig_dir}/{params$markdown_id}_energy_intake_boxplot_means.{format}"),
       bg = "white",
       width = 2.8,
       height = 3,
       dpi = 300)
saveRDS(energy_intake_boxplot,
        glue("{rds_dir}/{params$markdown_id}_energy_intake_boxplot_means.rds"))

macronut_boxplots
ggsave(glue("{fig_dir}/{params$markdown_id}_macronutrients_boxplots_means.{format}"),
       bg = "white",
       width = 7,
       height = 3.5,
       dpi = 300)
saveRDS(macronut_boxplots,
        glue("{rds_dir}/{params$markdown_id}_macronutrients_boxplots_means.rds"))

fibre_related_boxplots
ggsave(glue("{fig_dir}/{params$markdown_id}_fibre_related_boxplots_means.{format}"),
       bg = "white",
       width = 7,
       height = 3,
       dpi = 300)
saveRDS(fibre_related_boxplots,
        glue("{rds_dir}/{params$markdown_id}_fibre_related_boxplots_means.rds"))
```


# Finalise
```{r}
sessionInfo()
```

