---
title: "Gut environment boxplots"
author: "Paula Rodriguez-Garcia"
date: "`r Sys.time()`"
output:
  html_document:
    output_dir: "html"
params: 
  markdown_id: "120" # "120_MTT2" means methane TT 2group
  methane_group: "Methane_group" # "Methane_TT_2group"
  
  cor_method_means: "spearman"
  stool_ph_var: "stool_ph_raw" 
---


# Setup

## Knit options
```{r setup}
knitr::opts_chunk$set(warning = FALSE, message = FALSE,
                       root.dir = rprojroot::find_rstudio_root_file())
```

## Libraries
```{r}
library(tibble)
library(dplyr)
library(magrittr)
library(tidyr)
library(readxl)
library(glue)
library(purrr)
library(patchwork)
library(ggplot2)
library(stringr)
library(ggpubr)
library(openxlsx)
library(flextable)
```

## Study paramters
```{r}
if (file.exists("../../study_parameters.R")) {
  source("../../study_parameters.R")
} else {
  source("study_parameters.R")
}

paulaR::show_rmd_params(params)
```


## Directories
```{r}
```

## Functions
```{r}
count_how_many_zero_per_column <- function (data){
    sapply(data, function(x) sum(x == 0, na.rm = T))
}
```


# Read data
## Meta
```{r}
meta_raw <- readRDS(meta_all_filename)
meta_means_raw <- readRDS(meta_means_filename)
```


# Prepare data

## Extract variables of interest
```{r}
gut_boxplots_vars <- c(
  "bloating", "constipation", "git_symptoms",
  "stool_freq", "bristol_scale", "bacterial_load_wet",
  params$stool_ph_var, "stool_energy", "stool_moisture", "cn_ratio")

gut_boxplots_varnames <- list("Bloating", "Constipation", "GIT symptoms",
                   "Stool frequency", "Stool consistency", "Bacterial load",
                   "Stool pH", "Energy density", "Stool moisture" ,"C:N ratio")

gut_boxplots_var_mapping <- gut_boxplots_varnames
names(gut_boxplots_var_mapping) <- gut_boxplots_vars

meta <- meta_raw %>% 
  select(Sample_ID, Subject_ID,
         !!sym(params$methane_group),
         all_of(gut_boxplots_vars)) %>% 
  filter(!is.na(!!sym(params$methane_group)))

meta_means <- meta_means_raw %>% 
  select(Subject_ID,
         !!sym(params$methane_group),
         all_of(gut_boxplots_vars)) %>% 
  filter(!is.na(!!sym(params$methane_group)))
```


# Set plot aesthetics
## Fontsize
```{r}
fontsize <- 14
```

## Y axis labels
```{r}
gut_boxplots_ylab_mapping <- list("Visual analog scale",
                       "Visual analog scale",
                       "Visual analog scale",
                       "Times per day",
                       "Bristol stool score",
                       "Cell counts/g wet feces",
                       "pH",
                       "J/g dry feces",
                       "Water content (%)",
                       "Carbon-to-nitrogen ratio"
                       )

names(gut_boxplots_ylab_mapping) <- gut_boxplots_vars
```

## Y axis aesthetics
```{r}
gut_boxplots_means_ylim1_mapping <- list(-1,
                       -1,
                       -1,
                       0,
                       1, # bss
                       2e10, #1e11,
                       6,
                       19000,
                       0.5, 
                       5)

names(gut_boxplots_means_ylim1_mapping) <- gut_boxplots_vars

gut_boxplots_means_ylim2_mapping <- c(6,
                       6,
                       6,
                       4,
                       7, # bss
                       2e11, #5e11,
                       7.5,
                       28000,
                       1, 
                       20)

names(gut_boxplots_means_ylim2_mapping) <- gut_boxplots_vars


gut_boxplots_means_break1_mapping <- list(0,
                       0,
                       0,
                       0,
                       1, # bss
                       2e10, #1e11,
                       6,
                       2000,
                       0.5,
                       5)

names(gut_boxplots_means_break1_mapping) <- gut_boxplots_vars

gut_boxplots_means_break2_mapping <- list(6,
                       6,
                       6,
                       4,
                       7, # bss
                      2e11, #  5e11,
                       7.5,
                       28000,
                       1,
                       20)

names(gut_boxplots_means_break2_mapping) <- gut_boxplots_vars


gut_boxplots_means_break3_mapping <- list(2,
                       2,
                       2,
                       1,
                       1, # bss
                      5e10, # 1e11,
                       0.5,
                       2000,
                       0.1, 
                       5)

names(gut_boxplots_means_break3_mapping) <- gut_boxplots_vars


gut_boxplots_ylim1_mapping <- list(-1,
                       -1,
                       -1,
                       0,
                       1, # bss
                      8e9, # 4e10,
                       4.5,
                       8000,
                       0.2, 
                       5)

names(gut_boxplots_ylim1_mapping) <- gut_boxplots_vars

gut_boxplots_ylim2_mapping <- c(10,
                       10,
                       10,
                       5,
                       7, # bss
                       4e11, # 1e12,
                       8.5,
                       28500,
                       1, 
                       20)

names(gut_boxplots_ylim2_mapping) <- gut_boxplots_vars



gut_boxplots_break1_mapping <- list(0,
                       0,
                       0,
                       0,
                       1, # bss
                       8e9, #4e10,
                       6,
                       2000,
                       0.5,
                       5)

names(gut_boxplots_break1_mapping) <- gut_boxplots_vars

gut_boxplots_break2_mapping <- list(6,
                       6,
                       6,
                       4,
                       7, # bss
                      4e11,# 1e12,
                       7.5,
                       28000,
                       1,
                       20)

names(gut_boxplots_break2_mapping) <- gut_boxplots_vars


gut_boxplots_break3_mapping <- list(2,
                       2,
                       2,
                       1,
                       1, # bss
                       5e10, # 2e11,
                       0.5,
                       2000,
                       0.1, 
                       5)

names(gut_boxplots_break3_mapping) <- gut_boxplots_vars


gut_boxplots_yperc_mapping <- list(waiver(),
                       waiver(),
                       waiver(),
                       waiver(),
                       waiver(), # bss
                       scientific_notation_fun,
                       waiver(),
                       scientific_notation_fun,
                       function(x) scales::percent(x, suffix = ""),
                       waiver())

names(gut_boxplots_yperc_mapping) <- gut_boxplots_vars

```

# Plot data
## Energy boxplot (means)
```{r}
x <- params$methane_group
y <- "stool_energy"
ylim <- c(19000, 28000)
ppos <- ylim[1] + ((ylim[2]-ylim[1]) * 0.9)
breaks <- seq(2e4, 2.8e4, 2e3)

f <- as.formula(paste0("scale(`", y, "`) ~ `", x, "`"))
model <- lm(f, meta_means)
pval <- as.numeric(summary(model)$coefficients[2, "Pr(>|t|)"])
test <- tibble(
  .y. = y,
  group1 = "HMP",
  group2 = "LMP",
  p = pval
) %>% 
  mutate(label = case_when(
    p < 0.001 ~ "p < 0.001",
    p > 0.1 ~ "p > 0.1",
    TRUE ~ glue("p = {round(p, 3)}")
    ),
    y.position = ppos
  )

fig_energy <- 
  meta_means %>% 
     select(!!sym(x), !!sym(y)) %>% 
     filter(!is.na(!!sym(x)), !is.na(!!sym(y))) %>% 
     ggplot(aes(x = !!sym(x), 
                y = !!sym(y),
                fill = !!sym(x))) + 
     geom_boxplot(outlier.shape = NA,
                  color = "black") +
     geom_jitter(aes(fill = !!sym(x)), 
                 color = "black",
                 shape = 21,
                 size = 1,
                 height = 0,
                 stroke = 1.2,
                 alpha = 0.3) +
     scale_fill_manual(values = study_colors$methane) +
    stat_pvalue_manual(
        test,
        label = "label",
        tip.length = 0.01,
        size = fontsize/3
      ) +
     labs(title = gut_boxplots_var_mapping[[y]],
          y = gut_boxplots_ylab_mapping[[y]],
          x = "",
          ) +
    scale_y_continuous(limits = ylim,
                       breaks = breaks,
                       labels = gut_boxplots_yperc_mapping[[y]],
                       expand = expansion(mult = c(0.05, 0.15))) +
    paulaR::white_theme(font_size = fontsize) +
    theme(legend.position = "none",
          plot.title = element_text(size = fontsize - 2))

fig_energy
```

## Microbial load boxplot (longitudinal)
```{r}
x <- params$methane_group
y <- "bacterial_load_wet"
ylim <- c(8.6e9, 2.5e11)
ppos <- ylim[1] + ((ylim[2]-ylim[1]) * 0.99)
breaks <- seq(5e10, 2.5e11, 5e10)

f <- as.formula(paste0("scale(`", y, "`) ~ `", x, "`+ (1|Subject_ID)"))

model <- lmerTest::lmer(f, data = meta)
pval <- as.numeric(summary(model)$coefficients[2, "Pr(>|t|)"])


test <- tibble(
  .y. = y,
  group1 = "LMP",
  group2 = "HMP",
  p = pval,
  y.position = ppos,
) %>% 
  mutate(label = case_when(
    p < 0.001 ~ "p < 0.001",
    p > 0.1 ~ "p > 0.1",
    TRUE ~ glue("p = {round(p, 3)}")
  ))

fig_load_long <- 
  meta %>% 
     select(!!sym(x), !!sym(y)) %>% 
     filter(!is.na(!!sym(x)), !is.na(!!sym(y))) %>% 
     ggplot(aes(x = !!sym(x), 
                y = !!sym(y),
                fill = !!sym(x))) + 
     geom_boxplot(outlier.shape = NA,
                  color = "black") +
     geom_jitter(aes(fill = !!sym(x)), 
                 color = "black",
                 shape = 21,
                 size = 1,
                 height = 0,
                 stroke = 1.2,
                 alpha = 0.1) +
     scale_fill_manual(values = study_colors$methane) +
    stat_pvalue_manual(
        test,
        label = "label",
        tip.length = 0.01,
        size = fontsize/3
      ) +
     labs(title = gut_boxplots_var_mapping[[y]],
          y = gut_boxplots_ylab_mapping[[y]],
          x = "",
          ) +
    scale_y_continuous(limits = ylim,
                       breaks = breaks,
                       labels = gut_boxplots_yperc_mapping[[y]],
                       expand = expansion(mult = c(0.05, 0.15))) +
    paulaR::white_theme(font_size = fontsize) +
    theme(legend.position = "none",
          plot.title = element_text(size = fontsize - 2))

fig_load_long
```

## Correlation energy and microbial load (means)
```{r}
x <- "stool_energy"
y <- "bacterial_load_wet"

xlim <- c(paulaR::trunc_signif(min(meta_means[[x]], na.rm = F), 3),
          paulaR::ceiling_signif(max(meta_means[[x]], na.rm = T), 3))
ylim <- c(paulaR::trunc_signif(min(meta_means[[y]], na.rm = F), 10),
          paulaR::ceiling_signif(max(meta_means[[y]], na.rm = T), 10))

test <- meta_means %>%
  summarise(
    r = cor(!!sym(x), !!sym(y), method = params$cor_method_means),
    p = cor.test(!!sym(x), !!sym(y), method = params$cor_method_means, exact = FALSE)$p.value
  ) %>%
  mutate(p_lab = ifelse(
    p < 0.001,
    "p < 0.001",
    glue("p = {round(p, 3)}")
    ),
    r_lab = glue("R = {round(r, 3)}"),
    label = glue("{r_lab}, {p_lab}"),
    y_pos = max(meta_means[[y]], na.rm = TRUE) * 0.95,
    x_pos = min(meta_means[[x]], na.rm = TRUE) * 1.08
   )

fig_energy_load_means <-
  meta_means %>% 
    ggplot(aes(x = !!sym(x),
               y = !!sym(y))
           ) +
    geom_point(aes(fill = !!sym(params$methane_group)), 
               shape = 21,
               stroke = 1.2,
               color = "black",
               alpha = 0.6,
               size = 3) +
    scale_fill_manual(values = study_colors$methane) +
    geom_smooth(method = "lm", color = "black", se = F) +
    annotate("text",
              x = test$x_pos,
              y = test$y_pos,
             label = test$label,
             size = fontsize/3,
             hjust = 0.5) +
    scale_x_continuous(limits = xlim,
                       breaks = seq(xlim[1], xlim[2], 2e3),
                       ) +
    scale_y_continuous(limits = ylim,
                       breaks = seq(ylim[1], ylim[2], 2e10),
                       labels = scientific_notation_fun
                       ) +
    labs(title = "",
         y = "Bacterial load\n(cell counts/g wet feces)",
         x = "Energy density\n(J/g dry feces)",
         fill = ""
    ) +
    paulaR::white_theme(font_size = fontsize) +
    theme(legend.position = "bottom",
          plot.margin = margin(5,5,5,5),
          axis.text.x = element_text(angle = 30, hjust = 1))

fig_energy_load_means
```



# Save results

## Boxplot energy and methane groups (means)
```{r}
format <- "png" # png # pdf
fig_energy
ggsave(glue("{fig_dir}/{params$markdown_id}_Energy_boxplot_means.{format}"),
       bg = "white",
       width = 3,
       height = 3,
       dpi = 300)

saveRDS(fig_energy, glue("{rds_dir}/{params$markdown_id}_Energy_boxplot_means.rds"))
```


## Boxplot microbial load and methane groups (longitudinal)
```{r}
format <- "png" # png # pdf
fig_load_long
ggsave(glue("{fig_dir}/{params$markdown_id}_Bacterial_load_boxplot_longitudinal.{format}"),
       bg = "white",
       width = 3,
       height = 3,
       dpi = 300)

saveRDS(fig_load_long, glue("{rds_dir}/{params$markdown_id}_Bacterial_load_boxplot_longitudinal.rds"))
```


## Correlation energy and microbial load (means)
```{r}
format <- "png" # png # pdf
fig_energy_load_means
ggsave(glue("{fig_dir}/{params$markdown_id}_Correlation_energy_bacterial_load_means.{format}"),
       bg = "white",
       width = 5,
       height = 4,
       dpi = 300)

saveRDS(fig_energy_load_means, glue("{rds_dir}/{params$markdown_id}_Correlation_energy_bacterial_load_means.rds"))
```


# Finalise
```{r}
sessionInfo()
```

