---
title: "Linear models and heatmaps: gut factors, HMP vs LMP; and adjusting for transit time"
author: "Paula Rodriguez-Garcia"
date: "`r Sys.time()`"
output:
  html_document:
    output_dir: "html"
params: 
  markdown_id: "121" # "121_MTT2" means methane TT 2group
  methane_group: "Methane_group" # "Methane_TT_2group"
  stool_ph_var: "stool_ph_raw"
  threshold_p: 0.05
  threshold_fdr: 0.1
---

# Setup

## Knit options
```{r setup}
knitr::opts_chunk$set(warning = FALSE, message = FALSE,
                       root.dir = rprojroot::find_rstudio_root_file())
```

## Libraries
```{r}
library(tibble)
library(dplyr)
library(magrittr)
library(tidyr)
library(readxl)
library(glue)
library(purrr)
library(patchwork)
library(ggplot2)
library(stringr)
library(ggpubr)
library(openxlsx)
library(flextable)
library(effsize)
```

## Study paramters
```{r}
if (file.exists("../../study_parameters.R")) {
  source("../../study_parameters.R")
} else {
  source("study_parameters.R")
}

paulaR::show_rmd_params(params)
```


## Directories
```{r}
```

## Functions
```{r}
```


# Read data
## Meta
```{r}
meta_raw <- readRDS(meta_all_filename)
meta_means_raw <- readRDS(meta_means_filename)
```


# Prepare data

## Select variables of interest
```{r}
gut_vars <- c(
  "bloating", "diarrhea", "constipation", "git_symptoms",
  "stool_freq", "bristol_scale", "bacterial_load_wet",
  params$stool_ph_var, "stool_energy", "stool_moisture", "cn_ratio")

gut_varnames <- list("Bloating", "Diarrhoea", "Constipation", "GIT symptoms",
                   "Stool frequency", "Stool consistency", "Microbial load",
                   "Stool pH", "Energy density", "Stool moisture", "C:N ratio")

gut_var_mapping <- gut_varnames
names(gut_var_mapping) <- gut_vars


tt_vars  <- c("GET", "SBTT", "CTT", "SLBTT", "WGTT")

meta <- meta_raw %>% 
  select(Subject_ID, !!sym(params$methane_group), all_of(gut_vars), all_of(tt_vars)) %>% 
  filter(!is.na(!!sym(params$methane_group)))

meta_means <- meta_means_raw %>% 
  select(Subject_ID, !!sym(params$methane_group), all_of(gut_vars), all_of(tt_vars)) %>% 
  filter(!is.na(!!sym(params$methane_group)))
```

## Get total HMP and LMP numbers
```{r}
# for means' data
group_counts_means <- meta_means %>% group_by(!!sym(params$methane_group)) %>% count()
total_hmp_means <- group_counts_means[[1,2]]
total_lmp_means <- group_counts_means[[2,2]]
total_all_means <- nrow(meta_means)


# for long data
group_counts <- meta %>% group_by(!!sym(params$methane_group)) %>% count()
total_hmp <- group_counts[[1,2]]
total_lmp <- group_counts[[2,2]]
total_all <- nrow(meta)

```

## Prepare combinations for adjustment
```{r}
combi <- expand_grid(y = gut_vars,
                     adjusted_by = tt_vars)
```


# Linear modelling
## Set which vars are subject-level (others are treated longitudinally)
```{r}
vars_for_means <- c("stool_energy", "cn_ratio") # subject-level
```

## No adjustment
```{r}
x <- params$methane_group

res_unadj <- map(gut_vars, function(y){
  
  # If subject-level variable, do a LM on the means data, without Subject ID
  if (y %in% vars_for_means){
    
    model_data <- meta_means %>%
        dplyr::select(all_of(c(x, y)), Subject_ID) %>% 
        tidyr::drop_na()
    
    subject_counts <- model_data %>% group_by(!!sym(x)) %>% count()
    n_hmp <- subject_counts[[1,2]]
    n_lmp <- subject_counts[[2,2]]
    n_total <- nrow(model_data)
    
    f <- as.formula(paste0("scale(`", y, "`) ~ `", x, "`"))
    
    model <- lm(f, data = model_data)
        
    pval <- as.numeric(summary(model)$coefficients[2, "Pr(>|t|)"])
    
    coef <- as.numeric(summary(model)$coefficients[2, "Estimate"])
      
    log2FC <- coef / log(2)
      
    se      <- as.numeric(summary(model)$coefficients[2, "Std. Error"])
    ci_low  <- coef - 1.96 * se
    ci_high <- coef + 1.96 * se
  
    ci_log <- coef + c(-1, 1) * 1.96 * se
    ci_log2 <- ci_log / log(2)
      

    tibble::tibble(response = gut_var_mapping[[y]],
               predictor = x,
               test = "linear model",
               adjusted_by = "",
               n_total = glue("{n_total}/{total_all_means}"),
               n_HMP = glue("{n_hmp}/{total_hmp_means}"),
               n_LMP = glue("{n_lmp}/{total_lmp_means}"),
               estimate = coef,
               log2FC = coef,
               CI_log2FC_lower = ci_log2[1],
               CI_log2FC_upper = ci_log2[2],
               p = pval,
               beta_confounder = NA,
               p_confounder = NA
               )
    
        
  # If longitudinal variable, use a linear mixed model
  } else {
  
    model_data <- meta %>%
        dplyr::select(all_of(c(x, y)), Subject_ID) %>% 
        tidyr::drop_na()
    
    subject_counts <- model_data %>% group_by(!!sym(x)) %>% count()
    n_hmp <- subject_counts[[1,2]]
    n_lmp <- subject_counts[[2,2]]
    n_total <- nrow(model_data)
    
    f <- as.formula(paste0("scale(`", y, "`) ~ `", x, "` + (1 | Subject_ID)"))
    model <- lmerTest::lmer(f, data = model_data)
    
    # extract pval
    pval <- as.numeric(summary(model)$coefficients[2, "Pr(>|t|)"])
    
    # extract log2fold change and CI
    coef <- as.numeric(summary(model)$coefficients[2, "Estimate"])
    log2FC <- coef / log(2)
    
    # extract 95% CI
    se <- as.numeric(summary(model)$coefficients[2, "Std. Error"])
    ci_log <- coef + c(-1, 1) * 1.96 * se
    ci_log2 <- ci_log / log(2)
    
    tibble::tibble(response = gut_var_mapping[[y]],
                   predictor = x,
                   test = "Linear model",
                   adjusted_by = "Subject_ID",
                   n_total = glue("{n_total}/{total_all}"),
                   n_HMP = glue("{n_hmp}/{total_hmp}"),
                   n_LMP = glue("{n_lmp}/{total_lmp}"),
                   estimate = coef,
                   log2FC = log2FC,
                   CI_log2FC_lower = ci_log2[1],
                   CI_log2FC_upper = ci_log2[2],
                   p = pval,
                   beta_confounder = NA,
                   p_confounder = NA
                   )
  }

}) %>% bind_rows() %>% 
  mutate(FDR = p.adjust(p, method = "BH"), .after = "p")
```

## Adjustment for transit time
```{r}
x <- params$methane_group

res_adj <-
  pmap(combi, function(y, adjusted_by){
    
    # If subject-level variable, do a LM on the means data, without Subject ID
    if (y %in% vars_for_means){
      
      model_data <- meta_means %>%
          dplyr::select(Subject_ID, all_of(c(x, y, adjusted_by))) %>% 
          tidyr::drop_na()
      
      subject_counts <- model_data %>% group_by(!!sym(x)) %>% count()
      n_hmp <- subject_counts[[1,2]]
      n_lmp <- subject_counts[[2,2]]
      n_total <- nrow(model_data)
    
      f <- as.formula(paste0("scale(`", y, "`) ~ `", x, "` + `", adjusted_by, "`"))
      model <- lm(f, data = model_data)
      
      # extract pval
      pval <- as.numeric(summary(model)$coefficients[2, "Pr(>|t|)"])
      
      # extract log2fold change and CI
      coef <- as.numeric(summary(model)$coefficients[2, "Estimate"])
      log2FC <- coef / log(2)
      
      # extract 95% CI
      se <- as.numeric(summary(model)$coefficients[2, "Std. Error"])
      ci_log <- coef + c(-1, 1) * 1.96 * se
      ci_log2 <- ci_log / log(2)
    
      coef_tt <- as.numeric(summary(model)$coefficients[3, "Estimate"])
      pval_tt <- as.numeric(summary(model)$coefficients[3, "Pr(>|t|)"])
    
    
      tibble::tibble(response = gut_var_mapping[[y]],
                     predictor = x,
                     test = "Linear model",
                     adjusted_by = glue("{adjusted_by}"),
                     n_total = glue("{n_total}/{total_all_means}"),
                     n_HMP = glue("{n_hmp}/{total_hmp_means}"),
                     n_LMP = glue("{n_lmp}/{total_lmp_means}"),
                    estimate = coef,
                     log2FC = log2FC,
                     CI_log2FC_lower = ci_log2[1],
                     CI_log2FC_upper = ci_log2[2],
                     p = pval,
                     beta_confounder = coef_tt,
                    p_confounder = pval_tt
      )
  
    # If longitudinal variable, use a linear mixed model
    } else {
  
      model_data <- meta %>%
          dplyr::select(Subject_ID, all_of(c(x, y, adjusted_by))) %>% 
          tidyr::drop_na()
      
      subject_counts <- model_data %>% group_by(!!sym(x)) %>% count()
      n_hmp <- subject_counts[[1,2]]
      n_lmp <- subject_counts[[2,2]]
      n_total <- nrow(model_data)
    
      f <- as.formula(paste0("scale(`", y, "`) ~ `", x, "` + `", adjusted_by, "` + (1 | Subject_ID)"))
      model <- lmerTest::lmer(f, data = model_data)
      
      # extract pval
      pval <- as.numeric(summary(model)$coefficients[2, "Pr(>|t|)"])
      
      # extract log2fold change and CI
      coef <- as.numeric(summary(model)$coefficients[2, "Estimate"])
      log2FC <- coef / log(2)
      
      # extract 95% CI
      se <- as.numeric(summary(model)$coefficients[2, "Std. Error"])
      ci_log <- coef + c(-1, 1) * 1.96 * se
      ci_log2 <- ci_log / log(2)
    
      coef_tt <- as.numeric(summary(model)$coefficients[3, "Estimate"])
      pval_tt <- as.numeric(summary(model)$coefficients[3, "Pr(>|t|)"])
    
    
      tibble::tibble(response = gut_var_mapping[[y]],
                     predictor = x,
                     test = "Linear mixed model",
                     adjusted_by = glue("{adjusted_by}, Subject_ID"),
                     n_total = glue("{n_total}/{total_all}"),
                     n_HMP = glue("{n_hmp}/{total_hmp}"),
                     n_LMP = glue("{n_lmp}/{total_lmp}"),
                    estimate = coef,
                     log2FC = log2FC,
                     CI_log2FC_lower = ci_log2[1],
                     CI_log2FC_upper = ci_log2[2],
                     p = pval,
                     beta_confounder = coef_tt,
                    p_confounder = pval_tt
                     )
    }

}) %>% bind_rows() %>% 
  group_by(adjusted_by) %>% 
  mutate(FDR = p.adjust(p, method = "BH"), .after = "p") %>% 
  ungroup()
```

## Combine all results
```{r}
res <- bind_rows(res_unadj, res_adj) %>% 
  arrange(response) %>% 
    # assign a variable type label
  mutate(response_type = case_when(
    response %in% c("Bloating", "Diarrhoea", "GIT symptoms", "Constipation") ~ "GIT symptoms",
    response %in% c("Stool frequency", "Stool consistency") ~ "Def. patterns",
    response %in% c("Microbial load", "Energy density", "Stool pH", "Stool moisture", "C:N ratio") ~ "Stool characteristics"
  )) %>% 
  mutate(response_type = factor(response_type, levels = c(
    "GIT symptoms", "Def. patterns", "Stool characteristics"
  ))) %>% 
  # readjust multiple testing
  group_by(adjusted_by, response_type) %>% 
   mutate(FDR = p.adjust(p, method = "BH"), .after = "p") %>% 
  ungroup() %>% 
    mutate(signif_p = get_significance_asterisks(p, asterisks_p), .after = "p") %>%
    mutate(signif_FDR = get_significance_asterisks(FDR, asterisks_fdr), .after = "FDR")
```

## Extra: p- and q-values for load and energy
```{r}
res %>% 
  filter(adjusted_by %in% c("", "Subject_ID"),
         response %in% c("Stool moisture", 
                         "Microbial load",
                         "Energy density")) %>% 
  select(response, adjusted_by, p, FDR) %>% 
  flextable()
```


# Heatmap of results
- do not use fold changes, but standardized β. This works directly with coef/estimate because the y var was standardized before fitting the model. 
- only show adjustment for WGTT

## Prepare data and aesthetics for heatmap
```{r}
res_data <- 
  res %>% 
  mutate(response = case_when(
    response == "GIT symptoms" ~ "Overall symptoms",
    TRUE ~ response
  )) %>% 
   filter(adjusted_by ==  "Subject_ID" |
          adjusted_by ==  "" |
         str_detect(adjusted_by, "WGTT")
         ) %>% 
  # assign a variable type label
  mutate(response_type = case_when(
    response %in% c("Bloating",  "Diarrhoea", "GIT symptoms", "Overall symptoms", "Constipation") ~ "GIT symptoms",
    response %in% c("Stool frequency", "Stool consistency") ~ "Def. patterns",
    response %in% c("Microbial load", "Energy density", "Stool pH", "Stool moisture", "C:N ratio") ~ "Stool characteristics"
  )) %>% 
  mutate(response_type = factor(response_type, levels = c(
    "GIT symptoms", "Def. patterns", "Stool characteristics"
  ))) %>% 
  # readjust multiple testing
  group_by(adjusted_by, response_type) %>% 
   mutate(FDR = p.adjust(p, method = "BH"), .after = "p") %>% 
  ungroup() %>% 
  mutate(signif_FDR = get_significance_asterisks(FDR, asterisks_fdr), .after = "FDR") %>%
  # rename adjustment label
  mutate(adjustment_label = case_when(
    str_detect(adjusted_by, "^Subject_ID$") ~ "Minimal adjustment",
     adjusted_by == "" ~ "Minimal adjustment",
    str_detect(adjusted_by, "WGTT") ~ "Adjusted for WGTT",
  ),
  adjustment_label = factor(adjustment_label, levels = c(
    "Minimal adjustment", 
    "Adjusted for WGTT"
  ))) %>% 
  # assign a "significance label"
  mutate(
    significance_type = case_when(
      FDR < params$threshold_fdr ~ "fdr",
      p < params$threshold_p & FDR >= params$threshold_fdr ~ "nominal",
      TRUE ~ "ns"
    )
  ) %>% 
  # reorder responses
  mutate(response = factor(response, levels = c(
    "Overall symptoms",
    "Bloating",
    "Constipation",
    "Diarrhoea",
    "Stool frequency",
    "Stool consistency",
    "Stool moisture",
    "Stool pH",
    "C:N ratio",
    "Microbial load",
    "Energy density"
  )))
  
  
  
  # color gradients
gc_color_max_value <- res_data %>% 
  pull(estimate) %>%
  abs() %>% max(na.rm = TRUE) %>% 
  paulaR::ceiling_signif(digits = 0)

gc_fill_scale <- scale_fill_gradient2(
  low = study_colors$methane[[1]],
  mid = "white",
  high = study_colors$methane[[2]],
  midpoint = 0,
  limits = c(-gc_color_max_value, gc_color_max_value),
  breaks = c(-gc_color_max_value, 0, gc_color_max_value),  
  guide = guide_colorbar(title.position = "top"),
  name = "Standardized β",
 labels = c(
    glue("{-gc_color_max_value}\nhigher\nin HMP"),
    "0",
    glue("{gc_color_max_value}\nhigher\nin LMP")
  )
)

fontsize <- 16
fontsize_x <- 14
```

## Horizontal heatmap
```{r}
hm <- 
  res_data %>% 
  ggplot(aes(x = response,
             y = adjustment_label)) +

  # heatmap tiles
  geom_tile(aes(fill = estimate), color = "grey20") +

  # circles nominal P
  geom_point(
    data = res_data %>% filter(significance_type == "nominal"),
    aes(x = response, 
        y = adjustment_label,
        shape = significance_type),
    color = "black",
    size = 3,
    stroke = 1,
    fill = NA,
    inherit.aes = FALSE,
    show.legend = FALSE
  ) +
  
  # circles FDR
  geom_point(
    data = res_data %>% filter(significance_type == "fdr"),
    aes(x = response, 
        y = adjustment_label,
        shape = significance_type),
    color = "black",
    size = 3,
    stroke = 1,
    fill = "black",                   
    inherit.aes = FALSE,
    show.legend = TRUE
  ) +

  # tile color scale
  gc_fill_scale +

  # shape scale
  scale_shape_manual(
    name = "",
    values = c(nominal = 1, fdr = 21),
    labels = c(nominal = "p < 0.05", fdr = "q < 0.1")
  ) +
  
  facet_wrap(~ response_type,
             scales = "free_x",
             space = "free_x") +

  labs(y = "HMP vs LMP", 
        x = NULL) +

  scale_x_discrete(position = "top") +
  scale_y_discrete(limits = rev) +
  
  paulaR::white_theme(font_size = fontsize) +
  theme(
    axis.text.x = element_text(angle = -45,
                               hjust = 1,
                               size = fontsize_x,
                               color = "black"),
    axis.title.y = ggtext::element_textbox_simple(
            fill = "gray80",        
            color = "black",       
            box.color = "black",    
            size = fontsize - 4,
            linetype = 1,
           linewidth = 0.75,         
            padding = margin(4,4,4,4),
            margin = margin(b = 5, t = 5),
            orientation = "left-rotated",  
            halign = 0.5,       
            valign = 0.5        
          ),
    axis.text.y = element_text(color = "black"),
    legend.title = element_text(size = fontsize-4, face = "bold"),
    legend.text = element_text(size = fontsize-4),
    legend.position = "bottom",
    legend.location = "horizontal",
    panel.grid = element_blank()
  )
  
hm 
```

## Vertical heatmap
```{r}
hm_vertical <-
  res_data %>% 
  ggplot(aes(y = response,
             x = adjustment_label)) +

  # heatmap tiles
  geom_tile(aes(fill = estimate), color = "grey20") +

  # circles nominal P
  geom_point(
    data = res_data %>% filter(significance_type == "nominal"),
    aes(y = response, 
        x = adjustment_label,
        shape = significance_type),
    color = "black",
    size = 3,
    stroke = 1,
    fill = NA,
    inherit.aes = FALSE,
    show.legend = FALSE
  ) +
  
  # circles FDR
  geom_point(
    data = res_data %>% filter(significance_type == "fdr"),
    aes(y = response, 
        x = adjustment_label,
        shape = significance_type),
    color = "black",
    size = 3,
    stroke = 1,
    fill = "black",                   
    inherit.aes = FALSE,
    show.legend = TRUE
  ) +

  # tile color scale
  gc_fill_scale +

  # shape scale
  scale_shape_manual(
    name = "",
    values = c(nominal = 21, fdr = 21),
    labels = c(nominal = "p < 0.05", fdr = "q < 0.1")
  ) +
  
  facet_wrap(~ response_type,
             scales = "free_y",
              space = "free_y",
             strip.position = "right") +

  labs(x = "HMP vs LMP", 
        y = NULL) +

  scale_x_discrete(position = "top") +
  scale_y_discrete(limits = rev) +
  
  paulaR::white_theme(font_size = fontsize) +
  

theme(
    axis.text.y = element_text(size = 12,
                               color = "black"),
    axis.title.x.top = ggtext::element_textbox_simple(
            fill = "gray80",        
            color = "black",       
            box.color = "black",    
            size = fontsize - 4,
            linetype = 1,
           linewidth = 0.75,         
            padding = margin(4,4,4,4),
            margin = margin(b = 5, t = 5),
            orientation = "upright",  
            halign = 0.5,       
            valign = 0.5        
          ),
    strip.text.y.right = element_text(size = 12),
    axis.text.x = element_text(color = "black", size = 12, angle = -30, hjust = 1),
    legend.title = element_text(size = fontsize-4, face = "bold"),
    legend.text = element_text(size = fontsize-4),
    legend.position = "bottom",
    legend.box = "vertical",
    legend.location = "vertical",
    panel.grid = element_blank()
  )

hm_vertical
```


# Save results


## Linear models
```{r}
saveRDS(res,
           glue("{rds_dir}/{params$markdown_id}_gut_factors_methane_group_tt_linear_models.rds"))


saveRDS(res,
           glue("{rds_dir}/{params$markdown_id}_gut_factors_methane_group_tt_linear_models.rds"))
```

## Heatmaps
```{r}
format <- "png"

hm
ggsave(glue("{fig_dir}/{params$markdown_id}_results_heatmap_horizontal.{format}"),
       bg = "white",
       width = 8,
       height = 4,
       dpi = 400
       )
saveRDS(hm, glue("{rds_dir}/{params$markdown_id}_results_heatmap_horizontal.rds"))

hm_vertical
ggsave(glue("{fig_dir}/{params$markdown_id}_results_heatmap_vertical.{format}"),
       bg = "white",
       width = 3,
       height = 8.5,
       dpi = 400
       )

saveRDS(hm_vertical, glue("{rds_dir}/{params$markdown_id}_results_heatmap_vertical.rds"))
```



# Finalise
```{r}
sessionInfo()
```

