---
title: "PCA gut environment"
author: "Paula Rodriguez-Garcia"
date: "`r Sys.time()`"
output:
  html_document:
    output_dir: "html"
params: 
  markdown_id: "122"
  methane_group: "Methane_group"
  stool_ph_var: "stool_ph_raw"
---

NOTES: 
- This markdown only uses the means' data, as for some variables, there are only 2 repeated measures, which would lead to a lot of NAs and therefore loss of information.


# Setup

## Knit options
```{r setup}
knitr::opts_chunk$set(warning = FALSE, message = FALSE,
                       root.dir = rprojroot::find_rstudio_root_file())
```

## Libraries
```{r}
library(tibble)
library(dplyr)
library(magrittr)
library(glue)
library(patchwork)
library(ggplot2)
library(ggpubr)
library(ggrepel)
```

## Study paramters
```{r}
if (file.exists("../../study_parameters.R")) {
  source("../../study_parameters.R")
} else {
  source("study_parameters.R")
}

paulaR::show_rmd_params(params)
```


## Directories
```{r}
```

## Functions
```{r}
count_how_many_zero_per_column <- function (data){
    sapply(data, function(x) sum(x == 0, na.rm = T))
}

get_rows_with_all_na <- function(data, vars) {
  data %>%
    dplyr::filter(dplyr::if_all(all_of(vars),
                                is.na))
}

scientific_notation_fun <- function(x) {
  parse(text = gsub(
    "e\\+?", "%*%10^",
    formatC(x, format = "e", digits = 1)
  ))
}
```


# Read data
## Meta
```{r}
meta_raw <- readRDS(meta_means_filename)
```


# Prepare data

## Check variables of interest
```{r}
pca_vars <- c(
  "bloating",
  "constipation",
  "git_symptoms",
  
  "stool_freq",
  "bristol_scale",
  "stool_moisture",
  params$stool_ph_var,
  "stool_energy",
  "bacterial_load_wet",
  
  "ph_duodenum",
  "ph_small_bowel",
  "ph_proximal_colon",
  "ph_distal_colon",
  "ph_sigmoid_colon",
  "ph_rectum",
  
  
  "SBTT",
  "CTT",
  #"WGTT" # not included because of the subjects with no WGTT data
  
  "fasting_H2",
  "archaea_inv_ct"
  #"smithii_inv_ct" # not included because of 31 subjects with non-detected smithii
  )

pca_varnames <- c(
  
  "Bloating",
  "Constipation",
  "GIT symptoms",
  
  "Stool frequency",
  "Stool consistency",
  "Stool moisture",
  "Stool pH", 
  "Microbial load",
  "Energy density",
  
  "Duodenal pH",
  "Small intestinal pH",
  "Proximal colonic pH",
  "Distal colonic pH",
  "Sigmoid colonic pH",
  "Rectal pH",
  
  "Small int. transit time",
  "Colonic transit time",
  #"Whole-gut transit time",
  
  "Breath hydrogen",
  "Archaea"
  #"M. smithii"
)


pca_vars_mapping <- pca_varnames
names(pca_vars_mapping) <- pca_vars


meta_raw %>%
  select(Subject_ID, params$methane_group, all_of(pca_vars)) %>%
  paulaR::count_how_many_na_per_column()
   
      
meta <- meta_raw %>% 
  select(Subject_ID, params$methane_group, all_of(pca_vars)) %>% 
  na.omit() 
```

# PCA
## Settings
```{r}
top_k <- 10
fontsize <- 14
loadings_factor <- 10

loadings_offset_y_pos <- 0.01
loadings_offset_y_neg <- -0.01

loadings_offset_x_pos <- 0.01
loadings_offset_x_neg <- -0.01
```

## Compute PCA
```{r}
pca <- prcomp(meta %>%
                select(-Subject_ID, -!!sym(params$methane_group)),
              scale = T)

scores <- meta %>% 
  select(Subject_ID, params$methane_group) %>% 
  bind_cols(data.frame(pca$x[ , 1:top_k]))

loadings <- pca$rotation[ , 1:top_k] %>% 
  as_tibble(rownames = "var") %>% 
  mutate(varname = pca_vars_mapping[var], .after = "var") %>% 
  mutate(
    y_offset = ifelse(PC2 > 0, loadings_offset_y_pos, loadings_offset_y_neg),
    x_offset = ifelse(PC1 > 0, loadings_offset_x_pos, loadings_offset_x_neg),
    label_x = PC1*loadings_factor + x_offset,
    label_y = PC2*loadings_factor + y_offset
  )

labels <- loadings$varname

importances <- summary(pca)
```

## Scores plot
```{r}
scores_plot <- 
  scores %>% 
  ggplot() + 
  geom_point(aes(PC1,
                 PC2,
                 color = !!sym(params$methane_group),
                 fill = !!sym(params$methane_group)
                 ),
             shape = 21, alpha = 0.5, size = 3) +
  geom_hline(yintercept = 0,
             color = "gray80",
             linetype = "dashed") +
  geom_vline(xintercept = 0,
             color = "gray80",
             linetype = "dashed")  +
  stat_ellipse(data = scores, 
               aes(x = PC1, y = PC2, fill = !!sym(params$methane_group)),
               geom = "polygon",
               alpha = 0.2) +
  scale_fill_manual(values = study_colors$methane) +
  scale_color_manual(values = study_colors$methane) +
  labs(x = glue("PC1 ({as.character(signif(importances$importance[2,1]*100, 3))}%)"),
       y = glue("PC2 ({as.character(signif(importances$importance[2,2]*100, 3))}%)")) +
  
  geom_segment(data = loadings %>% select(-var, -varname),
               aes(x = 0,
                   y = 0,
                   xend = PC1*loadings_factor,
                   yend = PC2*loadings_factor),
             color = "black",
             arrow = arrow(angle = 25, 
                           length = unit(4, "pt"))
             ) +
  geom_text_repel(data = loadings %>% select(-var, -varname),
                  aes(x = label_x,
                      y = label_y,
                      label = labels),
            segment.color = "gray30",
            size = fontsize/3,
            vjust = 1,
            hjust = 1,
            max.overlaps = Inf,
            segment.linetype = "dashed"
            ) +
  paulaR::white_theme(font_size = fontsize) +
  theme(legend.position = "none")

scores_plot
```

## PC1 boxplot
```{r}
ylim <- c(paulaR::ceiling_signif(min(scores$PC1), 0),
          paulaR::ceiling_signif(max(scores$PC1), 0))

breaks <- seq(ylim[1], ylim[2], 2)

test <- compare_means(as.formula(paste("PC1 ~", params$methane_group)),
                      scores,
                      "wilcox.test",
                      exact = FALSE) %>%
  mutate(label = case_when(
    p < 0.001 ~ "p < 0.001",
    p > 0.1 ~ "p > 0.1",
    TRUE ~ glue("p = {round(p, 3)}")
    ),
    y.position = max(scores$PC1)*1.05)

boxplot_pc1 <- 
  scores %>% 
  ggplot(aes(x = !!sym(params$methane_group),
             y = PC1,
             group = !!sym(params$methane_group))) +
  geom_boxplot(aes(fill = !!sym(params$methane_group)),
               alpha = 0.5, 
               outlier.shape = NA) +
  geom_jitter(aes(fill = !!sym(params$methane_group),
                  color = !!sym(params$methane_group)),
              shape = 21,
              alpha = 0.5,
              height = 0,
              size = 3)+
  scale_fill_manual(values = study_colors$methane) +
  scale_color_manual(values = study_colors$methane) +
  guides(color = "none") +
  stat_pvalue_manual(
        test,
        label = "label",
        tip.length = 0.01,
        size = fontsize/3
      ) +
  scale_y_continuous(limits = ylim, 
                       breaks = breaks) + 
  coord_flip() +
  labs(x = "",
       y = glue("PC1 ({as.character(signif(importances$importance[2,1]*100, 3))}%)"),
       fill = "") +
  paulaR::white_theme(font_size = fontsize) +
  theme(legend.position = c(0.98, 0.999),
        legend.justification = c("right", "top"),
        legend.title = element_blank(),
        legend.background = element_rect(fill = "white"))

boxplot_pc1
```


## PC2 boxplot (not saved)
```{r}
ylim <- c(paulaR::ceiling_signif(min(scores$PC2), 0),
          paulaR::ceiling_signif(max(scores$PC2), 0))

breaks <- seq(ylim[1], ylim[2], 2)

test <- compare_means(as.formula(paste("PC2 ~", params$methane_group)),
                      scores,
                      "wilcox.test",
                      exact = FALSE) %>%
  mutate(label = case_when(
    p < 0.001 ~ "p < 0.001",
    p > 0.1 ~ "p > 0.1",
    TRUE ~ glue("p = {round(p, 3)}")
    ),
    y.position = max(scores$PC1)*1.05)

boxplot_pc2 <- 
  scores %>% 
  ggplot(aes(x = !!sym(params$methane_group),
             y = PC2,
             group = !!sym(params$methane_group))) +
  geom_boxplot(aes(fill = !!sym(params$methane_group)),
               alpha = 0.5, 
               outlier.shape = NA) +
  geom_jitter(aes(fill = !!sym(params$methane_group),
                  color = !!sym(params$methane_group)),
              shape = 21,
              alpha = 0.5,
              height = 0,
              size = 3)+
  scale_fill_manual(values = study_colors$methane) +
  scale_color_manual(values = study_colors$methane) +
  guides(color = "none") +
  stat_pvalue_manual(
        test,
        label = "label",
        tip.length = 0.01,
        size = fontsize/3
      ) +
  scale_y_continuous(limits = ylim, 
                       breaks = breaks) + 
  coord_flip() +
  labs(x = "",
       y = glue("PC2 ({as.character(signif(importances$importance[2,2]*100, 3))}%)"),
       fill = "") +
  paulaR::white_theme(font_size = fontsize) +
  theme(legend.position = c(0.98, 0.999),
        legend.justification = c("right", "top"),
        legend.title = element_blank(),
        legend.background = element_rect(fill = "white"))

boxplot_pc2
```

## Join both (PCA and PC1 boxplot)
```{r}
pca_joined <- scores_plot / boxplot_pc1 + 
  plot_layout(heights = c(2, 1)) +
  plot_annotation(tag_levels = "A")

pca_joined
```

# Save results

## Subfigures
```{r}
format <- "png" # png # pdf

scores_plot
ggsave(glue("{fig_dir}/{params$markdown_id}_Gut_environment_PCA_scores.{format}"),
       bg = "white",
       width = 8,
       height = 6,
       dpi = 300)

boxplot_pc1
ggsave(glue("{fig_dir}/{params$markdown_id}_Gut_environment_PC1_boxplot.{format}"),
       bg = "white",
       width = 8,
       height = 3,
       dpi = 300)
```

## Joined Figure (supplementary material)
```{r}
format <- "png" # png # pdf

pca_joined
ggsave(glue("{fig_dir}/{params$markdown_id}_Gut_environment_PCA.{format}"),
       bg = "white",
       width = 8,
       height = 8,
       dpi = 300)
```




# Finalise
```{r}
sessionInfo()
```

