---
title: "SmartPill data: pH and TT across HMP and LMP"
author: "Paula Rodriguez-Garcia"
date: "`r Sys.time()`"
output:
  html_document:
    output_dir: "html"
params: 
  markdown_id: "123" # "123_MTT2" means methane TT 2group
  methane_group: "Methane_group" # "Methane_TT_2group"
  stool_ph_var: "stool_ph_raw" 
---

# Setup

## Knit options
```{r setup}
knitr::opts_chunk$set(warning = FALSE, message = FALSE,
                       root.dir = rprojroot::find_rstudio_root_file())
```

## Libraries
```{r}
library(dplyr)
library(magrittr)
library(tidyr)
library(glue)
library(purrr)
library(patchwork)
library(ggplot2)
library(ggpubr)
library(openxlsx)
library(flextable)
```

## Study paramters
```{r}
if (file.exists("../../study_parameters.R")) {
  source("../../study_parameters.R")
} else {
  source("study_parameters.R")
}

paulaR::show_rmd_params(params)
```


## Directories
```{r}
```

## Functions
```{r}
count_how_many_zero_per_column <- function (data){
    sapply(data, function(x) sum(x == 0, na.rm = T))
}

get_rows_with_all_na <- function(data, vars) {
  data %>%
    dplyr::filter(dplyr::if_all(all_of(vars),
                                is.na))
}

scientific_notation_fun <- function(x) {
  parse(text = gsub(
    "e\\+?", "%*%10^",
    formatC(x, format = "e", digits = 1)
  ))
}
```


# Read data
## Meta
```{r}
meta_raw <- readRDS(meta_means_filename)
```


# Prepare data

## Extract variables of interest
```{r}
fig_ph_vars <- c("ph_duodenum",
  "ph_small_bowel",
  "ph_proximal_colon",
  "ph_distal_colon",
  "ph_sigmoid_colon",
  "ph_rectum")

fig_ph_varnames <- c("Duodenum",
                    "Small intestine",
                    "Proximal colon",
                    "Distal colon",
                    "Sigmoid colon",
                    "Rectum")

fig_tt_vars <- c("GET", "SBTT", "CTT", "SLBTT", "WGTT")

fig_tt_varnames <- c("Gastric emptying time",
  "Small bowel transit time",
                    "Colonic transit time",
                    "Intestinal transit time",
                    "Whole-gut transit time")

other_vars <- c("ph_colon",
                params$stool_ph_var # fecal pH
                )

other_varnames <- c("Colonic pH",
                    "Fecal pH") 


all_vars_mapping <- c(fig_ph_varnames, 
                      fig_tt_varnames, 
                      other_varnames)

names(all_vars_mapping) <- c(fig_ph_vars, 
                      fig_tt_vars, 
                      other_vars)


meta <- meta_raw %>% 
  select(Subject_ID,
         !!sym(params$methane_group),
         all_of(c(fig_ph_vars,
                  fig_tt_vars,
                  other_vars))
         ) %>% 
  filter(!is.na(!!sym(params$methane_group)))


paulaR::count_how_many_na_per_column(meta)
# 11 subjects with all missing data (did not agree to smartpill)
# 14-15 subjects without colonic pH
# 20 subjects without WGTT

# Identify the 11 subjects without smartpill data at all
ids_no_smartpill <- meta %>%
  get_rows_with_all_na(fig_ph_vars) %>% 
  pull(Subject_ID)
```


## Aesthetics, general
```{r}
fontsize <- 14
```

# Figure pH trajectories
## Prep data for fig pH trajectories
```{r}
fig_ph_data <- meta %>%
  select(Subject_ID,
         !!sym(params$methane_group),
         fig_ph_vars) %>% 
  # Remove only the 11 subjects with no smartpill data at all
   filter(!if_all(c(fig_ph_vars), is.na)) %>% 
  rename(!!!setNames(fig_ph_vars, fig_ph_varnames)) %>% 
  pivot_longer(-c(Subject_ID, !!sym(params$methane_group)),
               names_to = "segment", 
               values_to = "pH") %>% 
  filter(!is.na(pH)) %>% 
  mutate(segment = factor(segment,
                          levels = fig_ph_varnames),
         !!sym(params$methane_group) := factor(!!sym(params$methane_group)))

fig_ph_mean_data <- fig_ph_data %>%
  group_by(!!sym(params$methane_group), segment) %>%
  summarise(mean_pH = mean(pH, na.rm = TRUE)) %>% 
  filter(segment %in% fig_ph_varnames)

fig_ph_pval <- fig_ph_data %>% 
  group_by(segment) %>%
  summarise(
    p = wilcox.test(pH ~ !!sym(params$methane_group),
                    exact = F)$p.value) %>%
    mutate(label = case_when(
    p < 0.001 ~ "p < 0.001",
    p > 0.1 ~ "",
    TRUE ~ glue("* (p = {round(p, 3)})")
    )) %>% 
  mutate(y_position = max(fig_ph_data$pH, na.rm = TRUE) - 1.3
)

fig_ph_pval
```

## Plot fig pH trajectories
```{r}
ylim <- c(paulaR::trunc_signif(min(meta %>% select(fig_ph_vars), na.rm = T), 0),
          paulaR::ceiling_signif(max(meta %>% select(fig_ph_vars), na.rm = T), 0))
breaks <- seq(ylim[1], ylim[2], 1)

fig_ph <- 
fig_ph_data %>% 
  ggplot(aes(x = segment, 
             y = pH,
             group = Subject_ID, 
             color = !!sym(params$methane_group))) +
  geom_point(alpha = 0.6) +  
  geom_line(aes(group = Subject_ID), alpha = 0.3)  +  
  geom_line(data = fig_ph_mean_data, 
            aes(group = !!sym(params$methane_group),
                y = mean_pH),
            size = 2.5) +  
  geom_text(
    data = fig_ph_pval,
    aes(
      x = segment,
      y = y_position,
      label = label
    ),
    inherit.aes = FALSE,
    angle = 45,
    size = 5,
    vjust = 0.5,
    hjust = 0
  ) + 
  scale_color_manual(values = c(study_colors$methane[[1]], 
                                study_colors$methane[[2]])) +
  scale_y_continuous(limits = ylim,
                     breaks = breaks) +
  labs(x = "", y = "pH", color = "") +
  paulaR::white_theme(font_size = fontsize) +
  theme(axis.text.x = element_text(angle = 30,
                                   hjust = 1),
        legend.position = c(0.98, 0.03),
        legend.justification = c("right", "bottom"),
        legend.title = element_blank(),
        legend.background = element_rect(fill = "white")
        ) 

fig_ph
```

# Boxplots transit time
## Boxplot SBTT
```{r}
x <- params$methane_group
y <- "SBTT"
ylim <- c(0, paulaR::ceiling_signif(max(meta[[y]], na.rm = T), 1))
ppos <- ylim[1] + ((ylim[2]-ylim[1]) * 0.9)
breaks <- seq(ylim[1], ylim[2], 5)

test <- compare_means(as.formula(paste(y, "~", x)),
                      meta,
                      "wilcox.test",
                      exact = FALSE) %>%
  mutate(label = case_when(
    p < 0.001 ~ "p < 0.001",
    p > 0.1 ~ "p > 0.1",
    TRUE ~ glue("p = {round(p, 3)}")
    ),
    y.position = ppos
    )

fig_SBTT <- 
  meta %>% 
     select(!!sym(x), !!sym(y)) %>% 
     filter(!is.na(!!sym(x)), !is.na(!!sym(y))) %>% 
     ggplot(aes(x = !!sym(x), 
                y = !!sym(y),
                fill = !!sym(x))) + 
     geom_boxplot(outlier.shape = NA,
                  color = "black") +
     geom_jitter(aes(fill = !!sym(x)), 
                 color = "black",
                 shape = 21,
                 size = 3,
                 height = 0,
                 stroke = 1.2,
                 alpha = 0.6) +
     scale_fill_manual(values = study_colors$methane) +
    stat_pvalue_manual(
        test,
        label = "label",
        tip.length = 0.01,
        size = fontsize/3
      ) +
     labs(title = all_vars_mapping[[y]],
          y = "Hours",
          x = "",
          ) +
    scale_y_continuous(limits = ylim,
                       breaks = breaks,
                       expand = expansion(mult = c(0.05, 0.15))) +
    paulaR::white_theme(font_size = fontsize) +
    theme(legend.position = "none",
          plot.title = element_text(size = fontsize - 2))

fig_SBTT
```

## Boxplot CTT
```{r}
x <- params$methane_group
y <- "CTT"
ylim <- c(0, paulaR::ceiling_signif(max(meta[[y]], na.rm = T), 1)+10)
ppos <- ylim[1] + ((ylim[2]-ylim[1]) * 0.95)
breaks <- seq(ylim[1], ylim[2], 20)

test <- compare_means(as.formula(paste(y, "~", x)),
                      meta,
                      "wilcox.test",
                      exact = FALSE) %>%
  mutate(label = case_when(
    p < 0.001 ~ "p < 0.001",
    p > 0.1 ~ "p > 0.1",
    TRUE ~ glue("p = {round(p, 3)}")
    ),
    y.position = ppos
    )

fig_CTT <- 
  meta %>% 
     select(!!sym(x), !!sym(y)) %>% 
     filter(!is.na(!!sym(x)), !is.na(!!sym(y))) %>% 
     ggplot(aes(x = !!sym(x), 
                y = !!sym(y),
                fill = !!sym(x))) + 
     geom_boxplot(outlier.shape = NA,
                  color = "black") +
     geom_jitter(aes(fill = !!sym(x)), 
                 color = "black",
                 shape = 21,
                 size = 3,
                 height = 0,
                 stroke = 1.2,
                 alpha = 0.6) +
     scale_fill_manual(values = study_colors$methane) +
    stat_pvalue_manual(
        test,
        label = "label",
        tip.length = 0.01,
        size = fontsize/3
      ) +
     labs(title = all_vars_mapping[[y]],
          y = "Hours",
          x = "",
          ) +
    scale_y_continuous(limits = ylim,
                       breaks = breaks,
                       expand = expansion(mult = c(0.05, 0.15))) +
    paulaR::white_theme(font_size = fontsize) +
    theme(legend.position = "none",
          plot.title = element_text(size = fontsize - 2))

fig_CTT
```

## Boxplot SLBTT
```{r}
x <- params$methane_group
y <- "SLBTT"
ylim <- c(0, paulaR::ceiling_signif(max(meta[[y]], na.rm = T), 1)+10)
ppos <- ylim[1] + ((ylim[2]-ylim[1]) * 0.95)
breaks <- seq(ylim[1], ylim[2], 20)

test <- compare_means(as.formula(paste(y, "~", x)),
                      meta,
                      "wilcox.test",
                      exact = FALSE) %>%
  mutate(label = case_when(
    p < 0.001 ~ "p < 0.001",
    p > 0.1 ~ "p > 0.1",
    TRUE ~ glue("p = {round(p, 3)}")
    ),
    y.position = ppos
    )

fig_SLBTT <- 
  meta %>% 
     select(!!sym(x), !!sym(y)) %>% 
     filter(!is.na(!!sym(x)), !is.na(!!sym(y))) %>% 
     ggplot(aes(x = !!sym(x), 
                y = !!sym(y),
                fill = !!sym(x))) + 
     geom_boxplot(outlier.shape = NA,
                  color = "black") +
     geom_jitter(aes(fill = !!sym(x)), 
                 color = "black",
                 shape = 21,
                 size = 3,
                 height = 0,
                 stroke = 1.2,
                 alpha = 0.6) +
     scale_fill_manual(values = study_colors$methane) +
    stat_pvalue_manual(
        test,
        label = "label",
        tip.length = 0.01,
        size = fontsize/3
      ) +
     labs(title = all_vars_mapping[[y]],
          y = "Hours",
          x = "",
          ) +
    scale_y_continuous(limits = ylim,
                       breaks = breaks,
                       expand = expansion(mult = c(0.05, 0.15))) +
    paulaR::white_theme(font_size = fontsize) +
    theme(legend.position = "none",
          plot.title = element_text(size = fontsize - 2))

fig_SLBTT
```

## Boxplot WGTT
```{r}
x <- params$methane_group
y <- "WGTT"
ylim <- c(0, paulaR::ceiling_signif(max(meta[[y]], na.rm = T), 1))
ppos <- ylim[1] + ((ylim[2]-ylim[1]) * 0.95)
breaks <- seq(ylim[1], ylim[2], 20)

test <- compare_means(as.formula(paste(y, "~", x)),
                      meta,
                      "wilcox.test",
                      exact = FALSE) %>%
  mutate(label = case_when(
    p < 0.001 ~ "p < 0.001",
    p > 0.1 ~ "p > 0.1",
    TRUE ~ glue("p = {round(p, 3)}")
    ),
    y.position = ppos
    )

fig_WGTT <- 
  meta %>% 
     select(!!sym(x), !!sym(y)) %>% 
     filter(!is.na(!!sym(x)), !is.na(!!sym(y))) %>% 
     ggplot(aes(x = !!sym(x), 
                y = !!sym(y),
                fill = !!sym(x))) + 
     geom_boxplot(outlier.shape = NA,
                  color = "black") +
     geom_jitter(aes(fill = !!sym(x)), 
                 color = "black",
                 shape = 21,
                 size = 3,
                 height = 0,
                 stroke = 1.2,
                 alpha = 0.6) +
     scale_fill_manual(values = study_colors$methane) +
    stat_pvalue_manual(
        test,
        label = "label",
        tip.length = 0.01,
        size = fontsize/3
      ) +
     labs(title = all_vars_mapping[[y]],
          y = "Hours",
          x = "",
          ) +
    scale_y_continuous(limits = ylim,
                       breaks = breaks,
                       expand = expansion(mult = c(0.05, 0.15))) +
    paulaR::white_theme(font_size = fontsize) +
    theme(legend.position = "none",
          plot.title = element_text(size = fontsize - 2))

fig_WGTT
```

# Other figures
## Boxplot prox colonic pH
```{r}
x <- params$methane_group
y <- "ph_proximal_colon"
ylim <- c(paulaR::trunc_signif(min(meta[[y]], na.rm = T), 0), 
          paulaR::ceiling_signif(max(meta[[y]], na.rm = T), 0))
ppos <- 8
breaks <- seq(ylim[1], ylim[2], 0.5)

test <- compare_means(as.formula(paste(y, "~", x)),
                      meta,
                      "wilcox.test",
                      exact = FALSE) %>%
  mutate(label = case_when(
    p < 0.001 ~ "p < 0.001",
    p > 0.1 ~ "p > 0.1",
    TRUE ~ glue("p = {round(p, 3)}")
    ),
    y.position = ppos
    )

fig_pcph <- 
  meta %>% 
     select(!!sym(x), !!sym(y)) %>% 
     filter(!is.na(!!sym(x)), !is.na(!!sym(y))) %>% 
     ggplot(aes(x = !!sym(x), 
                y = !!sym(y),
                fill = !!sym(x))) + 
     geom_boxplot(outlier.shape = NA,
                  color = "black") +
     geom_jitter(aes(fill = !!sym(x)), 
                 color = "black",
                 shape = 21,
                 size = 3,
                 height = 0,
                 stroke = 1.2,
                 alpha = 0.6) +
     scale_fill_manual(values = study_colors$methane) +
    stat_pvalue_manual(
        test,
        label = "label",
        tip.length = 0.01,
        size = fontsize/3
      ) +
     labs(title = "Proximal colonic pH",
          y = "Hours",
          x = "",
          ) +
    scale_y_continuous(limits = ylim,
                       breaks = breaks,
                       expand = expansion(mult = c(0.05, 0.15))) +
    paulaR::white_theme(font_size = fontsize) +
    theme(legend.position = "none",
          plot.title = element_text(size = fontsize - 2))

fig_pcph
```

## Other pH/TT boxplots
```{r}
other_pH_tt_boxplots <- map(other_vars, function(y){
  
  x <- params$methane_group
  
  # ylim <- c(paulaR::trunc_signif(min(meta[[y]], na.rm = T), 0), 
  #           paulaR::ceiling_signif(max(meta[[y]], na.rm = T), 0))
  ylim <- c(5, 9)
  breaks <- seq(ylim[1], ylim[2], 1)
    

  
  ppos <- ylim[1] + ((ylim[2]-ylim[1]) * 0.95)


  test <- compare_means(as.formula(paste0("`", y, "`~", x)),
                        meta,
                        "wilcox.test",
                        exact = FALSE) %>%
    mutate(label = case_when(
      p < 0.001 ~ "p < 0.001",
      p > 0.1 ~ "p > 0.1",
      TRUE ~ glue("p = {round(p, 3)}")
      ),
      y.position = ppos
      )

  meta %>% 
     select(!!sym(x), !!sym(y)) %>% 
     filter(!is.na(!!sym(x)), !is.na(!!sym(y))) %>% 
     ggplot(aes(x = !!sym(x), 
                y = !!sym(y),
                fill = !!sym(x))) + 
     geom_boxplot(outlier.shape = NA,
                  color = "black") +
     geom_jitter(aes(fill = !!sym(x)), 
                 color = "black",
                 shape = 21,
                 size = 3,
                 height = 0,
                 stroke = 1.2,
                 alpha = 0.6) +
     scale_fill_manual(values = study_colors$methane) +
    stat_pvalue_manual(
        test,
        label = "label",
        tip.length = 0.01,
        size = fontsize/3
      ) +
     labs(title = all_vars_mapping[[y]],
          y = "Hours",
          x = "",
          ) +
    scale_y_continuous(limits = ylim,
                       breaks = breaks,
                       expand = expansion(mult = c(0.05, 0.15))) +
    paulaR::white_theme(font_size = fontsize) +
    theme(legend.position = "none",
          plot.title = element_text(size = fontsize - 2))

}) %>% wrap_plots(ncol = length(other_vars))

other_pH_tt_boxplots
```

## Correlation breath methane and WGTT
```{r}
# this figure, incl R and P, can be found in Rmd 110
# to reproduce supplementary figure, make sure to use in 110's params$methane_group = "Methane_TT_2group"
```


# Extra numbers for paper

## pH values Wilcoxon stat tests, repeated with FDR
```{r}
ph_pval <- fig_ph_data %>% 
  group_by(segment) %>%
  summarise(
mean_HMP = mean(pH[!!sym(params$methane_group) == "HMP"], na.rm = TRUE),
    sd_HMP   = sd(pH[!!sym(params$methane_group) == "HMP"], na.rm = TRUE),
    
    mean_LMP = mean(pH[!!sym(params$methane_group) == "LMP"], na.rm = TRUE),
    sd_LMP   = sd(pH[!!sym(params$methane_group) == "LMP"], na.rm = TRUE),
  p = wilcox.test(pH ~ !!sym(params$methane_group),
                    exact = F)$p.value) %>%
  mutate(FDR = p.adjust(p, method = "BH"))  %>%
    mutate(signif_p = get_significance_asterisks(p, asterisks_p), .after = "p") %>%
    mutate(signif_FDR = get_significance_asterisks(FDR, asterisks_fdr), .after = "FDR") %>% 
  rename("pH segment" = segment)

ph_pval %>% 
  mutate(across(where(is.numeric), round, 3)) %>% 
  flextable()
```

## Transit time Wilcoxon stat tests, repeated with FDR
```{r}
tt_pval <- map(fig_tt_vars, function(y){
     
    x <- params$methane_group
    
    varname  <- all_vars_mapping[[y]]  
    
    res <- compare_means(as.formula(paste0("`", y, "` ~ `", x, "`")),
                            meta,
                            "wilcox.test",
                            exact = FALSE) %>% 
      rename("var" = `.y.`) %>% 
      select(-`p.adj`, -`p.format`, -`p.signif`) %>% 
      relocate(method, .after = "group2") %>% 
      mutate(varname = varname, .after = "var") %>% 
      mutate(
      mean_HMP = mean(meta[[y]][meta[[x]] == "HMP"], na.rm = TRUE),
      sd_HMP   = sd(meta[[y]][meta[[x]] == "HMP"], na.rm = TRUE),

      mean_LMP = mean(meta[[y]][meta[[x]] == "LMP"], na.rm = TRUE),
      sd_LMP   = sd(meta[[y]][meta[[x]] == "LMP"], na.rm = TRUE)
    )

    
    return(res)
}) %>% bind_rows() %>% 
  mutate(FDR = p.adjust(p, method = "BH")) %>% 
  mutate(signif_p = case_when(
      p < 0.001 ~ "***",
      p < 0.01  ~ "**",
      p < 0.05  ~ "*",
      TRUE            ~ ""
  )) %>% 
  mutate(signif_FDR = case_when(
    FDR < 0.001 ~ "***",
    FDR < 0.01  ~ "**",
    FDR < 0.1 ~ "*",
    TRUE ~ ""
  )) %>% 
  select("Variable" = varname,
           mean_HMP, sd_HMP,
    mean_LMP, sd_LMP,
         p, FDR, contains("signif"))

tt_pval %>% 
  mutate(across(where(is.numeric), round, 3)) %>% 
  flextable()
```

## Numbers WGTT & &CTT & ITT & prox colonic pH HMP vs LMP
```{r}
meta %>% 
  select(!!sym(params$methane_group), WGTT, SLBTT, CTT, ph_proximal_colon) %>% 
  group_by(!!sym(params$methane_group)) %>% 
  summarise(mean_WGTT = mean(WGTT, na.rm = T),
            sd_WGTT = sd(WGTT, na.rm = T),
            mean_CTT = mean(CTT, na.rm = T),
            sd_CTT = sd(CTT, na.rm = T),
            mean_ITT = mean(SLBTT, na.rm = T),
            sd_ITT = sd(SLBTT, na.rm = T),
            mean_prox_colon_pH = mean(ph_proximal_colon, na.rm = T),
            sd_prox_colon_pH = sd(ph_proximal_colon, na.rm = T)
            ) %>% 
  mutate(across(where(is.numeric), round, 1)) %>%
  flextable()
```



# Save results


## pH and TT figures, as images
```{r}
format <- "png" # png # pdf

fig_ph
ggsave(glue("{fig_dir}/{params$markdown_id}_fig_ph.{format}"),
       bg = "white",
       width = 6,
       height = 4,
       dpi = 300)

fig_SBTT
ggsave(glue("{fig_dir}/{params$markdown_id}_fig_SBTT.{format}"),
       bg = "white",
       width = 3,
       height = 3,
       dpi = 300)

fig_CTT
ggsave(glue("{fig_dir}/{params$markdown_id}_fig_CTT.{format}"),
       bg = "white",
       width = 3,
       height = 3,
       dpi = 300)

fig_SLBTT
ggsave(glue("{fig_dir}/{params$markdown_id}_fig_SLBTT.{format}"),
       bg = "white",
       width = 3,
       height = 3,
       dpi = 300)

fig_WGTT
ggsave(glue("{fig_dir}/{params$markdown_id}_fig_WGTT.{format}"),
       bg = "white",
       width = 3,
       height = 3,
       dpi = 300)

fig_pcph
ggsave(glue("{fig_dir}/{params$markdown_id}_fig_proximal_colonic_pH.{format}"),
       bg = "white",
       width = 3,
       height = 3,
       dpi = 300)
```

## pH and TT figures, as rds
```{r}
saveRDS(fig_ph, glue("{rds_dir}/{params$markdown_id}_fig_ph.rds"))

saveRDS(fig_SBTT, glue("{rds_dir}/{params$markdown_id}_fig_SBTT.rds"))

saveRDS(fig_CTT, glue("{rds_dir}/{params$markdown_id}_fig_CTT.rds"))

saveRDS(fig_SLBTT, glue("{rds_dir}/{params$markdown_id}_fig_SLBTT.rds"))

saveRDS(fig_WGTT, glue("{rds_dir}/{params$markdown_id}_fig_WGTT.rds"))
```


## Wilcox tests (pH and transit time)
```{r}
write.xlsx(ph_pval, 
            glue("{tab_dir}/{params$markdown_id}_pH_methane_groups_wilcoxon.xlsx"))


write.xlsx(tt_pval, 
            glue("{tab_dir}/{params$markdown_id}_transit_time_methane_groups_wilcoxon.xlsx"))

ph_pval %>% 
  saveRDS(glue("{rds_dir}/{params$markdown_id}_ph_methane_group_wilcoxon.rds"))
tt_pval %>% 
  saveRDS(glue("{rds_dir}/{params$markdown_id}_tt_methane_group_wilcoxon.rds"))
```


# Finalise
```{r}
sessionInfo()
```

