---
title: "Alpha and beta diversity"
author: "Paula Rodriguez-Garcia"
date: "`r Sys.time()`"
output:
  html_document:
    output_dir: "html"
params: 
  markdown_id: "141" # "141_MTT2" means methane TT 2group
  methane_group: "Methane_group"
  
  microbiome_filename: asv_rar_relabu_filename
  beta_distance: "bray"
---

# Setup

## Knit options
```{r setup}
knitr::opts_chunk$set(warning = FALSE, message = FALSE,
                       root.dir = rprojroot::find_rstudio_root_file())
```

## Libraries
```{r}
library(tibble)
library(dplyr)
library(magrittr)
library(tidyr)
library(glue)
library(purrr)
library(ggplot2)
library(stringr)
library(ggpubr)
library(openxlsx)
library(flextable)
library(vegan)
```

## Study paramters
```{r}
if (file.exists("../../study_parameters.R")) {
  source("../../study_parameters.R")
} else {
  source("study_parameters.R")
}

paulaR::show_rmd_params(params)
```


## Directories
```{r}
```

## Functions
```{r}
```

## General aesthetics
```{r}
fontsize <- 14
```

# Read data
## Meta
```{r}
meta_raw <- readRDS(meta_all_filename)
```

## Microbiome (μb) data
```{r}
ub_raw     <- readRDS(get(params$microbiome_filename))
```


# Prepare data
## Meta and microbiome data sets
```{r}
meta <- meta_raw %>% 
  filter(!is.na(!!sym(params$methane_group)))

ub   <- ub_raw %>% 
  filter(Sample_ID %in% meta$Sample_ID)
```

## Extract variables of interest for linear models
```{r}
tt_vars <- c("GET", "SBTT", "CTT", "SLBTT", "WGTT")

other_vars_of_interest <- c("fasting_CH4",
                            "smithii_inv_ct",
                            "archaea_inv_ct",
                            "stool_energy",
                            "bacterial_load_wet",
                            "ph_proximal_colon")
```

## Prepare microbiome data for beta diversity
```{r}
ub_df     <- ub %>% 
  column_to_rownames("Sample_ID") %>%
  as.data.frame()


## Sample/subject lists for PERMANOVA
sample_list <- tibble(
  Sample_ID = ub$Sample_ID
) %>% 
  mutate(Subject_ID = str_split_i(Sample_ID, "-", 1)) %>% 
  left_join(meta %>% 
              select(Subject_ID, !!sym(params$methane_group)) %>% 
              distinct(),
            by = "Subject_ID")

subject_list <- meta %>% 
  select(Subject_ID, !!sym(params$methane_group)) %>% 
  distinct()
```


# Alpha diversity
## Calculate alpha diversity
```{r}
alpha_div <- tibble(
  Sample_ID = ub$Sample_ID,
  Subject_ID = str_split_i(Sample_ID, "-", 1), 
  richness  = rowSums(ub %>% select(-Sample_ID) > 0),
  shannon   = vegan::diversity(
    ub %>% select(-Sample_ID) %>% as.data.frame(),
    "shannon")
  ) %>% 
  left_join(meta %>%
              select(Sample_ID, params$methane_group,
                     all_of(tt_vars),
                     all_of(other_vars_of_interest)),
            by = "Sample_ID") %>% 
  arrange(Sample_ID)
```

## Figure: compare alpha diversity between groups (longitudinal)

### Richness
```{r}
x <- params$methane_group
y <- "richness"
title <- "Richness"


ylim <- c(0, paulaR::ceiling_signif(max(alpha_div[[y]], na.rm = T), 1))
ppos <- ylim[1] + ((ylim[2]-ylim[1]) * 0.99)
breaks <- seq(ylim[1], ylim[2], 50)

  test <- 
        lmerTest::lmer(as.formula(paste0("`", y, "` ~ `", x, "` + (1|Subject_ID)")),
                              alpha_div) %>%
        broom.mixed::tidy() %>% 
        filter(term == paste0(params$methane_group, "LMP")) %>% 
        select(estimate, p = p.value) %>% 
        mutate(label = case_when(
                p < 0.001 ~ "p < 0.001",
                p > 0.1   ~ "p > 0.1",
                TRUE ~ glue("p = {round(p, 3)}")
                ),
               y.position = ppos,
               group1 = "LMP", 
               group2 = "HMP"
               )

fig_richness <- 
  alpha_div %>% 
     select(!!sym(x), !!sym(y)) %>% 
     filter(!is.na(!!sym(x)), !is.na(!!sym(y))) %>% 
     ggplot(aes(x = !!sym(x), 
                y = !!sym(y),
                fill = !!sym(x))) + 
     geom_boxplot(outlier.shape = NA,
                  color = "black") +
     geom_jitter(aes(fill = !!sym(x)), 
                 color = "black",
                 shape = 21,
                 size = 1,
                 stroke = 1.2,
                 height = 0,
                 alpha = 0.1) +
     scale_fill_manual(values = study_colors$methane) +
     stat_pvalue_manual(
        test,
        label = "label",
        tip.length = 0.01,
        size = fontsize/3
     ) +
     labs(title = title,
          y = "",
          x = "",
          ) +
    scale_y_continuous(limits = ylim,
                       breaks = breaks,
                       expand = expansion(mult = c(0.05, 0.15))) +
    paulaR::white_theme(font_size = fontsize) +
    theme(legend.position = "none",
          plot.title = element_text(size = fontsize - 2))

fig_richness
```

### Shannon
```{r}
x <- params$methane_group
y <- "shannon"
title <- "Shannon index"


ylim <- c(0, paulaR::ceiling_signif(max(alpha_div[[y]], na.rm = T), 0))
ppos <- ylim[1] + ((ylim[2]-ylim[1]) * 0.99)
breaks <- seq(ylim[1], ylim[2], 1)


test <- 
        lmerTest::lmer(as.formula(paste0("`", y, "` ~ `", x, "` + (1|Subject_ID)")),
                              alpha_div) %>%
        broom.mixed::tidy() %>% 
        filter(term == paste0(params$methane_group, "LMP")) %>% 
        select(estimate, p = p.value) %>% 
        mutate(label = case_when(
                p < 0.001 ~ "p < 0.001",
                p > 0.1   ~ "p > 0.1",
                TRUE ~ glue("p = {round(p, 3)}")
                ),
               y.position = ppos, 
               group1 = "LMP", 
               group2 = "HMP"
               )

fig_shannon <- 
  alpha_div %>% 
     select(!!sym(x), !!sym(y)) %>% 
     filter(!is.na(!!sym(x)), !is.na(!!sym(y))) %>% 
     ggplot(aes(x = !!sym(x), 
                y = !!sym(y),
                fill = !!sym(x))) + 
     geom_boxplot(outlier.shape = NA,
                  color = "black") +
     geom_jitter(aes(fill = !!sym(x)), 
                 color = "black",
                 shape = 21,
                 size = 1,
                 height = 0,
                 stroke = 1.2,
                 alpha = 0.2) +
     scale_fill_manual(values = study_colors$methane) +
     stat_pvalue_manual(
        test,
        label = "label",
        tip.length = 0.01,
        size = fontsize/3
     ) +
     labs(title = title,
          y = "",
          x = "",
          ) +
    scale_y_continuous(limits = ylim,
                       breaks = breaks,
                       expand = expansion(mult = c(0.05, 0.15))) +
    paulaR::white_theme(font_size = fontsize) +
    theme(legend.position = "none",
          plot.title = element_text(size = fontsize - 2))

fig_shannon
```


## Linear models (group comparison, longitudinal)

### Adjustment only by Subject_ID as random effect
```{r}
x <- params$methane_group

alpha_vars <- c("richness", "shannon")

total_all <- nrow(alpha_div)
total_hmp <- nrow(alpha_div %>% filter(!!sym(params$methane_group) == "HMP"))
total_lmp <- nrow(alpha_div %>% filter(!!sym(params$methane_group) == "LMP"))

lm_alpha_unadj <- map(alpha_vars, function(y){

  model_data <- alpha_div %>%
      dplyr::select(all_of(c(x, y, "Subject_ID"))) %>% 
      tidyr::drop_na()
  
  subject_counts <- model_data %>% group_by(!!sym(x)) %>% count()
  n_hmp <- subject_counts[[1,2]]
  n_lmp <- subject_counts[[2,2]]
  n_total <- nrow(model_data)
  
  f <- as.formula(paste0("scale(`", y, "`) ~ `", x, "` + (1|Subject_ID)"))
  model <- lmerTest::lmer(f, data = model_data)
    
    
  # extract pval
  pval <- as.numeric(summary(model)$coefficients[2, "Pr(>|t|)"])
  
  # extract log2fold change and CI
  coef <- as.numeric(summary(model)$coefficients[2, "Estimate"])
  log2FC <- coef / log(2)
  
  # extract 95% CI
  se <- as.numeric(summary(model)$coefficients[2, "Std. Error"])
  ci_log <- coef + c(-1, 1) * 1.96 * se
  ci_log2 <- ci_log / log(2)

  
  tibble::tibble(response = y,
                 predictor = x,
                 adjusted_by = "(1|Subject_ID)",
                 n_total = glue("{n_total}/{total_all}"),
                 n_HMP = glue("{n_hmp}/{total_hmp}"),
                 n_LMP = glue("{n_lmp}/{total_lmp}"),
                 beta_estimate = coef,
                 log2FC = log2FC,
                 CI_log2FC_lower = ci_log2[1],
                 CI_log2FC_upper = ci_log2[2],
                 p = pval,
                 estimate_confounder = NA,
                 p_confounder = NA
                 )

}) %>% bind_rows() %>% 
  mutate(FDR = p.adjust(p, method = "BH"), .after = "p") %>% 
    mutate(signif_p = get_significance_asterisks(p, asterisks_p), .after = "p") %>%
    mutate(signif_FDR = get_significance_asterisks(FDR, asterisks_fdr), .after = "FDR")
```

### Adjustment by transit time as fixed effect and Subject_ID as random effect
```{r}
x <- params$methane_group

combi <- expand_grid(y = alpha_vars,
                     fixed_effect = tt_vars)

lm_alpha_adj <-
  pmap(combi, function(y, fixed_effect){
  
    model_data <- alpha_div %>%
      dplyr::select(all_of(c(x, y, fixed_effect, "Subject_ID"))) %>% 
      tidyr::drop_na()
  
  subject_counts <- model_data %>% group_by(!!sym(x)) %>% count()
  n_hmp <- subject_counts[[1,2]]
  n_lmp <- subject_counts[[2,2]]
  n_total <- nrow(model_data)

  f <- as.formula(paste0("scale(`", y, "`) ~ `", x, "` + `", fixed_effect, "` + (1|Subject_ID)"))
  model <- lmerTest::lmer(f, data = model_data)
  
  # extract pval
  pval <- as.numeric(summary(model)$coefficients[2, "Pr(>|t|)"])
  
  # extract log2fold change and CI
  coef <- as.numeric(summary(model)$coefficients[2, "Estimate"])
  log2FC <- coef / log(2)
  
  # extract 95% CI
  se <- as.numeric(summary(model)$coefficients[2, "Std. Error"])
  ci_log <- coef + c(-1, 1) * 1.96 * se
  ci_log2 <- ci_log / log(2)
  
  coef_tt <- as.numeric(summary(model)$coefficients[3, "Estimate"])
  pval_tt <- as.numeric(summary(model)$coefficients[3, "Pr(>|t|)"])


  tibble::tibble(response = y,
                 predictor = x,
                 adjusted_by = glue("{fixed_effect} + (1|Subject_ID)"),
                 n_total = glue("{n_total}/{total_all}"),
                 n_HMP = glue("{n_hmp}/{total_hmp}"),
                 n_LMP = glue("{n_lmp}/{total_lmp}"),
                  beta_estimate = coef,
                 log2FC = log2FC,
                 CI_log2FC_lower = ci_log2[1],
                 CI_log2FC_upper = ci_log2[2],
                 p = pval,
                 estimate_confounder = coef_tt,
                 p_confounder = pval_tt
                 )

}) %>% bind_rows() %>% 
  group_by(adjusted_by) %>% 
  mutate(FDR = p.adjust(p, method = "BH"), .after = "p") %>% 
  ungroup() %>%
    mutate(signif_p = get_significance_asterisks(p, asterisks_p), .after = "p") %>%
    mutate(signif_FDR = get_significance_asterisks(FDR, asterisks_fdr), .after = "FDR")
```

### Combine all results
```{r}
lm_alpha <- bind_rows(lm_alpha_unadj, lm_alpha_adj) %>% 
  arrange(response)

lm_alpha %>% 
  select(-predictor) %>% 
   mutate(across(where(is.numeric), round, 3)) %>% 
  flextable()
```


## Linear models (correlation to vars of interest)
```{r}
# variables with subject-level means
vars_for_means <- c("fasting_CH4",
                    "smithii_inv_ct",
                    "archaea_inv_ct",
                    "stool_energy",
                    "ph_proximal_colon") 


combi_cor <- expand_grid(
  x = other_vars_of_interest,
  y = c("richness", "shannon"),
)


cor_alpha <- purrr::pmap(combi_cor, function(x, y){
    
  # If subject-level variable, run LM without subject adjustment
 if (x %in% vars_for_means) {

       n_total <- alpha_div$Subject_ID %>% unique() %>% length()
          
       # compute means
       model_data <- alpha_div %>%
         dplyr::select(any_of(c(x, y, "Subject_ID"))) %>% 
         group_by(Subject_ID) %>% 
         summarise(across(all_of(c(x, y)),
                          ~ mean(.x, na.rm = TRUE)))  %>%
         na.omit() 
       
       n_tested <- nrow(model_data)
          
       # note both alpha/response variable and predictor variable are scaled
       f <- as.formula(paste0("scale(`", y, "`) ~ scale(`", x, "`)"))

       random_effect <- ""
       model <- lm(f, data = model_data)
        
       pval <- as.numeric(summary(model)$coefficients[2, "Pr(>|t|)"])
    
       coef <- as.numeric(summary(model)$coefficients[2, "Estimate"])
       se      <- as.numeric(summary(model)$coefficients[2, "Std. Error"])
       ci_low  <- coef - 1.96 * se
       ci_high <- coef + 1.96 * se
        
       output <- tibble::tibble(
         response = y,
         predictor = x,
         test = "linear model",
         adjusted_by = glue(""),
         n_total = glue("{n_tested}/{n_total}"),
         estimate_type = "standardized beta coefficient",
         estimate = coef,
         CI_estimate_lower = ci_low,
         CI_estimate_upper = ci_high,
         p = pval 
        )
       
       
      # If longitudinal variable, run linear mixed model adjusting for subject
   } else {
      
      model_data <- alpha_div %>%
              dplyr::select(any_of(c(x, y, "Subject_ID"))) %>% 
              na.omit() 
      
      n_total <- nrow(alpha_div)
      n_tested <- nrow(model_data)
      
      # note both alpha/response variable and predictor variable are scaled
      f <- as.formula(paste0("scale(`", y, "`) ~ scale(`", x, "`) + (1|Subject_ID)"))
  
      random_effect <- "(1|Subject_ID)"
      model <- lmerTest::lmer(f, data = model_data)

      pval <- as.numeric(summary(model)$coefficients[2, "Pr(>|t|)"])
    
      coef <- as.numeric(summary(model)$coefficients[2, "Estimate"])
      se      <- as.numeric(summary(model)$coefficients[2, "Std. Error"])
      ci_low  <- coef - 1.96 * se
      ci_high <- coef + 1.96 * se
  
      output <- tibble::tibble(
         response = y,
         predictor = x,
         test = "linear mixed model",
         adjusted_by = glue("{random_effect}"),
         n_total = glue("{n_tested}/{n_total}"),
         estimate_type = "standardized beta coefficient",
         estimate = coef,
         CI_estimate_lower = ci_low,
         CI_estimate_upper = ci_high,
         p = pval
                     )
   }
     
     return(output)
    
    
  }) %>% bind_rows() %>% 
  group_by(adjusted_by) %>%
  mutate(FDR = p.adjust(p, method = "BH"), .after = "p") %>%
  ungroup() %>% 
  mutate(signif_p = get_significance_asterisks(p, asterisks_p), .after = "p") %>%
    mutate(signif_FDR = get_significance_asterisks(FDR, asterisks_fdr), .after = "FDR")


# see results
cor_alpha %>%
  arrange(predictor) %>%
  select(predictor, response, estimate, p, FDR) %>% 
  mutate(estimate = round(estimate,3),
         p = round(p, 4),
         FDR = round(FDR, 4)) %>% 
  flextable()
```


## Some numbers for paper
```{r}
lm_alpha %>% 
  filter(adjusted_by %in% c("(1|Subject_ID)")) %>% 
  select(response, adjusted_by, beta_estimate, p, FDR) %>% 
   mutate(across(where(is.numeric), round, 5)) %>% 
  flextable()


# might be needed
cor_alpha %>%
  arrange(predictor) %>%
  select(predictor, response, estimate, p, FDR) %>% 
  mutate(estimate = round(estimate,3),
         p = round(p, 4),
         FDR = round(FDR, 4)) %>% 
  flextable()
```



# Beta diversity
## Compute beta diversity and PCoA
```{r}
# Distances
dist_mat <- vegdist(ub_df, method = params$beta_distance)

# PCoA
pcoa <- cmdscale(dist_mat, k = nrow(ub_df)-1, eig = TRUE)

positive_axes <- which(pcoa$eig > 0)


pcoa_tib <- 
  pcoa$points[, positive_axes] %>% 
  as_tibble(rownames = "Sample_ID",
            .name_repair = "minimal") %>%
  setNames(c("Sample_ID",
             paste0("PCo", seq_len(length(positive_axes)))
             )) %>%
  mutate(Subject_ID = str_split_i(Sample_ID, "-", 1),
         .after = "Sample_ID") %>% 
  left_join(meta %>% 
              select(Sample_ID, !!sym(params$methane_group)),
            by = "Sample_ID")

# Variance explained
var_exp <- pcoa$eig / sum(pcoa$eig)
pc1_lab <- glue("PCo1 ({round(var_exp[1] * 100, 1)} %)")
pc2_lab <- glue("PCo2 ({round(var_exp[2] * 100, 1)} %)")


# Subject centroids
# using all PCos with positive eigenvalues
subject_centroids_tib <- pcoa_tib %>%
  group_by(Subject_ID, !!sym(params$methane_group)) %>%
  summarise(across(starts_with("PC"), mean)) %>% 
  ungroup()
```

## PERMANOVA

### On the distances between the centroids
```{r}
centroid_dist_mat <- 
    subject_centroids_tib %>% 
    select(Subject_ID, starts_with("PC")) %>% 
    column_to_rownames("Subject_ID") %>% 
    as.data.frame() %>% 
    dist()

# Necessary check
stopifnot(all.equal(subject_list$Subject_ID,
                    subject_centroids_tib$Subject_ID))


formula <- stringr::str_c(
  "centroid_dist_mat ~",
  params$methane_group
) %>%
  stats::as.formula()

set.seed(631)

adonis_centroids <- adonis2(
  formula,
  data = subject_list,
  permutations = 1000
)

adonis_centroids_table <- 
  tibble(
  Term = rownames(adonis_centroids),
  R2 = adonis_centroids$R2,
  F = adonis_centroids$F,
  p_value = adonis_centroids$`Pr(>F)`
)

adonis_centroids_table %>% flextable()

permanova_label <- glue(
  "R^2 == {round(adonis_centroids_table$R2[1], 2)} ~ ',' ~ P == {round(adonis_centroids_table$p_value[1], 3)}"
)
```

## Plot subject centroids (PCoA)
```{r}
fig_pcoa <- 
  ggplot() +
  # individual samples
  geom_point(
    data = pcoa_tib,
    aes(x = PCo1,
        y = PCo2,
        color = !!sym(params$methane_group)),
    alpha = 0.2,
    size = 1
  ) +
  # subject centroids
  geom_point(
    data = subject_centroids_tib,
    aes(x = PCo1, y = PCo2,
        color = !!sym(params$methane_group),
        group = !!sym(params$methane_group)),
    size = 3
  ) +
  # ellipse
  stat_ellipse(
    data = subject_centroids_tib,
    aes(x = PCo1,
        y = PCo2,
        color = !!sym(params$methane_group),
        group = !!sym(params$methane_group)
        ),
    type = "norm",
    linetype = 2
  ) +
  annotate("text", 
           x = -0.4,
           y = 0.4,
          label = permanova_label,
          hjust = 0,
          parse = T
          ) +
  scale_color_manual(values = c(study_colors$methane[[1]],
                                study_colors$methane[[2]])) +
  labs(
    x = pc1_lab,
    y = pc2_lab,
    color = "",
    title = ""
  ) +
  paulaR::white_theme(font_size = fontsize + 2) +
    theme(plot.title = element_text(size = fontsize - 4),
          legend.position = c(0.98, 0.03),
          legend.justification = c("right", "bottom"),
          legend.title = element_blank(),
          legend.background = element_rect(fill = "white")
          )

fig_pcoa
```



# Save results

## Alpha and beta div data
```{r}
saveRDS(alpha_div,
        glue("{rds_dir}/{params$markdown_id}_alpha_div.rds"))

saveRDS(dist_mat,
        glue("{rds_dir}/{params$markdown_id}_{params$beta_distance}_distance_matrix.rds"))


saveRDS(pcoa_tib,
        glue("{rds_dir}/{params$markdown_id}_{params$beta_distance}_pcoa_table.rds"))

saveRDS(subject_centroids_tib,
        glue("{rds_dir}/{params$markdown_id}_{params$beta_distance}_subject_centroids_table.rds"))
```

## Subfigures, as images
```{r}
format <- "png" # png # pdf

fig_richness
ggsave(glue("{fig_dir}/{params$markdown_id}_boxplot_richness.{format}"),
       bg = "white",
       width = 3,
       height = 3,
       dpi = 400)

fig_shannon
ggsave(glue("{fig_dir}/{params$markdown_id}_boxplot_shannon.{format}"),
       bg = "white",
       width = 3,
       height = 3,
       dpi = 400)

fig_pcoa
ggsave(glue("{fig_dir}/{params$markdown_id}_beta_div_pcoa.{format}"),
       bg = "white",
       width = 4,
       height = 4,
       dpi = 400)
```

## Subfigures, as plot rds
```{r}
saveRDS(fig_richness,
        glue("{rds_dir}/{params$markdown_id}_boxplot_richness.rds"))

saveRDS(fig_shannon,
        glue("{rds_dir}/{params$markdown_id}_boxplot_shannon.rds"))


saveRDS(fig_pcoa,
        glue("{rds_dir}/{params$markdown_id}_microbiome_pcoa.rds"))
```


## Linear model results
```{r}
write.xlsx(lm_alpha,
           glue("{tab_dir}/{params$markdown_id}_alpha_div_methane_tt_linear_models.xlsx"))

lm_alpha %>% 
  saveRDS(glue("{rds_dir}/{params$markdown_id}_alpha_div_methane_tt_linear_models.rds"))


write.xlsx(cor_alpha,
           glue("{tab_dir}/{params$markdown_id}_alpha_div_correlation_linear_models.xlsx"))

cor_alpha %>% 
  saveRDS(glue("{rds_dir}/{params$markdown_id}_alpha_div_correlation_linear_models.rds"))
```

## PERMANOVA results
```{r}
write.xlsx(adonis_centroids_table,
           glue("{tab_dir}/{params$markdown_id}_permanova_subject_centroids.xlsx"))
```


# Finalise
```{r}
sessionInfo()
```

