---
title: "Microbiome comparison, HMP vs LMP, linear models"
author: "Paula Rodriguez-Garcia"
date: "`r Sys.time()`"
output:
  html_document:
    output_dir: "html"
params: 
  markdown_id: "142_genus_counts" # can be: "142_asv_counts" or "142_genus_counts"
  methane_group: "Methane_group"
  
  microbiome_filename: genus_counts_filename # can be: "asv_counts_filename" or "genus_counts_filename"
  
  tax_level: "genus" # can be: "ASV" or "genus"
  is_ub_means: FALSE # should be: FALSE
  
  # thresholds: equal or above
  abu_filter_perc: 0.01 # 0.01 perc, this works if microbiome data rowSums() is 100
  prev_filter_perc: 0.1 # 10% of samples
  
  pseudocount: 1e-6
---

# Setup

## Knit options
```{r setup}
knitr::opts_chunk$set(warning = FALSE, message = FALSE,
                       root.dir = rprojroot::find_rstudio_root_file())
```

## Libraries
```{r}
library(tibble)
library(dplyr)
library(magrittr)
library(tidyr)
library(readxl)
library(glue)
library(purrr)
library(patchwork)
library(ggplot2)
library(stringr)
library(ggpubr)
library(openxlsx)
library(flextable)
library(phyloseq)
library(vegan)
library(effsize)
```

## Study parameters
```{r}
if (file.exists("../../study_parameters.R")) {
  source("../../study_parameters.R")
} else {
  source("study_parameters.R")
}

paulaR::show_rmd_params(params)
```


## Directories
```{r}
```

## Functions
```{r}
```


# Read data
## Meta
```{r}
if (params$is_ub_means){
  meta_raw <- readRDS(meta_means_filename)
} else {
  meta_raw <- readRDS(meta_all_filename)
}
```

## Microbiome (μb) data
```{r}
ub_raw     <- readRDS(get(params$microbiome_filename))

# for abundance filtering, one needs the rel abundances file
ub_rar_relabu_filename <- params$microbiome_filename %>%
  str_replace("counts", "rar_relabu")
ub_rar_relabu_raw <- readRDS(get(ub_rar_relabu_filename))

taxonomic_annotation <- readRDS(taxonomic_annotation_filename)
```

# Clean data
```{r}
meta <- meta_raw %>% 
  filter(!is.na(!!sym(params$methane_group)))


if (params$is_ub_means){
  ub   <- ub_raw %>% 
  filter(Subject_ID %in% meta$Subject_ID)
  
  ub_rar_relabu   <- ub_rar_relabu_raw %>% 
  filter(Subject_ID %in% meta$Subject_ID)
  
} else {
  ub   <- ub_raw %>% 
  filter(Sample_ID %in% meta$Sample_ID)
  
   ub_rar_relabu   <- ub_rar_relabu_raw %>% 
  filter(Sample_ID %in% meta$Sample_ID)
}
```


# Prepare data

## Prev and abu filtering
```{r}
prev     <- colMeans(ub[, -1] > 0)
prev_filter <- prev >= params$prev_filter_perc


# for abundance filtering, one actually needs the rel abu file
mean_abu <- colMeans(ub_rar_relabu[, -1]) 
abu_filter  <- mean_abu >= params$abu_filter_perc


filtered_taxa <- names(abu_filter & prev_filter)[abu_filter & prev_filter]
length(filtered_taxa)
```

## Prepare data for models
```{r}
# confounders
tt_vars <- c("GET", "SBTT", "CTT", "SLBTT", "WGTT")


if (params$is_ub_means){
  
  # only used for visualizations, not for the models
  ub_meta <- ub %>% 
    # filter taxa
    select(Subject_ID, all_of(filtered_taxa)) %>% 
    left_join(meta %>% 
                select(Subject_ID,
                       !!sym(params$methane_group),
                       all_of(tt_vars)
                       ),
              by = "Subject_ID") %>% 
    relocate(!!sym(params$methane_group), .after = "Subject_ID")
  
  # clr-transformed data 
  ub_clr_meta <- 
    ub %>%
  # filter taxa
  select(Subject_ID, all_of(filtered_taxa)) %>% 
  # clr transformation
  mutate(across(all_of(filtered_taxa), ~ .x + as.numeric(params$pseudocount))) %>%  
  { bind_cols(
      select(., Subject_ID),
      compositions::clr(select(., all_of(filtered_taxa)))
      )
  } %>%
    left_join(meta %>% 
                select(Subject_ID,
                       !!sym(params$methane_group),
                       all_of(tt_vars)
                       ),
              by = "Subject_ID") %>% 
    relocate(!!sym(params$methane_group), .after = "Subject_ID")
  
  
} else {
  
  # only used for visualizations, not for the models
  ub_meta <-
     ub %>%
  mutate(Subject_ID = str_split_i(Sample_ID, "-", 1), .after = "Sample_ID") %>% 
  # filter taxa
  select(Sample_ID, Subject_ID, all_of(filtered_taxa)) %>% 
  left_join(meta %>% 
              select(Sample_ID,
                     !!sym(params$methane_group),
                     all_of(tt_vars)
                     ),
            by = "Sample_ID") %>% 
  relocate(!!sym(params$methane_group),
           .after = "Subject_ID")
    
   
 # clr-transformed data
  ub_clr_meta <- 
    ub %>%
  mutate(Subject_ID = str_split_i(Sample_ID, "-", 1), .after = "Sample_ID") %>% 
  # filter taxa
  select(Sample_ID, Subject_ID, all_of(filtered_taxa)) %>% 
  # clr transformation
  mutate(across(all_of(filtered_taxa), ~ .x + as.numeric(params$pseudocount))) %>%  
  { bind_cols(
      select(., Sample_ID, Subject_ID),
      compositions::clr(select(., all_of(filtered_taxa)))
      )
  } %>%
  left_join(meta %>% 
              select(Sample_ID,
                     !!sym(params$methane_group),
                     all_of(tt_vars)
                     ),
            by = "Sample_ID") %>% 
  relocate(!!sym(params$methane_group), .after = "Subject_ID")
}
```


# Linear models

## Adjustment only by Subject_ID as random effect (adjustment only if longitudinal)
```{r}
x <- params$methane_group
total_all <- nrow(ub_meta)
total_hmp <- nrow(ub_meta %>% filter(!!sym(params$methane_group) == "HMP"))
total_lmp <- nrow(ub_meta %>% filter(!!sym(params$methane_group) == "LMP"))
  

lm_unadj_res <- map(filtered_taxa, function(y){

  # using CLR data
  model_data <- ub_clr_meta %>%
      dplyr::select(all_of(c(x, y, "Subject_ID"))) %>% 
      tidyr::drop_na()
  
  subject_counts <- model_data %>% group_by(!!sym(x)) %>% count()
  n_hmp <- subject_counts[[1,2]]
  n_lmp <- subject_counts[[2,2]]
  n_total <- nrow(model_data)
  
  if (params$is_ub_means){
    
    f <- as.formula(paste0("`", y, "` ~ `", x, "`"))
    random_effect <- ""
    model <- lm(f, data = model_data)
  
  } else {
    f <- as.formula(paste0("`", y, "` ~ `", x, "` + (1|Subject_ID)"))
    random_effect <- "(1|Subject_ID)"
    model <- lmerTest::lmer(f, data = model_data)
  }
    
  # extract pval
  pval <- as.numeric(summary(model)$coefficients[2, "Pr(>|t|)"])
  
  # extract log2fold change and CI
  coef <- as.numeric(summary(model)$coefficients[2, "Estimate"])
  log2FC <- coef / log(2)
  
  # extract 95% CI
  se <- as.numeric(summary(model)$coefficients[2, "Std. Error"])
  ci_log <- coef + c(-1, 1) * 1.96 * se
  ci_log2 <- ci_log / log(2)
  
  sd_y <- sd(model_data[[y]], na.rm = TRUE)
  coef_std <- coef / sd_y

  tibble::tibble(response = y,
                 predictor = x,
                 adjusted_by = random_effect,
                 n_total = glue("{n_total}/{total_all}"),
                 n_HMP = glue("{n_hmp}/{total_hmp}"),
                 n_LMP = glue("{n_lmp}/{total_lmp}"),
                 estimate = coef,
                 estimate_std = coef_std,
                 log2FC = log2FC,
                 CI_log2FC_lower = ci_log2[1],
                 CI_log2FC_upper = ci_log2[2],
                 p = pval,
                 beta_confounder = NA,
                 p_confounder = NA
                 )

}) %>% bind_rows() %>% 
  mutate(FDR = p.adjust(p, method = "BH"), .after = "p") %>%
    mutate(signif_p = get_significance_asterisks(p, asterisks_p), .after = "p") %>%
    mutate(signif_FDR = get_significance_asterisks(FDR, asterisks_fdr), .after = "FDR")

if (params$tax_level == "ASV"){
    lm_unadj_res %<>% 
    left_join(taxonomic_annotation %>% 
                  select(
                    !!sym(params$tax_level),
                    closest_tax),
                by = c("response" = params$tax_level)) %>% 
      relocate(c(closest_tax), .after = response)
  
  }
```

## Adjustment by transit time as fixed effect and Subject_ID as random effect (Subject ID adjustment only if longitudinal)
```{r}
x <- params$methane_group
combi <- expand_grid(y = filtered_taxa,
                     fixed_effect = tt_vars)
total_all <- nrow(ub_meta)
total_hmp <- nrow(ub_meta %>% filter(!!sym(params$methane_group) == "HMP"))
total_lmp <- nrow(ub_meta %>% filter(!!sym(params$methane_group) == "LMP"))


lm_adj_res <-
  pmap(combi, function(y, fixed_effect){
  
  # using CLR data
  model_data <- ub_clr_meta %>%
      dplyr::select(all_of(c(x, y, fixed_effect, "Subject_ID"))) %>% 
      tidyr::drop_na()

  
    subject_counts <- model_data %>% group_by(!!sym(x)) %>% count()
    n_hmp <- subject_counts[[1,2]]
    n_lmp <- subject_counts[[2,2]]
    n_total <- nrow(model_data)
  
    if (params$is_ub_means){
    
      random_effect <- ""
      f <- as.formula(paste0("`", y, "` ~ `", x, "` + `", fixed_effect, "`"))
      model <- lm(f, data = model_data)

  } else {
      random_effect <- " + (1|Subject_ID)"
      f <- as.formula(paste0("`", y, "` ~ `", x, "` + `", fixed_effect, "` + (1|Subject_ID)"))
      
      model <- lmerTest::lmer(f, data = model_data)
  }
    
  # extract pval
  pval <- as.numeric(summary(model)$coefficients[2, "Pr(>|t|)"])
  
  # extract log2fold change and CI
  coef <- as.numeric(summary(model)$coefficients[2, "Estimate"])
  log2FC <- coef / log(2)
  
  # extract 95% CI
  se <- as.numeric(summary(model)$coefficients[2, "Std. Error"])
  ci_log <- coef + c(-1, 1) * 1.96 * se
  ci_log2 <- ci_log / log(2)
  
    # extract std beta (in CLR units if model data not QMP)
  sd_y <- sd(model_data[[y]], na.rm = TRUE)
  coef_std <- coef / sd_y
    

  coef_tt <- as.numeric(summary(model)$coefficients[3, "Estimate"])
  pval_tt <- as.numeric(summary(model)$coefficients[3, "Pr(>|t|)"])


  tibble::tibble(response = y,
                 predictor = x,
                 adjusted_by = glue("{fixed_effect}{random_effect}"),
                 n_total = glue("{n_total}/{total_all}"),
                 n_HMP = glue("{n_hmp}/{total_hmp}"),
                 n_LMP = glue("{n_lmp}/{total_lmp}"),
                 estimate = coef,
                 estimate_std = coef_std,
                 log2FC = log2FC,
                 CI_log2FC_lower = ci_log2[1],
                 CI_log2FC_upper = ci_log2[2],
                 p = pval,
                 beta_confounder = coef_tt,
                 p_confounder = pval_tt
                 )

}) %>% bind_rows() %>% 
  group_by(adjusted_by) %>% 
  mutate(FDR = p.adjust(p, method = "BH"), .after = "p") %>% 
  ungroup() %>%
    mutate(signif_p = get_significance_asterisks(p, asterisks_p), .after = "p") %>%
    mutate(signif_FDR = get_significance_asterisks(FDR, asterisks_fdr), .after = "FDR")


if (params$tax_level == "ASV"){
    lm_adj_res %<>% 
    left_join(taxonomic_annotation %>% 
                  select(
                    !!sym(params$tax_level),
                    closest_tax),
                by = c("response" = params$tax_level)) %>% 
      relocate(c(closest_tax), .after = response)
  
  }
```

## Combine all results
```{r}
lm_res <- bind_rows(lm_unadj_res, lm_adj_res) %>% 
  arrange(response) %>% 
  mutate(signif_p_confounder = get_significance_asterisks(p_confounder, asterisks_p))
```



# Extra plots
## Ind bacterium distribution
```{r}
if (params$tax_level == "genus"){
  
  x <- "Methane_group" #params$methane_group
  fontsize <- 14
  
  vars <- c("Agathobacter",
            "Anaerobutyricum",
            "Anaerostipes",
            "Bifidobacterium",
            "Beduinibacterium", 
            "Blautia",
            "Butyrivibrio",
            "Clostridium_IV",
            "Eubacterium",
            "Faecalibacterium",
            "Intestinimonas",
            "Lactobacillus", 
            "Massilimicrobiota",
            "Monoglobus",
            "Oscillibacter",
            "Pseudoflavonifractor",
            "Streptococcus"
            )
  
bact_boxplots <-  map(vars, function(y){
    ylim <- c(0, paulaR::ceiling_signif(max(ub_meta[[y]], na.rm = T), 0))
    
    if (x %in% colnames(ub_meta)){
      model_data <- ub_meta %>% 
           select(!!sym(x), !!sym(y)) %>% 
           na.omit()
    } else {
      
      if (params$is_ub_means){
          model_data <- ub_meta %>% 
            left_join(meta %>% select(!!sym(x), Subject_ID),
                      by = "Subject_ID") %>% 
           select(!!sym(x), !!sym(y)) %>% 
           na.omit()
      } else {
          model_data <- ub_meta %>% 
            left_join(meta %>% select(!!sym(x), Sample_ID),
                      by = "Sample_ID") %>% 
           select(!!sym(x), !!sym(y)) %>% 
           na.omit()
      }
    }
    
    subject_counts <- model_data %>% 
      group_by(group = !!sym(x)) %>% 
      count()
    
    labels <- setNames(
      paste0(subject_counts$group, "\n(N=", subject_counts$n, ")"),
      subject_counts$group
    )
    
    
    test <- compare_means(as.formula(paste(y, "~", x)),
                          model_data,
                          "wilcox.test",
                          exact = FALSE) %>%
      mutate(label = ifelse(
        p < 0.001,
        "p < 0.001",
        glue("p = {round(p, 3)}")
        ),
        y.position = max(model_data[[y]], na.rm = TRUE) * 1.05
        )
    
    bact_groups_boxplot <- 
      model_data %>% 
         select(!!sym(x), !!sym(y)) %>% 
         filter(!is.na(!!sym(x)), !is.na(!!sym(y))) %>% 
         ggplot(aes(x = !!sym(x), 
                    y = !!sym(y),
                    fill = !!sym(x))) + 
         geom_boxplot(outlier.shape = NA,
                      color = "black") +
         geom_jitter(aes(fill = !!sym(x)), 
                     color = "black",
                     shape = 21,
                     size = 1,
                     height = 0,
                     stroke = 1.2,
                     alpha = 0.6) +
         scale_fill_manual(values = study_colors[[x]]) +
        stat_pvalue_manual(
            test,
            label = "label",
            tip.length = 0.01,
            size = fontsize/5
          ) +
      scale_x_discrete(labels = labels) +
         labs(title = "",
              y = glue("{y} (%)"),
              x = "",
              ) +
        coord_cartesian(ylim = ylim) +
        scale_y_continuous(expand = expansion(mult = c(0.05, 0.1))) +
        paulaR::white_theme(font_size = fontsize) +
        theme(legend.position = "none")
    
    bact_groups_boxplot
  }) %>% wrap_plots() 

bact_boxplots 

}


```

## Christensenella/ceae
```{r}
if (!params$is_ub_means & params$tax_level == "genus" & params$methane_group == "Methane_group"){

filtered_taxa %>% str_extract("Christ") # it must have been filtered out after abundance or prevalence filters

# do we really have these taxa in the overall file?
taxonomic_annotation %>% filter(Family == "Christensenellaceae")  # yes, 8 ASV

christ_asv <- taxonomic_annotation %>% filter(Family == "Christensenellaceae") %>% arrange(closest_tax) %>% pull(ASV) %>% unique()
christ_gen <- taxonomic_annotation %>% filter(Family == "Christensenellaceae") %>% arrange(closest_tax) %>%  pull(genus_or_closest) %>% unique()
christ_colnames <- c("Christensenellaceae", christ_gen, christ_asv)


# it seems only beduinibacterium passed the filter
prev[["Christensenella"]] # 29%
prev[["Beduinibacterium"]] # 24%
prev[["Unclassified Christensenellaceae"]] # 61%

mean_abu[["Christensenella"]] #  0.002%
mean_abu[["Beduinibacterium"]] # 0.010%
mean_abu[["Unclassified Christensenellaceae"]] # 0.008%


# get the data
asv_counts  <- readRDS(asv_counts_filename)
genus_counts  <- readRDS(genus_counts_filename)
family_counts  <- readRDS(family_counts_filename)


asv_rar_relabu  <- readRDS(asv_rar_relabu_filename)
genus_rar_relabu  <- readRDS(genus_rar_relabu_filename)
family_rar_relabu  <- readRDS(family_rar_relabu_filename)


# clr transformation
asv_clr <- asv_counts %>%
  mutate(across(-c(Sample_ID), ~ .x + as.numeric(params$pseudocount))) %>%  
  { bind_cols(
      select(., Sample_ID),
      compositions::clr(select(., -c(Sample_ID))))
      
  } %>%
    left_join(meta %>% 
                select(Sample_ID,
                       !!sym(params$methane_group)
                       ),
              by = "Sample_ID") %>% 
    relocate(!!sym(params$methane_group), .after = "Sample_ID")

genus_clr <- genus_counts %>%
  mutate(across(-c(Sample_ID), ~ .x + as.numeric(params$pseudocount))) %>%  
  { bind_cols(
      select(., Sample_ID),
      compositions::clr(select(., -c(Sample_ID))))
      
  } %>%
    left_join(meta %>% 
                select(Sample_ID,
                       !!sym(params$methane_group)
                       ),
              by = "Sample_ID") %>% 
    relocate(!!sym(params$methane_group), .after = "Sample_ID")

family_clr <- family_counts %>%
  mutate(across(-c(Sample_ID), ~ .x + as.numeric(params$pseudocount))) %>%  
  { bind_cols(
      select(., Sample_ID),
      compositions::clr(select(., -c(Sample_ID))))
      
  } %>%
    left_join(meta %>% 
                select(Sample_ID,
                       !!sym(params$methane_group)
                       ),
              by = "Sample_ID") %>% 
    relocate(!!sym(params$methane_group), .after = "Sample_ID")


christ_counts <- asv_counts %>% 
  select(Sample_ID, all_of(christ_asv)) %>% 
  left_join(genus_counts %>% select(Sample_ID, all_of(christ_gen)),
            by = "Sample_ID") %>% 
  left_join(family_counts %>% select(Sample_ID, "Christensenellaceae"),
            by = "Sample_ID") %>% 
  left_join(meta %>% select(Sample_ID, !!sym(params$methane_group)),
            by = "Sample_ID") %>% 
  mutate(Subject_ID = str_split_i(Sample_ID, "-", 1), .after = "Sample_ID") %>% 
  relocate(!!sym(params$methane_group), .after = "Subject_ID")


christ_rar_relabu <- asv_rar_relabu %>% 
  select(Sample_ID, all_of(christ_asv)) %>% 
  left_join(genus_rar_relabu %>% select(Sample_ID, all_of(christ_gen)),
            by = "Sample_ID") %>% 
  left_join(family_rar_relabu %>% select(Sample_ID, "Christensenellaceae"),
            by = "Sample_ID") %>% 
  left_join(meta %>% select(Sample_ID, !!sym(params$methane_group)),
            by = "Sample_ID") %>% 
  mutate(Subject_ID = str_split_i(Sample_ID, "-", 1), .after = "Sample_ID") %>% 
  relocate(!!sym(params$methane_group), .after = "Subject_ID")


christ_clr <- asv_clr %>% 
  select(Sample_ID, !!sym(params$methane_group), all_of(christ_asv)) %>% 
  left_join(genus_clr %>% select(Sample_ID, all_of(christ_gen)),
            by = "Sample_ID") %>% 
  left_join(family_clr %>% select(Sample_ID, "Christensenellaceae"),
            by = "Sample_ID") %>% 
  mutate(Subject_ID = str_split_i(Sample_ID, "-", 1), .after = "Sample_ID") 


x <- params$methane_group
fontsize <- 14

christ_boxplots <- map(christ_colnames[1:4], function(y){
  
  model_data <- christ_clr %>% 
    select(!!sym(x), !!sym(y), Subject_ID)
  
  plot_data <- christ_rar_relabu %>% 
     select(!!sym(x), !!sym(y), Subject_ID)
  
  
  f <- as.formula(paste0("`", y, "` ~  + `", x, "` + (1|Subject_ID)"))
  
  model <- lmerTest::lmer(f, data = model_data)
  
  pval <- as.numeric(summary(model)$coefficients[2, "Pr(>|t|)"])
  coef <- as.numeric(summary(model)$coefficients[2, "Estimate"])
  print(y)
  print(coef)
  
  pval_pos <- 15
  
  plot_p_aes <- tibble(
    .y. = y,
    group1 = "HMP",
    group2 = "LMP",
    p = pval
  ) %>%  mutate(label = ifelse(
        p < 0.001,
        "p < 0.001",
        glue("p = {round(p, 3)}")
        ),
        y.position = pval_pos
        )
  
  ylim <- c(0, paulaR::ceiling_signif(max(plot_data[[y]], na.rm = T), -1))
  
  title <- ifelse(str_detect(y, "ASV"),
                  glue("{y} ({taxonomic_annotation %>% filter(ASV == y) %>% pull(closest_tax)})"),
                  y)
  
  plot_data %>% 
         select(!!sym(x), !!sym(y)) %>% 
         filter(!is.na(!!sym(x)), !is.na(!!sym(y))) %>% 
         ggplot(aes(x = !!sym(x), 
                    y = !!sym(y),
                    fill = !!sym(x))) + 
         geom_boxplot(outlier.shape = NA,
                      color = "black") +
         geom_jitter(aes(fill = !!sym(x)), 
                     color = "black",
                     shape = 21,
                     size = 1,
                     height = 0,
                     stroke = 1.2,
                     alpha = 0.6) +
         scale_fill_manual(values = study_colors[[x]]) +
        stat_pvalue_manual(
            plot_p_aes,
            label = "label",
            tip.length = 0.01,
            size = fontsize - 10
          ) +
      scale_x_discrete() +
         labs(title = title,
              y = glue("Rel. abundance (%)"),
              x = "",
              ) +
        scale_y_continuous(
        trans = scales::pseudo_log_trans(
        base = 2, sigma = 0.0001),
        limits = c(-0.00025, pval_pos),
       breaks = c(0, 0.001, 0.01, 0.1, 1),
        ) +
        paulaR::white_theme(font_size = fontsize) +
        theme(legend.position = "none",
              title = element_text(size = fontsize-4),
              axis.title.y = element_text(size = fontsize))
  
  
}) %>% wrap_plots(ncol = 2) + plot_annotation(tag_levels = "A")


christ_boxplots


}
```


# Extra numbers for the paper
```{r}
if (params$tax_level == "genus"){
  # butyrivibrio
  lm_res %>% 
    filter(response == "Butyrivibrio") %>% 
    select(response, adjusted_by, p, FDR) %>% 
    mutate(p = round(p, 4),
           FDR = round(FDR, 4)) %>% 
    flextable()
}
```




# Save results


## Linear models
```{r}
write.xlsx(lm_res,
             glue("{tab_dir}/{params$markdown_id}_ub_methane_group_tt_linear_models.xlsx"))
saveRDS(lm_res,
             glue("{rds_dir}/{params$markdown_id}_ub_methane_group_tt_linear_models.rds"))
```

## Individual bacterial boxplots
```{r}
if (params$tax_level == "genus" & !str_detect(params$markdown_id, "QMP")){
  format <- "png" # png # pdf
  
  bact_boxplots + plot_layout(ncol = 3)
  ggsave(glue("{fig_dir}/{params$markdown_id}_boxplots_ind_genera_vertical.{format}"),
         bg = "white",
         width = 12,
         height = 20,
         dpi = 300)
  
  bact_boxplots + plot_layout(ncol = 5)
  ggsave(glue("{fig_dir}/{params$markdown_id}_boxplots_ind_genera_horizontal.{format}"),
         bg = "white",
         width = 20,
         height = 12,
         dpi = 300)
  
  christ_boxplots
  ggsave(glue("{fig_dir}/{params$markdown_id}_Christensenellaceae_family_genera_boxplots.{format}"),
         bg = "white",
         width = 8,
         height = 7,
         dpi = 300)
}
```




# Finalise
```{r}
sessionInfo()
```

