---
title: "Microbiome correlations to continuous variables, linear models"
author: "Paula Rodriguez-Garcia"
date: "`r Sys.time()`"
output:
  html_document:
    output_dir: "html"
params: 
  markdown_id: "143_genus_counts_mixed" # can be: "143_asv_counts_mixed" or "143_genus_counts_mixed" 
  methane_group: "Methane_group"
  
  microbiome_filename: genus_counts_filename # can be: "asv_counts_filename", or "genus_counts_filename"
  
  tax_level: "genus" # can be: "ASV" or "genus"
  is_ub_means: FALSE # should be: FALSE
  mixed: TRUE # should be: TRUE
      # mixed means that we are doing the correlations differently
      # if the variable has 2 or less repeated measures, we do LM on means data, not adjusting for Subject ID
      # if the variable has 3 or more repeated measured, we do LMM adjusting for Subject ID
      # this is also done like this in markdown 121 and 171
  
  # thresholds: equal or above
  abu_filter_perc: 0.01 # 0.01 perc, this works if microbiome data rowSums() is 100
  prev_filter_perc: 0.1 # 10% of samples
  
  cor_method: "spearman"
  pseudocount: 1e-6
---


# Setup

## Knit options
```{r setup}
knitr::opts_chunk$set(warning = FALSE, message = FALSE,
                       root.dir = rprojroot::find_rstudio_root_file())
```

## Libraries
```{r}
library(tibble)
library(dplyr)
library(magrittr)
library(tidyr)
library(glue)
library(purrr)
library(ggplot2)
library(stringr)
library(openxlsx)
library(flextable)
```

## Study paramters
```{r}
if (file.exists("../../study_parameters.R")) {
  source("../../study_parameters.R")
} else {
  source("study_parameters.R")
}

paulaR::show_rmd_params(params)
```


## Directories
```{r}
```

## Functions
```{r}
```

## General aesthetics
```{r}
fontsize <- 14
```

# Read data
## Meta
```{r}
if (params$is_ub_means){
  meta_raw <- readRDS(meta_means_filename)
} else {
  meta_raw <- readRDS(meta_all_filename)
}
```

## Microbiome (μb) data
```{r}
ub_raw     <- readRDS(get(params$microbiome_filename))


# for abundance filtering, one needs the rel abundances file
ub_rar_relabu_filename <- params$microbiome_filename %>% str_replace("counts", "rar_relabu")
ub_rar_relabu_raw <- readRDS(get(ub_rar_relabu_filename))


taxonomic_annotation <- readRDS(taxonomic_annotation_filename)
```

# Clean data
```{r}
# meta & ub: nothing needed for now
meta <- meta_raw %>% 
  filter(!is.na(!!sym(params$methane_group)))


if (params$is_ub_means){
  ub   <- ub_raw %>% 
  filter(Subject_ID %in% meta$Subject_ID)
  
   ub_rar_relabu   <- ub_rar_relabu_raw %>% 
  filter(Subject_ID %in% meta$Subject_ID)
   
   
} else {
  ub   <- ub_raw %>% 
  filter(Sample_ID %in% meta$Sample_ID)
  
  ub_rar_relabu   <- ub_rar_relabu_raw %>% 
  filter(Sample_ID %in% meta$Sample_ID)
}


```


# Prepare data

## Prev and abu filtering
```{r}
prev     <- colMeans(ub[, -1] > 0)
prev_filter <- prev >= params$prev_filter_perc


# for abundance filtering, one actually needs the rel abu file
mean_abu <- colMeans(ub_rar_relabu[, -1]) 
abu_filter  <- mean_abu >= params$abu_filter_perc

filtered_taxa <- names(abu_filter & prev_filter)[abu_filter & prev_filter]
```

## Prepare data for models
```{r}
vars_of_interest <- c("fasting_CH4",
                      "fasting_H2",
                      "smithii_inv_ct",
                      "archaea_inv_ct",
                      "ph_proximal_colon",
                      "stool_energy", 
                      "bacterial_load_wet"
                      )
# confounders
tt_vars <- c("GET", "SBTT", "CTT", "SLBTT", "WGTT")

# variables where we actually need to use means (if mixed)
vars_for_means <- c("stool_energy",
                    "cn_ratio",
                    "ph_small_bowel",
                    "ph_colon",
                    "ph_proximal_colon",
                    "ph_distal_colon",
                    "ph_sigmoid_colon",
                    "ph_rectum",
                    "fasting_CH4",
                    "fasting_H2",
                    "smithii_inv_ct",
                    "archaea_inv_ct",
                    "glp1")


if (params$is_ub_means){
  
  # not used for the models
  ub_meta <- ub %>% 
    # filter taxa
    select(Subject_ID, all_of(filtered_taxa)) %>% 
    left_join(meta %>% 
                select(Subject_ID,
                       !!sym(params$methane_group),
                       all_of(tt_vars),
                       all_of(vars_of_interest)
                       ),
              by = "Subject_ID") %>% 
    relocate(!!sym(params$methane_group), .after = "Subject_ID")
  
  # clr-transformed data
  ub_clr_meta <- 
    ub %>%
  # filter taxa
  select(Subject_ID, all_of(filtered_taxa)) %>% 
  # clr transformation
  mutate(across(all_of(filtered_taxa), ~ .x + as.numeric(params$pseudocount))) %>%  
  { bind_cols(
      select(., Subject_ID),
      compositions::clr(select(., all_of(filtered_taxa)))
      )
  } %>%
    left_join(meta %>% 
                select(Subject_ID,
                       !!sym(params$methane_group),
                       all_of(tt_vars),
                       all_of(vars_of_interest)
                       ),
              by = "Subject_ID") %>% 
    relocate(!!sym(params$methane_group), .after = "Subject_ID")
  
  
} else {
  # not used for the models
  ub_meta <- ub %>%
  mutate(Subject_ID = str_split_i(Sample_ID, "-", 1), .after = "Sample_ID") %>% 
  # filter taxa
  select(Sample_ID, Subject_ID, all_of(filtered_taxa)) %>% 
  left_join(meta %>% 
              select(Sample_ID,
                     !!sym(params$methane_group),
                     all_of(tt_vars),
                       all_of(vars_of_interest)
                     ),
            by = "Sample_ID") %>% 
  relocate(!!sym(params$methane_group), .after = "Subject_ID")
  
  
   # clr-transformed data
  ub_clr_meta <- ub %>%
  mutate(Subject_ID = str_split_i(Sample_ID, "-", 1), .after = "Sample_ID") %>% 
  # filter taxa
  select(Sample_ID, Subject_ID, all_of(filtered_taxa)) %>% 
  # clr transformation
  mutate(across(all_of(filtered_taxa), ~ .x + as.numeric(params$pseudocount))) %>%  
  { bind_cols(
      select(., Sample_ID, Subject_ID),
      compositions::clr(select(., all_of(filtered_taxa)))
      )
  } %>%
  left_join(meta %>% 
              select(Sample_ID,
                     !!sym(params$methane_group),
                     all_of(tt_vars),
                     all_of(vars_of_interest)
                     ),
            by = "Sample_ID") %>% 
  relocate(!!sym(params$methane_group), .after = "Subject_ID")
   
  
  if (params$mixed){
    
    # then compute mean counts
    ub_meta_means <-  ub_meta %>% 
      select(-Sample_ID, -!!sym(params$methane_group)) %>% 
      group_by(Subject_ID) %>% 
       summarise(across(everything(), ~ mean(.x, na.rm = TRUE))) %>% 
      left_join(ub_meta %>% 
                  select(Subject_ID, !!sym(params$methane_group)) %>%
                  distinct(),
                by = "Subject_ID") %>% 
      relocate(!!sym(params$methane_group), .after = "Subject_ID")
    
    # CLR-transform the mean counts
   ub_clr_meta_means <-
      ub_meta_means %>%
        # clr transformation
        mutate(across(all_of(filtered_taxa), ~ .x + as.numeric(params$pseudocount))) %>%  
  { bind_cols(
      select(., Subject_ID),
      compositions::clr(select(., all_of(filtered_taxa)))
      )
  } %>%
  left_join(ub_meta_means %>% 
              select(Subject_ID,
                     !!sym(params$methane_group),
                     all_of(tt_vars),
                     all_of(vars_of_interest)
                     ),
            by = "Subject_ID") %>% 
  relocate(!!sym(params$methane_group), .after = "Subject_ID")
  
  
}}
```



# Linear models

## Adjustment only by Subject_ID as random effect (adjustment only if longitudinal)
```{r}
n_total <- nrow(ub_meta)

combi <- expand_grid(y = filtered_taxa,
                       x = vars_of_interest)

lm_unadj_res <- pmap(combi, function(y, x){

  if (params$mixed){
      
      model_data <- ub_clr_meta_means %>% 
        select(all_of(c(x, y, "Subject_ID"))) %>% 
      tidyr::drop_na()
      
    } else {
  model_data <- ub_clr_meta %>%
      dplyr::select(all_of(c(x, y, "Subject_ID"))) %>% 
      tidyr::drop_na()
  }
  
  n_tested <- nrow(model_data)
  
  if (params$is_ub_means | params$mixed){
    
    f <- as.formula(paste0("`", y, "` ~ scale(`", x, "`)"))
    random_effect <- ""
    model <- lm(f, data = model_data)
  
  } else {

    f <- as.formula(paste0("`", y, "` ~ scale(`", x, "`) + (1|Subject_ID)"))
    random_effect <- "(1|Subject_ID)"
    model <- lmerTest::lmer(f, data = model_data)

  }
    
  # extract pvalue
  pval <- as.numeric(summary(model)$coefficients[2, "Pr(>|t|)"])

    
  # extract unstandardized coefficient
  coef    <- as.numeric(summary(model)$coefficients[2, "Estimate"])
  se      <- as.numeric(summary(model)$coefficients[2, "Std. Error"])
  ci_low  <- coef - 1.96 * se
  ci_high <- coef + 1.96 * se
  
  
  # extract std beta (in CLR units if model data not QMP)
  sd_y <- sd(model_data[[y]], na.rm = TRUE)
  coef_std <- coef / sd_y
  
  tibble::tibble(response = y,
                 predictor = x,
                 adjusted_by = random_effect,
                 n_total = glue("{n_tested}/{n_total}"),
                 estimate_type = "beta coefficient",
                 estimate = coef,
                 estimate_std = coef_std,
                 CI_estimate_lower = ci_low,
                 CI_estimate_upper = ci_high,
                 p = pval,
                 estimate_confounder = NA,
                 p_confounder = NA
                 )

}) %>% bind_rows() %>% 
  group_by(predictor) %>% 
  mutate(FDR = p.adjust(p, method = "BH"), .after = "p") %>%
  ungroup() %>% 
    mutate(signif_p = get_significance_asterisks(p, asterisks_p), .after = "p") %>%
    mutate(signif_FDR = get_significance_asterisks(FDR, asterisks_fdr), .after = "FDR")

if (params$tax_level == "ASV"){
    lm_unadj_res %<>% 
    left_join(taxonomic_annotation %>% 
                  select(
                    !!sym(params$tax_level),
                    closest_tax),
                by = c("response" = params$tax_level)) %>% 
      relocate(c(closest_tax), .after = response)
  
  }
```

## Adjustment by transit time as fixed effect and Subject_ID as random effect (Subject ID adjustment only if longitudinal)
```{r}
n_total <- nrow(ub_meta)

combi <- expand_grid(y = filtered_taxa,
                     x = vars_of_interest,
                     fixed_effect = tt_vars)


lm_adj_res <-
  pmap(combi, function(y, x, fixed_effect){

    if (params$mixed){
      
      model_data <- ub_clr_meta_means %>% 
        select(all_of(c(x, y, fixed_effect, "Subject_ID"))) %>% 
      tidyr::drop_na()
      
    } else {
  model_data <- ub_clr_meta %>%
      dplyr::select(all_of(c(x, y, fixed_effect, "Subject_ID"))) %>% 
      tidyr::drop_na()
  }
  
    n_tested <- nrow(model_data)
  
    random_effect <- if (params$is_ub_means | params$mixed) "" else " + (1|Subject_ID)"

    f <- as.formula(
      paste0("`", y, "` ~ scale(`", x, "`) + scale(`", fixed_effect, "`)", random_effect)
    )

   output <- tryCatch({

      model <- if (params$is_ub_means  | params$mixed) {
        lm(f, data = model_data)
      } else {
        lmerTest::lmer(f, data = model_data)
      }

      sm <- summary(model)$coefficients
      
      # extract pvalue
      pval <- as.numeric(sm[2, "Pr(>|t|)"])
    
        
      # extract unstandardized coefficient
      coef    <- as.numeric(sm[2, "Estimate"])
      se      <- as.numeric(sm[2, "Std. Error"])
      ci_low  <- coef - 1.96 * se
      ci_high <- coef + 1.96 * se
  
    # extract std beta (in CLR units if model data not QMP)
  sd_y <- sd(model_data[[y]], na.rm = TRUE)
  coef_std <- coef / sd_y
      

      tibble::tibble(
        response = y,
        predictor = x,
        adjusted_by = glue::glue("{fixed_effect}{random_effect}"),
        n_total = glue::glue("{n_tested}/{n_total}"),
       estimate_type = "beta coefficient",
       estimate = coef,
       estimate_std = coef_std,
       CI_estimate_lower = ci_low,
       CI_estimate_upper = ci_high,
        p = pval,
        estimate_confounder = as.numeric(sm[3, "Estimate"]),
        p_confounder = as.numeric(sm[3, "Pr(>|t|)"])
      )

    }, error = function(e) {

      tibble::tibble(
        response = y,
        predictor = x,
        adjusted_by = glue::glue("{fixed_effect}{random_effect}"),
        n_total = glue::glue("{n_tested}/{n_total}"),
        estimate_type = "beta coefficient",
        estimate = NA_real_,
        estimate_std = NA_real_,
        CI_estimate_lower = NA_real_,
        CI_estimate_upper = NA_real_,
        p = NA_real_,
        estimate_confounder = NA_real_,
        p_confounder = NA_real_
      )
    })
        
        return(output)

}) %>% bind_rows() %>% 
  group_by(predictor, adjusted_by) %>% 
  mutate(FDR = p.adjust(p, method = "BH"), .after = "p") %>% 
  ungroup() %>%
    mutate(signif_p = get_significance_asterisks(p, asterisks_p), .after = "p") %>%
    mutate(signif_FDR = get_significance_asterisks(FDR, asterisks_fdr), .after = "FDR")


if (params$tax_level == "ASV"){
    lm_adj_res %<>% 
    left_join(taxonomic_annotation %>% 
                  select(
                    !!sym(params$tax_level),
                    closest_tax),
                by = c("response" = params$tax_level)) %>% 
      relocate(c(closest_tax), .after = response)
  
  }
```

## Combine all results
```{r}
lm_res <- bind_rows(lm_unadj_res, lm_adj_res) %>% 
  arrange(response, predictor) %>% 
  # final fdr
    group_by(adjusted_by) %>%
  mutate(FDR = p.adjust(p, method = "BH"), .after = "p") %>% 
  ungroup() %>%
    mutate(signif_FDR = get_significance_asterisks(FDR, asterisks_fdr), .after = "FDR")
```

# Extra numbers for paper
```{r}
if (params$tax_level == "genus"){
  # smithii and beduinibacterium
  lm_unadj_res %>% 
    filter(str_detect(predictor, "smithii"), response == "Beduinibacterium") %>% 
    select(response, predictor, estimate_std, p, FDR) %>% 
    mutate(estimate_std = round(estimate_std, 3),
           p = round(p, 4), 
           FDR = round(FDR, 4)) %>% 
    flextable()
}
```



# Individual figures (potential suplementary)
##  Beduinibacterium and smithii
```{r}
# only if genus
if (params$tax_level == "genus"){
  
  x <- "smithii_inv_ct"
  y <- "Beduinibacterium"
  
  if (params$is_ub_means){
    
    res <- np_cor_res %>% 
      filter(predictor == x,
             response == y)
    
    caption <- glue("{str_to_title(params$cor_method)} correlation.")
    
    r_or_beta <- "R"
    
  } else {
    
    res <- lm_res %>% 
      filter(predictor == x,
             response == y,
             adjusted_by == ""
             )

    r_or_beta <- "Standardized β"

  }
  
  plot_data <- ub_rar_relabu %>% 
    select(Sample_ID, !!sym(y)) %>% 
    left_join(meta %>% 
                select(Sample_ID,
                       Subject_ID,
                       !!sym(params$methane_group),
                       !!sym(x)),
              by = "Sample_ID") %>% 
    # compute subject means
    group_by(Subject_ID, !!sym(params$methane_group)) %>% 
    summarize(across(where(is.numeric), mean, na.rm = TRUE)) %>% 
    ungroup()
  
  xlim <- c(paulaR::trunc_signif(min(plot_data[[x]], na.rm = T), -2),
          paulaR::ceiling_signif(max(plot_data[[x]], na.rm = T), -2))
  
  ylim <- c(-0.02,
            paulaR::ceiling_signif(max(plot_data[[y]], na.rm = T), -2))
  
  
  res_plot <- tibble(
      r = res$estimate_std,
      p = res$p
    ) %>%
    mutate(p_lab = ifelse(
      res$p < 0.001,
      "p < 0.001",
      glue("p = {round(res$p, 3)}")
      ),
      r_lab = glue("{r_or_beta} = {round(r, 3)}"),
      label = glue("{r_lab}, {p_lab}"),
      y_pos = max(plot_data[[y]], na.rm = TRUE),
      x_pos = min(plot_data[[x]], na.rm = TRUE)
     )
  

  # Note the pseudolog scale!!!
  fig_bedu_smi <-
    plot_data %>% 
        select(!!sym(x), !!sym(y), !!sym(params$methane_group)) %>% 
    na.omit() %>% 
      ggplot(aes(x = !!sym(x),
                 y = !!sym(y))
             ) +
      geom_point(aes(fill = !!sym(params$methane_group)), 
                 shape = 21,
                 stroke = 1.2,
                 color = "black",
                 alpha = 0.6,
                 size = 3) +
      scale_fill_manual(values = study_colors$methane) +
      geom_smooth(method = "lm", color = "black", se = F) +
      annotate("text",
               x = xlim[1], 
               y = res_plot$y_pos + 0.3,
               label = res_plot$label,
               size = fontsize/3,
               hjust = 0) +
      scale_x_continuous(
        limits = xlim,
         breaks = seq(xlim[1], xlim[2], 0.01)
                         ) +
      scale_y_continuous(
        trans = scales::pseudo_log_trans(
        base = 2, sigma = 0.0001),
       limits = c(-0.00025, 1),
       breaks = c(0, 0.001, 0.01, 0.1, 1)
      ) +
      labs(title = "",
           x = "M. smithii qPCR signal (1/Ct)",
           y = glue("{y},\nrelative abundance (%)"),
           fill = "",
      ) +
      paulaR::white_theme(font_size = fontsize) +
      theme(legend.position = "bottom")
  
  fig_bedu_smi
}
```


# Save results

## Linear models
```{r}
write.xlsx(lm_res,
             glue("{tab_dir}/{params$markdown_id}_ub_correlations_linear_models.xlsx"))

saveRDS(lm_res,
             glue("{rds_dir}/{params$markdown_id}_ub_correlations_linear_models.rds"))
```

## Figure correlations continuous vars, as images
```{r}
# only if genus
if (params$tax_level == "genus"){
format <- "png" # png # pdf

fig_bedu_smi
ggsave(glue("{fig_dir}/{params$markdown_id}_correlation_Beduinibacterium_Msmithii.{format}"),
       plot = fig_bedu_smi,
       bg = "white",
       width = 4,
       height = 4.5,
       dpi = 400)
}
```

## Figure correlations continuous vars, as plot rds
```{r}
# only if genus
if (params$tax_level == "genus"){
saveRDS(fig_bedu_smi,
        glue("{rds_dir}/{params$markdown_id}_correlation_Beduinibacterium_Msmithii.rds"))
}
```

# Finalise
```{r}
sessionInfo()
```

