---
title: "Microbiome comparison + correlation heatmaps"
author: "Paula Rodriguez-Garcia"
date: "`r Sys.time()`"
output:
  html_document:
    output_dir: "html"
params: 
  markdown_id: "144_genus_counts_mixed" # can be: "144_genus_counts_mixed" or "144_asv_counts_mixed"
  methane_group: "Methane_group"
  
  gc_res_filename: "142_genus_counts_ub_methane_group_tt_linear_models" 
  # !! it assumes it is an .rds file in folder rds_dir
  # can be:
  # - "142_genus_counts_ub_methane_group_tt_linear_models" 
  # - "142_asv_counts_ub_methane_group_tt_linear_models" 

  
  cor_res_filename: "143_genus_counts_mixed_ub_correlations_linear_models"
  # !! it assumes it is an .rds file in folder rds_dir
  # can be:
  # - "143_genus_counts_mixed_ub_correlations_linear_models"
  # - "143_asv_counts_mixed_ub_correlations_linear_models"

  
  tax_level: "genus" # can be: "ASV" or "genus"
  is_gc_means: FALSE # should be FALSE
  is_cor_means: FALSE # should be FALSE
  
  threshold_p: 0.05
  threshold_fdr: 0.1
  filter_unclassified: TRUE # can be true (recommended, for reproducibility) or false
  
---

# Setup

## Knit options
```{r setup}
knitr::opts_chunk$set(warning = FALSE, message = FALSE,
                       root.dir = rprojroot::find_rstudio_root_file())
```

## Libraries
```{r}
library(tibble)
library(dplyr)
library(magrittr)
library(tidyr)
library(readxl)
library(glue)
library(purrr)
library(patchwork)
library(ggplot2)
library(stringr)
library(ggpubr)
library(openxlsx)
library(flextable)
library(effsize)
```

## Study parameters
```{r}
if (file.exists("../../study_parameters.R")) {
  source("../../study_parameters.R")
} else {
  source("study_parameters.R")
}

paulaR::show_rmd_params(params)
```


## Directories
```{r}
```

## Functions
```{r}
```


# Read data
## Results' data
```{r}
# group comparison
gc_res_raw     <- readRDS(glue("{rds_dir}/{params$gc_res_filename}.rds"))

# correlations
cor_res_raw     <- readRDS(glue("{rds_dir}/{params$cor_res_filename}.rds"))
```


# Prepare data

## Extract random effect
```{r}
if (params$is_gc_means){
  
  random_effect <- ""
  random_effect_plus <- ""

} else {
  random_effect <- "(1|Subject_ID)"
  random_effect_plus <- " + (1|Subject_ID)"
}

if (params$is_cor_means){
  cor_label <- "Correlation (R)"
  
} else {
  cor_label <- "Correlation\n(standardized β)"
}

gc_label <- "Group comparison\n(standardized β)"
```

## Identify whether to remove "unclassified" or not
```{r}
if (params$filter_unclassified){
  if (params$tax_level == "genus"){
    filter_unclassified_expr <- expr(!str_detect(response, "nclassified"))
  } else if (params$tax_level =="ASV"){
    filter_unclassified_expr <- expr(!str_detect(closest_tax, "nclassified"))
  }
  
} else {
  filter_unclassified_expr <- TRUE # do not filter anything
}
```



## GC results
```{r}
gc_res <- gc_res_raw %>% 
  # filter unclassified (if specified)
  filter(eval(filter_unclassified_expr)) %>% 
  # filter the adjustments we are interested in
  filter(adjusted_by == random_effect |
          adjusted_by == glue("WGTT{random_effect_plus}")
         ) %>% 
  # filter the responses with any (nominally or FDR) significant results
  group_by(response) %>% 
  mutate(is_signif_unadj = case_when(
    params$tax_level == "genus" ~ any(adjusted_by == random_effect & p < params$threshold_p),
    params$tax_level == "ASV" ~ any(adjusted_by == random_effect & FDR < params$threshold_fdr),
    )) %>%
  ungroup() %>%
  filter(is_signif_unadj) %>%
  select(-is_signif_unadj) %>% 
  # assign a "significance label"
  mutate(
    significance_type = case_when(
      FDR < params$threshold_fdr ~ "fdr",
      p < params$threshold_p & FDR >= params$threshold_fdr ~ "nominal",
      TRUE ~ "ns"
    )
  ) %>% 
  # add an "adjustment"/fixed effect" label
  mutate(adjustment_label = case_when(
    str_detect(adjusted_by, "^\\(1\\|Subject_ID\\)$") ~ "Minimal adjustment",
    str_detect(adjusted_by, "WGTT") ~ "Adjusted for WGTT",
    adjusted_by == "" ~ "No adjustment",
  ),
  adjustment_label = factor(adjustment_label, levels = c(
    "No adjustment",
    "Minimal adjustment", 
    "Adjusted for WGTT"
  ))
  ) %>% 
  # add the taxa label
  mutate(
    tax_label = coalesce(
      if (params$tax_level == "genus") response else NULL,
      if (params$tax_level == "ASV") str_c(closest_tax, " (", response, ")") else NULL
    )
  ) %>% 
  # trim very long taxa
  mutate(
    tax_label = if_else(
      str_length(tax_label) > 40,          
      str_c(word(tax_label, 1), " sp. (", response, ")"),                 
      tax_label                            
    )
  )
```

## Correlation results
```{r}
cor_res <- cor_res_raw %>% 
    # filter unclassified (if specified)
  filter(eval(filter_unclassified_expr)) %>% 
  # filter the significant responses from above
  filter(response %in% gc_res$response) %>% 
  # filter predictors of interest
  filter(predictor %in% c("fasting_CH4", "fasting_H2",
                          "archaea_inv_ct", "smithii_inv_ct")) %>% 
  # filter adjustment of interest
  filter(adjusted_by == "") %>% 
   # assign a "significance label"
  mutate(
    significance_type = case_when(
      FDR < params$threshold_fdr ~ "fdr",
      p < params$threshold_p & FDR >= params$threshold_fdr ~ "nominal",
      TRUE ~ "ns"
    )
  ) %>% 

  # add an "adjustment"/fixed effect" label
  mutate(predictor_label = case_when(
  #  predictor == "stool_energy" ~ "Energy density",
 #   predictor == "bacterial_load_wet" ~ "Microbial load",
  #  predictor == "bacterial_load_dry" ~ "Microbial load (cell counts/g dry feces)",
    predictor == "smithii_inv_ct" ~ "M. smithii",
    predictor == "archaea_inv_ct" ~ "Archaea",
    predictor == "fasting_CH4" ~ "Fasting breath methane",
    predictor == "fasting_H2" ~ "Fasting breath hydrogen",
 # predictor == "ph_proximal_colon" ~ "Proximal colonic pH",
  ),
  predictor_label = factor(
    predictor_label, 
    levels = c(
       "Fasting breath methane",
       "Fasting breath hydrogen",
       "Archaea",
       "M. smithii"
    #   "Microbial load",
      # "Microbial load (cell counts/g dry feces)",
     #  "Energy density",
    #  "Proximal colonic pH"
  ))
  ) %>% 
  # add the taxa label
  mutate(
    tax_label = coalesce(
      if (params$tax_level == "genus") response else NULL,
      if (params$tax_level == "ASV") str_c(closest_tax, " (", response, ")") else NULL
    )
  ) %>% 
  # trim very long taxa
  mutate(
    tax_label = if_else(
      str_length(tax_label) > 40,          
      str_c(word(tax_label, 1), " sp. (", response, ")"),                 
      tax_label                            
    )
  )
```


## Heatmap aesthetics
```{r}
# color gradients
gc_color_max_value <- gc_res %>% 
  pull(estimate_std) %>%
  abs() %>% max(na.rm = TRUE) %>% 
  paulaR::ceiling_signif(digits = 0)

gc_fill_scale <- scale_fill_gradient2(
  low = study_colors$methane[[1]],
  mid = "white",
  high = study_colors$methane[[2]],
  midpoint = 0,
  limits = c(-gc_color_max_value, gc_color_max_value),
  breaks = c(-gc_color_max_value, 0, gc_color_max_value),  
  guide = guide_colorbar(title.position = "top"),
  name = gc_label,
 labels = c(
    glue("{-gc_color_max_value}\nhigher\nin HMP"),
    "0",
    glue("{gc_color_max_value}\nhigher\nin LMP")
  )
)


cor_color_max_value <- cor_res %>% 
  pull(estimate_std) %>%
  abs() %>% max(na.rm = TRUE) %>% 
  paulaR::ceiling_signif(digits = 0)

cor_fill_scale <- scale_fill_gradient2(
  low = study_colors$correlation[[1]],
  mid = study_colors$correlation[[2]],
  high = study_colors$correlation[[3]],  
    midpoint = 0,
    limits = c(-cor_color_max_value, cor_color_max_value),
    breaks = c(-cor_color_max_value, 0, cor_color_max_value),
    guide = guide_colorbar(title.position = "top"),
    name = cor_label
  )

fontsize <- 16
fontsize_x <- ifelse(params$tax_level == "ASV", 6, 14)

```

## Clustering based on results
```{r}
cluster_matrix <- gc_res %>% 
   select(tax_label,
          x = adjustment_label,
          estimate_std) %>%
  # uncomment lines below to also use the corr results for clustering
  bind_rows(cor_res %>% 
              select(tax_label, 
                     x = predictor,
                     estimate_std)) %>% 
  #############################
  pivot_wider(
    names_from = x,
    values_from = estimate_std
  ) %>% 
  column_to_rownames("tax_label") %>% 
  as.matrix()

hc <- hclust(dist(cluster_matrix))
row_order <- rownames(cluster_matrix)[hc$order]
```


# Heatmap of results

## GC heatmap
```{r}
gc_hm <-
gc_res %>% 
  mutate(
    tax_label = factor(tax_label, levels = row_order)
  ) %>% 
  ggplot(aes(x = tax_label,
             y = adjustment_label)) +

  # heatmap tiles
  geom_tile(aes(fill = estimate_std), color = "grey20") +

  # circles nominal P
  geom_point(
    data = gc_res %>% filter(significance_type == "nominal"),
    aes(x = tax_label, 
        y = adjustment_label,
        shape = significance_type),
    color = "black",
    size = 3,
    stroke = 1,
    fill = NA,
    inherit.aes = FALSE,
    show.legend = FALSE
  ) +
  
  # circles FDR
  geom_point(
    data = gc_res %>% filter(significance_type == "fdr"),
    aes(x = tax_label, 
        y = adjustment_label,
        shape = significance_type),
    color = "black",
    size = 3,
    stroke = 1,
    fill = "black",                   
    inherit.aes = FALSE,
    show.legend = FALSE
  ) +

  # tile color scale
  gc_fill_scale +

  # shape scale
  scale_shape_manual(
    name = "",
    values = c(nominal = 1, fdr = 21),
    labels = c(nominal = "p < 0.05", fdr = "q < 0.1")
  ) +

  labs(y = "HMP vs LMP", 
       x = NULL) +

  scale_x_discrete(position = "top") +
  scale_y_discrete(limits = rev) +
  
  paulaR::white_theme(font_size = fontsize) +
  theme(
    axis.text.x = element_text(angle = -45,
                               hjust = 1,
                               size = fontsize_x,
                               color = "black"),
    axis.title.y = ggtext::element_textbox_simple(
            fill = "gray80",        
            color = "black",       
            box.color = "black",    
            size = fontsize - 6,
            linetype = 1,
           linewidth = 0.75,         
            padding = margin(4,4,4,4),
            margin = margin(b = 5, t = 5),
            orientation = "left-rotated",  
            halign = 0.5,       
            valign = 0.5        
          ),
    axis.text.y = element_text(color = "black"),
    legend.title = element_text(size = fontsize-4, face = "bold"),
    legend.text = element_text(size = fontsize-4),
    legend.position = "bottom",
    legend.location = "horizontal",
    panel.grid = element_blank()
  )

gc_hm

```

## Cor heatmap
```{r}
cor_hm <- 
cor_res %>% 
  mutate(
    tax_label = factor(tax_label, levels = row_order)
  ) %>% 
  ggplot(aes(x = tax_label, y = predictor_label)) +
  
  # heatmap tiles
  geom_tile(aes(fill = estimate_std), color = "grey20") +
  
  # circles nominal P
  geom_point(
    data = cor_res %>% filter(significance_type == "nominal"),
    aes(x = tax_label, 
        y = predictor_label,
        shape = significance_type),
    color = "black",
    size = 3,
    stroke = 1,
    fill = NA,
    inherit.aes = FALSE
  ) +
  # circles FDR
  geom_point(
    data = cor_res %>% filter(significance_type == "fdr"),
    aes(x = tax_label, 
        y = predictor_label,
        shape = significance_type),
    color = "black",
    size = 3,
    stroke = 1,
    fill = "black",                   
    inherit.aes = FALSE
  ) +

  # tile color scale
  cor_fill_scale +
  
    # shape scale
  scale_shape_manual(
    name = "",
    values = c(nominal = 1, fdr = 21),
    labels = c(nominal = "p < 0.05", fdr = "q < 0.1")
  ) +
  
guides(
  fill = guide_colourbar(order = 1,  title.position = "top"),
  shape = guide_legend(order = 2)
) +



  labs(
    y = "Correlation",
    x = NULL,
  ) +
  
  scale_x_discrete(position = "top") +
  scale_y_discrete(limits = rev) +
  
  paulaR::white_theme(font_size = fontsize) +
  theme(
    axis.text.x = element_text(angle = -45,
                               hjust = 1,
                               size = fontsize_x,
                               color = "black"),
    axis.text.y = element_text(color = "black"),
    axis.title.y = ggtext::element_textbox_simple(
            fill = "gray80",        
            color = "black",       
            box.color = "black",    
            size = fontsize - 6,
            linetype = 1,
           linewidth = 0.75,         
            padding = margin(4,4,4,4),
            margin = margin(b = 5, t = 5),
            orientation = "left-rotated",  
            halign = 0.5,       
            valign = 0.5        
          ),
    legend.title = element_text(size = fontsize - 4, face = "bold"),
    legend.text = element_text(size = fontsize-4),
    legend.position = "bottom",
    legend.box = "horizontal",
    legend.direction = "horizontal",
    legend.location = "horizontal",
    panel.grid = element_blank()
  )

cor_hm
```

## Join both plots
```{r}
joint_hm <- wrap_plots(list(gc_hm, cor_hm), nrow = 2) +
  plot_layout(
    guides = "collect", 
    axes = "collect_x",
    heights = c(2/10, 4/10)
    ) &
  theme(
    legend.position = "bottom",    
    legend.direction = "horizontal",
    legend.title.align = 0.5
    )

joint_hm
```

## A version for the ASV in vertical
```{r}
if (params$tax_level == "ASV"){
  
  gc_hm_v <- gc_hm + 
    scale_y_discrete(position = "right") +
    scale_x_discrete(limits = rev) +
    labs(
      title = "HMP vs LMP",
      y = NULL,
      x = NULL
    ) +
    guides(guide = guide_colorbar(
      title.position = "top")) +
    coord_flip() +
    theme(
      plot.title = ggtext::element_textbox_simple(
        fill = "gray80",        
            color = "black",       
            box.color = "black",    
            size = fontsize - 6,
            linetype = 1,
           linewidth = 0.75,         
            padding = margin(4,4,4,4),
            margin = margin(b = 5, t = 5),
            orientation = "upright",  
            halign = 0.5,       
            valign = 0.5 
      ),
      axis.text.y = element_text(
        size = fontsize - 6,
        angle = 0,
        color = "black"),
      
      axis.text.x = element_text(
        angle = -90,
        size = fontsize_x,
        color = "black"
      ),
    legend.position = "right",
    legend.direction = "horizontal"
    )
  
  gc_hm_v
  
  cor_hm_v <- cor_hm + 
    scale_y_discrete(position = "right") +
    scale_x_discrete(limits = rev) +
    labs(
      title = "Correlation",
      y = NULL,
      x = NULL
    ) +
    guides(guide = guide_colorbar(
      title.position = "top"),
      shape = guide_legend(
        ncol = 1
        ),
      fill = guide_colorbar(
        direction = "horizontal",
        title.position = "top",
        barwidth = unit(3, "cm"),
        barheight = unit(0.4, "cm")
      ),
    ) +
    coord_flip() +
    theme(
      plot.title = ggtext::element_textbox_simple(
        fill = "gray80",        
            color = "black",       
            box.color = "black",    
            size = fontsize - 6,
            linetype = 1,
           linewidth = 0.75,         
            padding = margin(4,4,4,4),
            margin = margin(b = 5, t = 5),
            orientation = "upright",  
            halign = 0.5,       
            valign = 0.5 
      ),
      axis.text.y = element_text(
        size = fontsize - 6,
        angle = 0,
        color = "black"),
      
      axis.text.x = element_text(
        angle = -90,
        size = fontsize_x,
        color = "black"
      ),
    legend.position = "right",
    legend.box = "vertical",
    )
  
  cor_hm_v
  
  joint_hm_v <-
    wrap_plots(
    list(gc_hm_v, cor_hm_v), ncol = 2) +
  plot_layout(
    guides = "collect", 
    axes = "collect_y",
    widths = c(2/10, 4/10)
    ) &
  theme(
    legend.position = "bottom", 
    legend.direction = "horizontal",
    legend.title.align = 0.5,
     legend.box.margin =  margin(l = -200)
    )

joint_hm_v
}
```


# Save results
```{r}
if (params$tax_level == "genus"){
  width <- 10
  height <- 6
} else if (params$tax_level == "ASV"){
  width <- 19
  height <- 7
}


format <- "png"

joint_hm 
ggsave(glue("{fig_dir}/{params$markdown_id}_microbiome_comparison_heatmap.{format}"),
       bg = "white",
       width = width,
       height = height,
       dpi = 400
       )

saveRDS(joint_hm,
        glue("{rds_dir}/{params$markdown_id}_microbiome_comparison_heatmap.rds"))

# the vertical heatmap if ASV
if (params$tax_level == "ASV"){
  
  joint_hm_v 
ggsave(glue("{fig_dir}/{params$markdown_id}_microbiome_comparison_heatmap_vertical.{format}"),
       bg = "white",
       width = 6,
       height = 14,
       dpi = 400
       )

saveRDS(joint_hm_v,
        glue("{rds_dir}/{params$markdown_id}_microbiome_comparison_heatmap_vertical.rds"))
  
}
```



# Finalise
```{r}
sessionInfo()
```

