---
title: "Clean metabolome data"
author: "Paula Rodriguez-Garcia"
date: "`r Sys.time()`"
output:
  html_document:
    output_dir: "html"
params: 
  markdown_id: "030"
---

There are 2 metabolomics data files to clean:
- Fecal SCFA
  - Contains measured concentrations of SCFA, measured via targeted LCMS at NEXS
- Plasma SCFA
  - Contains measured concentrations of SCFA, measured via targeted LCMS at KU/CBMR metabolomics platform


Additionally:
- This Rmarkdown also calculates subject-means for all resulting files.
- This Rmarkdown also contains the cleaning process of the plasma GLP-1 raw measurements.


The resulting output files will be:

- 1) fecal_scfa.rds
- 2) plasma_scfa.rds
- 3) plasma_glp1.rds

- 4) fecal_scfa_means.rds
- 5) plasma_scfa_means.rds
- 6) plasma_glp1_means.rds



# Setup
## Knit options
```{r setup}
knitr::opts_chunk$set(warning = FALSE, message = FALSE,
                      root.dir = rprojroot::find_rstudio_root_file())
```

## Libraries
```{r}
library(dplyr)
library(magrittr)
library(tidyr)
library(readxl)
library(glue)
library(stringr)
```

## Study paramters
```{r}
if (file.exists("../../study_parameters.R")) {
  source("../../study_parameters.R")
} else {
  source("study_parameters.R")
}

paulaR::show_rmd_params(params)
```

## Directories
```{r}
```


# Read data

## 1. Fecal SCFA/BCFA data
```{r}
f_scfa_raw <- read_excel(
  # path-to-raw-data
  )
```

## 2. Plasma SCFA
```{r}
p_scfa_raw <- read_excel(
  # path-to-raw-data
  )
```

## 3. Plasma GLP-1
```{r}
p_glp1_raw <- read_excel(
  # path-to-raw-data
  )
```

## 4. Meta
```{r}
meta_raw <- readRDS(meta_all_filename)
meta <- meta_raw
```


# 1. Clean fecal SCFA data
## 1.1) Clean and average file
```{r}
f_scfa <- f_scfa_raw %>%
  select(-ID) %>% 
  rename(Sample_ID = Sample,
         Isovalerate = Isavalerate) %>% 
  mutate(Subject_ID = str_split_i(Sample_ID, "-", 1), .before = Sample_ID) %>% 
  select(Sample_ID, everything(), -Day, -Subject_ID) %>% 
  arrange(Sample_ID)

# compute the means on the long files
f_scfa_means <- f_scfa %>%
   mutate(Subject_ID = str_split_i(Sample_ID, "-", 1), .keep = "unused", .before = 1) %>% 
  group_by(Subject_ID) %>% 
  summarise(across(where(is.numeric), mean, na.rm = TRUE)) %>% 
  arrange(Subject_ID)
```

## 1.2) Compute values per dry feces weight
```{r}
f_scfa_dry <-
  f_scfa %>% 
  left_join(meta %>% select(Sample_ID, stool_moisture),
            by = "Sample_ID") %>% 
  mutate(across(
    -c(Sample_ID, stool_moisture),
    ~ .x * (1 / (1 - stool_moisture))
  )) %>% 
  select(-stool_moisture)

# compute the means on the long files
f_scfa_dry_means <- f_scfa_dry %>%
   mutate(Subject_ID = str_split_i(Sample_ID, "-", 1), .keep = "unused", .before = 1) %>% 
  group_by(Subject_ID) %>% 
  summarise(across(where(is.numeric), mean, na.rm = TRUE)) %>% 
  arrange(Subject_ID)
```


# 2. Clean plasma SCFA
```{r}
p_scfa <-
  p_scfa_raw %>% 
  select("Subject_ID" = Subject,
         "Day" = `Time-point`,
         everything(),
         -`Name LCMS file`) %>% 
  naniar::replace_with_na_all(condition = ~str_detect(.x, "LOD")) %>% 
  mutate(Subject_ID = as.character(Subject_ID)) %>% 
  mutate(Sample_ID = str_c(Subject_ID, Day, sep = "-"), .before = 1) %>% 
  select(-Day, -Subject_ID) %>% 
  mutate(across(-all_of(c("Sample_ID")), as.double)) %>% 
  group_by(Sample_ID) %>% 
  summarise(across(-c(), mean, na.rm = T)) %>% 
  ungroup() %>% 
  mutate(across(
    where(is.numeric),
    ~ replace(., is.nan(.), NA_real_)
  )) %>% 
  mutate(Subject_ID = str_split_i(Sample_ID, "-", 1), 
         .after = "Sample_ID") %>% 
  # Rename columns 
  select(Sample_ID, 
         Acetate = `Acetic acid`,
         Propionate = `Propionic acid`,
         Butyrate = `Butyric acid`,
         Isobutyrate = `Isobutyric acid`,
         Valerate = `Valeric acid`,
         Isovalerate = `Isovaleric acid`,
         Caproate = `Hexanoic acid`
         ) %>% 
  arrange(Sample_ID)


p_scfa_means <- p_scfa %>%
  mutate(Subject_ID = str_split_i(Sample_ID, "-", 1), .keep = "unused", .before = 1) %>% 
  group_by(Subject_ID) %>% 
  summarise(across(where(is.numeric), mean, na.rm = TRUE)) %>% 
  ungroup() %>% 
  arrange(Subject_ID)
```

## 3. Clean plasma GLP-1
```{r}
p_glp1 <- 
  p_glp1_raw %>% 
  rename(Subject_ID = ID, 
         "day_2" = `pmol/L...2`,
         "day_9" = `pmol/L...3`) %>% 
    pivot_longer(cols = -Subject_ID,
               names_to = "Day",
               values_to = "glp1") %>% 
  mutate(Day = str_remove(Day, "day_")
         ) %>% 
  mutate(Sample_ID = str_c(Subject_ID, Day, sep = "-"), .before = 1) %>% 
  select(-Subject_ID, -Day) %>% 
  arrange(Sample_ID)


p_glp1_means <- p_glp1 %>%
   mutate(Subject_ID = str_split_i(Sample_ID, "-", 1), .keep = "unused", .before = 1) %>% 
  group_by(Subject_ID) %>% 
  summarise(across(where(is.numeric), mean, na.rm = TRUE)) %>% 
  ungroup() %>% 
  arrange(Subject_ID)
```



# Save data
## 1. Fecal SCFA
```{r}
# wet weight
saveRDS(f_scfa,
        glue("{rds_dir}/{params$markdown_id}_fecal_scfa.rds"))

saveRDS(f_scfa_means,
        glue("{rds_dir}/{params$markdown_id}_fecal_scfa_means.rds"))


# dry weight
saveRDS(f_scfa_dry,
        glue("{rds_dir}/{params$markdown_id}_fecal_scfa_dry.rds"))

saveRDS(f_scfa_dry_means,
        glue("{rds_dir}/{params$markdown_id}_fecal_scfa_dry_means.rds"))
```

## 2. Plasma SCFA
```{r}
saveRDS(p_scfa ,
        glue("{rds_dir}/{params$markdown_id}_plasma_scfa.rds"))

saveRDS(p_scfa_means,
        glue("{rds_dir}/{params$markdown_id}_plasma_scfa_means.rds"))
```

## 3. Plasma GLP-1
```{r}
saveRDS(p_glp1,
        glue("{rds_dir}/{params$markdown_id}_plasma_glp1.rds"))

saveRDS(p_glp1_means,
        glue("{rds_dir}/{params$markdown_id}_plasma_glp1_means.rds"))
```


# Finalise
```{r}
sessionInfo()
```

