---
title: "SCFA plasma-fecal correlation, methane group comparison, and correlation to continuous vars"
author: "Paula Rodriguez-Garcia"
date: "`r Sys.time()`"
output:
  html_document:
    output_dir: "html"
params: 
  markdown_id: "171_dry_long" # can be: "171_long" or "171_dry_long"
  
  fecal_scfa_filename: "fecal_scfa_dry_filename" # "fecal_scfa_filename" or "fecal_scfa_dry_filename" 
  
  methane_group: "Methane_group"
  
  cor_method: "spearman"
  
  is_mb_means: FALSE # must be FALSE
  
  pseudocount: 1e-6 # potentially needed for log-transformation
  threshold_p: 0.05
  threshold_fdr: 0.1
---

# Setup

## Knit options
```{r setup}
knitr::opts_chunk$set(warning = FALSE, message = FALSE,
                       root.dir = rprojroot::find_rstudio_root_file())
```

## Libraries
```{r}
library(tibble)
library(dplyr)
library(magrittr)
library(tidyr)
library(glue)
library(purrr)
library(patchwork)
library(ggplot2)
library(stringr)
library(ggpubr)
library(openxlsx)
library(flextable)
library(broom)
```

## Study paramters
```{r}
if (file.exists("../../study_parameters.R")) {
  source("../../study_parameters.R")
} else {
  source("study_parameters.R")
}

paulaR::show_rmd_params(params)
```


## Directories
```{r}
```

## Functions
```{r}
```

## General aesthetics
```{r}
fontsize <- 14
```

# Read data
## Meta
```{r}
meta_raw <- readRDS(meta_all_filename)
```

## SCFA data
```{r}
plasma_raw     <- readRDS(plasma_scfa_filename)
fecal_raw      <- readRDS(get(params$fecal_scfa_filename))
# extra
glp1_raw       <- readRDS(plasma_glp1_filename)
```

# Explore data
## General
```{r}
plasma_raw %>% colnames()
fecal_raw %>% colnames()
glp1_raw %>% colnames()

plasma_raw %>% dim() 
fecal_raw %>% dim() 
glp1_raw %>% dim() 

plasma_raw %>% mutate(Day = str_split_i(Sample_ID, "-", 2)) %>% group_by(Day) %>% count() # days 2 and 9
fecal_raw %>% mutate(Day = str_split_i(Sample_ID, "-", 2)) %>% group_by(Day) %>% count() # days 2,3,4,5,6
glp1_raw %>% mutate(Day = str_split_i(Sample_ID, "-", 2)) %>% group_by(Day) %>% count() # days 2 and 9
```


# Prepare data

## Add a prefix to the fecal and plasma variables
```{r}
plasma <- plasma_raw %>% rename_with(~ paste0("p_", .x), where(is.numeric)) 
fecal  <- fecal_raw  %>% rename_with(~ paste0("f_", .x), where(is.numeric))
glp1   <- glp1_raw

# meta - nothing needed
meta <- meta_raw

# for now only - remove ph from those with smartpill fucked
subjects_any_na_smartpill <- meta_raw %>% 
  select(Subject_ID,
         contains("TT"),
         -contains("Methane"), -contains("stool")) %>% 
  paulaR::show_rows_with_na() %>% 
  pull(Subject_ID) %>% unique()

meta <- meta_raw %>% 
  mutate(
    ph_stomach = ifelse(Subject_ID %in% subjects_any_na_smartpill, NA, ph_stomach),
    ph_duodenum = ifelse(Subject_ID %in% subjects_any_na_smartpill, NA, ph_duodenum),
    ph_small_bowel = ifelse(Subject_ID %in% subjects_any_na_smartpill, NA, ph_small_bowel),
    ph_colon = ifelse(Subject_ID %in% subjects_any_na_smartpill, NA, ph_colon),
    ph_proximal_colon = ifelse(Subject_ID %in% subjects_any_na_smartpill, NA, ph_proximal_colon),
    ph_distal_colon = ifelse(Subject_ID %in% subjects_any_na_smartpill, NA, ph_distal_colon),
    ph_sigmoid_colon = ifelse(Subject_ID %in% subjects_any_na_smartpill, NA, ph_sigmoid_colon),
    ph_rectum = ifelse(Subject_ID %in% subjects_any_na_smartpill, NA, ph_rectum),
    GET = ifelse(Subject_ID %in% subjects_any_na_smartpill, NA, GET),
    SBTT = ifelse(Subject_ID %in% subjects_any_na_smartpill, NA, SBTT),
    SLBTT = ifelse(Subject_ID %in% subjects_any_na_smartpill, NA, SLBTT),
    CTT = ifelse(Subject_ID %in% subjects_any_na_smartpill, NA, CTT),
    WGTT = ifelse(Subject_ID %in% subjects_any_na_smartpill, NA, WGTT),
    
  )
```

## Extract variables of interest
```{r}
plasma_vars <- plasma %>% select(where(is.numeric)) %>% colnames()
fecal_vars  <- fecal  %>% select(where(is.numeric)) %>% colnames()

plasma_vars_glp1 <- c(plasma_vars, "glp1")
fecal_vars_glp1  <- c(fecal_vars,  "glp1")

extra_vars <- c("fasting_CH4",
                "fasting_H2",
                "stool_energy",
                "bacterial_load_wet",
                "ph_small_bowel",
                "ph_colon",
                "ph_proximal_colon",
                "ph_distal_colon",
                "ph_sigmoid_colon",
                "ph_rectum",
                "smithii_inv_ct", 
                "archaea_inv_ct")

tt_vars <- c("GET", "SBTT", "CTT", "SLBTT", "WGTT")
# for now keep it simple
tt_vars <- tt_vars[4:5]

combi <- expand_grid(
  x = c(fecal_vars_glp1, extra_vars),
  y = c(plasma_vars, fecal_vars, "glp1"))

gc_vars <- c(plasma_vars, fecal_vars, "glp1")

combi_gc <- expand_grid(
  y = gc_vars,
  fixed_effect = c("", tt_vars)
)

combi_all <- expand_grid(
  x = c(fecal_vars_glp1, extra_vars),
  y = c(plasma_vars, fecal_vars, "glp1"),
  fixed_effect = c("", tt_vars)
)

vars_for_means <- c("stool_energy",
                    "cn_ratio",
                    "ph_small_bowel",
                    "ph_colon",
                    "ph_proximal_colon",
                    "ph_distal_colon",
                    "ph_sigmoid_colon",
                    "ph_rectum",
                    "fasting_CH4",
                    "fasting_H2",
                    "smithii_inv_ct",
                    "archaea_inv_ct",
                    "glp1")
```

## Combine all data sets
```{r}
data <- plasma %>% 
    full_join(fecal,
              by = "Sample_ID") %>% 
    full_join(glp1, by = "Sample_ID") %>% 
    mutate(Subject_ID = str_split_i(Sample_ID , "-", 1),
           .after = Sample_ID) %>% 
    left_join(meta %>%
                select(Sample_ID,
                       !!sym(params$methane_group),
                       sex,
                       all_of(tt_vars),
                       all_of(extra_vars)) %>%
                distinct(),
              by = "Sample_ID") %>% 
  filter(!is.na(!!sym(params$methane_group))) %>% 
  arrange(Sample_ID)
```


# 1) Statistical tests

## 1.1) Correlations - longitudinal except for "cross-sectional" variables
```{r}
if (!params$is_mb_means){

  
  cor_res <- pmap(combi_all, function(x, y, fixed_effect){
    

     if (x == y){
       output <- tibble::tibble(response = y,
                    predictor = x,
                    test = NA,
                    adjusted_by = NA,
                    n_total = NA,
                    estimate_type = NA,
                    estimate = NA,
                    estimate_std = NA,
                    p = NA
                   )

     } else if (x %in% vars_for_means) {
       
       n_total <- data$Subject_ID %>% unique() %>% length()
       if (fixed_effect == "") {
                # compute means
       model_data <- data %>%
      dplyr::select(any_of(c(x, y, "Subject_ID"))) %>% 
      group_by(Subject_ID) %>% 
      summarise(across(all_of(c(x, y)), ~ mean(.x, na.rm = TRUE))) %>% 
      mutate(!!sym(y) := !!sym(y) + as.numeric(params$pseudocount)) %>% 
      na.omit() %>% 
      # transform X already here
         mutate(!!sym(x) := case_when(
           x %in% c(plasma_vars, fecal_vars) ~ log(!!sym(x) + as.numeric(params$pseudocount)),
           TRUE ~ scale(!!sym(x))
           )) 
  
      n_tested <- nrow(model_data)
         
         
         
          f <- as.formula(paste0("log(`", y, "`) ~ `", x, "`"))
        } else {
          
      # compute means
       model_data <- data %>%
      dplyr::select(any_of(c(x, y, fixed_effect, "Subject_ID"))) %>% 
      group_by(Subject_ID) %>% 
      summarise(across(all_of(c(x, y, fixed_effect)), ~ mean(.x, na.rm = TRUE)))  %>%
      mutate(!!sym(y) := !!sym(y) + as.numeric(params$pseudocount)) %>% 
      na.omit() %>% 
      # transform X already here
         mutate(!!sym(x) := case_when(
           x %in% c(plasma_vars, fecal_vars) ~ log(!!sym(x) + as.numeric(params$pseudocount)),
           TRUE ~ scale(!!sym(x))
           )) 
  
      n_tested <- nrow(model_data)
          
          
          f <- as.formula(paste0("log(`", y, "`) ~ `", x, "` + scale(`", fixed_effect, "`)"))

        }
        
        random_effect <- ""
        model <- lm(f, data = model_data)
        
         pval <- as.numeric(summary(model)$coefficients[2, "Pr(>|t|)"])
    
      coef <- as.numeric(summary(model)$coefficients[2, "Estimate"])
        se      <- as.numeric(summary(model)$coefficients[2, "Std. Error"])
  ci_low  <- coef - 1.96 * se
  ci_high <- coef + 1.96 * se
        
  # extract std beta (in log-transformed units)
  sd_y <- sd(log(model_data[[y]]), na.rm = TRUE)
  coef_std <- coef / sd_y
       
       
       
       output <- tibble::tibble(
         response = y,
         predictor = x,
         test = "linear model",
         adjusted_by = glue("{fixed_effect}"),
         n_total = glue("{n_tested}/{n_total}"),
         estimate_type = "beta coefficient",
         estimate = coef,
         estimate_std = coef_std,
         CI_estimate_lower = ci_low,
         CI_estimate_upper = ci_high,
         p = pval )
       
       
       
     } else {
        
      model_data <- data %>%
      dplyr::select(any_of(c(x, y, fixed_effect, "Subject_ID"))) %>% 
      mutate(!!sym(y) := !!sym(y) + as.numeric(params$pseudocount)) %>% 
      na.omit() %>% 
      # transform X already here
         mutate(!!sym(x) := case_when(
           x %in% c(plasma_vars, fecal_vars) ~ log(!!sym(x) + as.numeric(params$pseudocount)),
           TRUE ~ scale(!!sym(x))
           )) 
      
       n_total <- nrow(data)
      n_tested <- nrow(model_data)
    
  
      # if only 1 measurement per subject, then do not adjust for subject
      if (nrow(model_data) == n_distinct(model_data$Subject_ID)){
        if (fixed_effect == "") {
          f <- as.formula(paste0("log(`", y, "`) ~ `", x, "`"))
        } else {
          
          f <- as.formula(paste0("log(`", y, "`) ~ `", x, "` + scale(`", fixed_effect, "`)"))

        }
        
          random_effect <- ""
          model <- lm(f, data = model_data)
           test_label <- "linear model"
       
          # linear mixed model 
      } else {
        
        if (fixed_effect == "") {
          f <- as.formula(paste0("log(`", y, "`) ~ `", x, "` + (1|Subject_ID)"))

        } else {
          
          f <- as.formula(paste0("log(`", y, "`) ~ `", x, "` + scale(`", fixed_effect, "`) + (1|Subject_ID)"))

        }

        random_effect <- "(1|Subject_ID)"
        model <- lmerTest::lmer(f, data = model_data)
        test_label <- "linear mixed model"
      }
    
      pval <- as.numeric(summary(model)$coefficients[2, "Pr(>|t|)"])
    
      coef <- as.numeric(summary(model)$coefficients[2, "Estimate"])
        se      <- as.numeric(summary(model)$coefficients[2, "Std. Error"])
  ci_low  <- coef - 1.96 * se
  ci_high <- coef + 1.96 * se
  
  # extract std beta (in log-transformed units)
  sd_y <- sd(log(model_data[[y]]), na.rm = TRUE)
  coef_std <- coef / sd_y
       
  
      
       
      output <- tibble::tibble(
         response = y,
         predictor = x,
         test = test_label,
         adjusted_by = glue("{fixed_effect}, {random_effect}"),
         n_total = glue("{n_tested}/{n_total}"),
         estimate_type = "beta coefficient",
         estimate = coef,
         estimate_std = coef_std,
         CI_estimate_lower = ci_low,
         CI_estimate_upper = ci_high,
         p = pval
                     )
     }
     
     return(output)
    
    
  }) %>% bind_rows() %>% 
  mutate(yx = str_c(response, predictor, sep = "-"),
         xy = str_c(predictor, response, sep = "-"),
         ) %>% 
  filter(yx != xy) %>% 
  select(-xy, -yx) %>% 
  mutate(type_response = case_when(
    str_detect(response, "p_") ~ "plasma",
    str_detect(response, "f_") ~ "fecal",
    str_detect(response, "glp1") ~ "glp1",
    TRUE ~ "stool"
    ),
    .after = "response") %>%
    mutate(type_predictor = case_when(
    str_detect(predictor, "p_") ~ "plasma",
    str_detect(predictor, "f_") ~ "fecal",
    str_detect(predictor, "glp1") ~ "glp1",
    str_detect(predictor, "ph_") ~ "pH",
    str_detect(predictor, "fasting_") ~ "breath",
    str_detect(predictor, "qpcr") ~ "methanogens",
    str_detect(predictor, "inv_ct") ~ "methanogens",
    TRUE ~ "stool"
    ),
    .after = "predictor") %>%
  group_by(type_response, type_predictor, adjusted_by) %>%
  mutate(FDR = p.adjust(p, method = "BH"), .after = "p") %>%
  ungroup() %>% 
  mutate(signif_p = get_significance_asterisks(p, asterisks_p), .after = "p") %>%
    mutate(signif_FDR = get_significance_asterisks(FDR, asterisks_fdr), .after = "FDR")
  
}
```

## 1.2) Methane group comparison - longitudinal
```{r}
if (!params$is_mb_means){
  x <- params$methane_group
  
  total_all <- nrow(data)
  total_hmp <- nrow(data %>% filter(!!sym(params$methane_group) == "HMP"))
  total_lmp <- nrow(data %>% filter(!!sym(params$methane_group) == "LMP"))
  
  gc_res <- pmap(combi_gc, function(y, fixed_effect){
    
    model_data <- data %>%
      dplyr::select(any_of(c(x, y, fixed_effect, "Subject_ID"))) %>% 
      mutate(!!sym(y) := !!sym(y) + as.numeric(params$pseudocount)) %>% 
      na.omit()
    
    subject_counts <- model_data %>%
      group_by(!!sym(x)) %>% 
      count()

    n_hmp <- subject_counts[[1,2]]
    n_lmp <- subject_counts[[2,2]]
    n_total <- n_lmp + n_hmp
    n_tested <- nrow(model_data)
    
    # if only 1 measurement per subject, then do not adjust for subject
    if (nrow(model_data) == n_distinct(model_data$Subject_ID)){
      
     if (fixed_effect == "") {
          f <- as.formula(paste0("log(`", y, "`) ~ `", x, "`"))
        } else {
          f <- as.formula(paste0("log(`", y, "`) ~ `", x, "` + scale(`", fixed_effect, "`)"))
        }
      
      random_effect <- ""
      model <- lm(f, data = model_data)
      model <- lmerTest::lmer(f, data = model_data)
      test_label <- "linear model"
      
    } else {

        if (fixed_effect == "") {
          f <- as.formula(paste0("log(`", y, "`) ~ `", x, "` + (1|Subject_ID)"))

        } else {
          
          f <- as.formula(paste0("log(`", y, "`) ~ `", x, "` + scale(`", fixed_effect, "`) + (1|Subject_ID)"))

        }
      
      
      random_effect <- "(1|Subject_ID)"
      model <- lmerTest::lmer(f, data = model_data)
      test_label <- "linear mixed model"
    }
    
    
      pval <- as.numeric(summary(model)$coefficients[2, "Pr(>|t|)"])
    
      coef <- as.numeric(summary(model)$coefficients[2, "Estimate"])
      log2FC <- coef / log(2)
      
      
        se      <- as.numeric(summary(model)$coefficients[2, "Std. Error"])
   ci_log <- coef + c(-1, 1) * 1.96 * se
      ci_log2 <- ci_log / log(2)
  
  # extract std beta (in log-transformed units)
  sd_y <- sd(log(model_data[[y]]), na.rm = TRUE)
  coef_std <- coef / sd_y
       
    tibble::tibble(response = y,
                   predictor = x,
                   test = test_label,
                   adjusted_by = glue("{fixed_effect}, {random_effect}"),
                   n_total = glue("{n_tested}/{n_total}"),
                   n_HMP = glue("{n_hmp}/{total_hmp}"),
                   n_LMP = glue("{n_lmp}/{total_lmp}"),
                    estimate_type = "log2FC",
         estimate = log2FC,
         estimate_std = coef_std,
          CI_log2FC_lower = ci_log2[1],
                     CI_log2FC_upper = ci_log2[2],
         p = pval
                   )
    
    }) %>% bind_rows() %>% 
 mutate(type_response = case_when(
    str_detect(response, "p_") ~ "plasma",
    str_detect(response, "f_") ~ "fecal",
    str_detect(response, "glp1") ~ "glp1",
    TRUE ~ "stool"
    ),
    .after = "response") %>%
  group_by(type_response, adjusted_by) %>%
  mutate(FDR = p.adjust(p, method = "BH"), .after = "p") %>%
  ungroup() %>% 
  mutate(signif_p = get_significance_asterisks(p, asterisks_p), .after = "p") %>%
    mutate(signif_FDR = get_significance_asterisks(FDR, asterisks_fdr), .after = "FDR")
}
```


# 2) Heatmap plasma + fecal SCFA

## 2.1) Edit results data for heatmap
```{r}
gc_fp_scfa_data <-
 gc_res %>% 
  filter(type_response %in% c("fecal", "plasma")) %>% 
   filter(adjusted_by ==  ", (1|Subject_ID)" |
        # str_detect(adjusted_by, "SLBTT") |
         str_detect(adjusted_by, "WGTT")
         ) %>% 
  mutate(response = str_replace_all(response, "f_", "")) %>%
  mutate(response = str_replace_all(response, "p_", "")) %>%
  mutate(adjustment_label = case_when(
    str_detect(adjusted_by, "WGTT") ~ "Adjusted for WGTT",
    str_detect(adjusted_by, "SLBTT") ~ "Adjusted for ITT",
    adjusted_by == ", (1|Subject_ID)" ~ "Minimal adjustment",
     adjusted_by == "" ~ "Minimal adjustment",
  ),
  adjustment_label = factor(adjustment_label, levels = c(
    "Minimal adjustment", 
  #  "Adjusted for ITT",
    "Adjusted for WGTT"
  ))) %>% 
    mutate(type_response = case_when(
    type_response == "plasma" ~ "Plasma SCFA",
    type_response == "fecal" ~ "Faecal SCFA",
    TRUE ~ "Other (check code)",
  )) %>% 
  # assign a "significance label"
  mutate(
    significance_type = case_when(
      FDR < params$threshold_fdr ~ "fdr",
      p < params$threshold_p & FDR >= params$threshold_fdr ~ "nominal",
      TRUE ~ "ns"
    )
  )


cor_fp_scfa_data <- cor_res %>% 
  filter(type_response %in% c("fecal", "plasma")) %>% 
  filter(type_predictor %in% c("pH", "breath", "methanogens")) %>% 
  filter(!predictor %in% c("ph_small_bowel", "ph_colon", "ph_rectum")) %>% 
  filter(adjusted_by == "" | adjusted_by == ", (1|Subject_ID)") %>% 
 mutate(response = str_replace_all(response, "f_", "")) %>%
  mutate(response = str_replace_all(response, "p_", "")) %>%
 mutate(type_predictor = forcats::fct_collapse(
    type_predictor,
    "Breath and\nmethanogens" = c("breath", "methanogens")
  )) %>% 
  mutate(type_response = case_when(
    type_response == "plasma" ~ "Plasma SCFA",
    type_response == "fecal" ~ "Faecal SCFA",
    TRUE ~ "Other (check code)",
  )) %>% 
   # assign a "significance label"
  mutate(
    significance_type = case_when(
      FDR < params$threshold_fdr ~ "fdr",
      p < params$threshold_p & FDR >= params$threshold_fdr ~ "nominal",
      TRUE ~ "ns"
    )
  ) %>% 
  # add an "adjustment"/fixed effect" label
  mutate(predictor_label = case_when(
     predictor == "smithii_inv_ct" ~ "M. smithii",
    predictor == "archaea_inv_ct" ~ "Archaea",
    predictor == "fasting_CH4" ~ "Fasting breath methane",
    predictor == "fasting_H2" ~ "Fasting breath hydrogen",
    predictor == "ph_proximal_colon" ~ "Proximal colonic pH",
    predictor == "ph_distal_colon" ~ "Distal colonic pH",
    predictor == "ph_sigmoid_colon" ~ "Sigmoid colonic pH",
  #  predictor == "ph_rectum" ~ "Rectal pH",
  #  predictor == "ph_small_bowel" ~ "Small bowel pH",
  #  predictor == "ph_colon" ~ "Colonic pH",
  TRUE ~ predictor
  ),
  predictor_label = factor(
    predictor_label,
    levels = c(
       "Fasting breath methane",
       "Fasting breath hydrogen",
       "Archaea",
       "M. smithii",
     # "Small bowel pH",
    #  "Colonic pH",
      "Proximal colonic pH",
      "Distal colonic pH",
      "Sigmoid colonic pH"
   #   "Rectal pH"
  ))
  )
```


## 2.2) Heatmap aesthetics
```{r}
# color gradients
gc_color_max_value <- gc_fp_scfa_data %>% 
  pull(estimate_std) %>%
  abs() %>% max(na.rm = TRUE) %>% 
  paulaR::ceiling_signif(digits = 0)

gc_fill_scale <- scale_fill_gradient2(
  low = study_colors$methane[[1]],
  mid = "white",
  high = study_colors$methane[[2]],
  midpoint = 0,
  limits = c(-gc_color_max_value, gc_color_max_value),
  breaks = c(-gc_color_max_value, 0, gc_color_max_value),  
  guide = guide_colorbar(title.position = "top"),
  name = "Group comparison\n(standardized β)",
 labels = c(
    glue("{-gc_color_max_value}\nhigher\nin HMP"),
    "0",
    glue("{gc_color_max_value}\nhigher\nin LMP")
  )
)


cor_color_max_value <- cor_fp_scfa_data %>% 
  pull(estimate_std) %>%
  abs() %>% max(na.rm = TRUE) %>% 
  paulaR::ceiling_signif(digits = 0)

cor_fill_scale <- scale_fill_gradient2(
  low = study_colors$correlation[[1]],
  mid = study_colors$correlation[[2]],
  high = study_colors$correlation[[3]],  
    midpoint = 0,
    limits = c(-cor_color_max_value, cor_color_max_value),
    breaks = c(-cor_color_max_value, 0, cor_color_max_value),
    guide = guide_colorbar(title.position = "top"),
    name = "Correlation\n(standardized β)"
  )

fontsize <- 14
fontsize_hm <- 16
```

## 2.3) GC heatmap
```{r}
gc_fp_scfa_hm <-
  gc_fp_scfa_data %>% 
  
  ggplot(aes(x = response,
             y = adjustment_label)) +

  # heatmap tiles
  geom_tile(aes(fill = estimate_std), color = "grey20") +

  facet_grid( ~ type_response,
             scales = "free", space = "free", switch = "y") + 
  
  # circles nominal P
  geom_point(
    data = gc_fp_scfa_data %>% filter(significance_type == "nominal"),
    aes(x = response, 
        y = adjustment_label,
        shape = significance_type),
    color = "black",
    size = 3,
    stroke = 1,
    fill = NA,
    inherit.aes = FALSE,
    show.legend = FALSE
  ) +
  
  # circles FDR
  geom_point(
    data = gc_fp_scfa_data %>% filter(significance_type == "fdr"),
    aes(x = response, 
        y = adjustment_label,
        shape = significance_type),
    color = "black",
    size = 3,
    stroke = 1,
    fill = "black",                   
    inherit.aes = FALSE,
    show.legend = FALSE
  ) +

  # tile color scale
  gc_fill_scale +

  # shape scale
  scale_shape_manual(
    name = "",
    values = c(nominal = 21, fdr = 21),
    labels = c(nominal = "p < 0.05", fdr = "q < 0.1")
  ) +

  labs(y = "HMP vs LMP", 
       x = NULL) +

  scale_x_discrete(position = "top") +
  scale_y_discrete(limits = rev) +
  
  paulaR::white_theme(font_size = fontsize_hm) +
  theme(
    axis.text.x = element_text(angle = -45,
                               hjust = 1,
                               size = fontsize_hm,
                               color = "black"),
        axis.title.y = ggtext::element_textbox_simple(
            fill = "gray80",        
            color = "black",       
            box.color = "gray80", #"black",    
            size = fontsize_hm - 6,
            linetype = 1,
           linewidth = 0.75,         
            padding = margin(4,4,4,4),
            margin = margin(b = 5, t = 5),
            orientation = "left-rotated",  
            halign = 0.5,       
            valign = 0.5        
          ),
    axis.text.y = element_text(color = "black"),
    legend.title = element_text(size = fontsize_hm-4, face = "bold"),
    legend.text = element_text(size = fontsize_hm-4),
    legend.position = "bottom",
    legend.location = "horizontal",
    panel.grid = element_blank()
  )

gc_fp_scfa_hm
```

## 2.4) Cor heatmap
```{r}
cor_fp_scfa_hm <-
  cor_fp_scfa_data %>% 
  ggplot(aes(x = response, y = predictor_label)) +
  
  # heatmap tiles
  geom_tile(aes(fill = estimate_std), color = "grey20") +
  
  facet_grid(type_predictor ~ type_response,
             scales = "free", space = "free", switch = "y") + 


  # circles nominal P
  geom_point(
    data = cor_fp_scfa_data %>% filter(significance_type == "nominal"),
    aes(x = response, 
        y = predictor_label,
        shape = significance_type),
    color = "black",
    size = 3,
    stroke = 1,
    fill = NA,
    inherit.aes = FALSE
  ) +
  # circles FDR
  geom_point(
    data = cor_fp_scfa_data %>% filter(significance_type == "fdr"),
    aes(x = response, 
        y = predictor_label,
        shape = significance_type),
    color = "black",
    size = 3,
    stroke = 1,
    fill = "black",                   
    inherit.aes = FALSE
  ) +

  # tile color scale
  cor_fill_scale +
  
    # shape scale
  scale_shape_manual(
    name = "",
    values = c(nominal = 21, fdr = 21),
    labels = c(nominal = "p < 0.05", fdr = "q < 0.1")
  ) +


  labs(
    y = NULL,
    x = NULL,
  ) +
  
  scale_x_discrete(position = "top") +
  scale_y_discrete(limits = rev) +
  
  guides(
    fill = guide_colorbar(order = 1,
                          title.position = "top"),
    shape = guide_legend(order = 2)
  ) + 
  
  paulaR::white_theme(font_size = fontsize_hm) +
  theme(
    axis.text.x = element_text(angle = -45,
                               hjust = 1,
                               size = fontsize_hm,
                               color = "black"),
    axis.text.y = element_text(color = "black"),
    axis.title.y = ggtext::element_textbox_simple(
            fill = "gray80",        
            color = "black",       
            box.color = "black",    
            size = fontsize_hm - 8,
            linetype = 1,
           linewidth = 0.75,         
            padding = margin(4,4,4,4),
            margin = margin(b = 5, t = 5),
            orientation = "left-rotated",  
            halign = 0.5,       
            valign = 0.5        
          ),
    legend.title = element_text(size = fontsize_hm - 4, face = "bold"),
    legend.text = element_text(size = fontsize_hm-4),
    legend.position = "bottom",
    legend.location = "horizontal",
    panel.grid = element_blank()
  )

cor_fp_scfa_hm
```

## 2.5) Join both plots
```{r}
joint_fp_scfa_hm <- wrap_plots(
  list(gc_fp_scfa_hm + theme(
    axis.text.x.top = element_text(angle = -30)
  ),
       cor_fp_scfa_hm + theme(
          axis.line.x.top = element_blank(), 
          strip.text.x = element_blank(),
          strip.background.x = element_blank(),
          axis.text.x.top = element_blank(),
          axis.ticks.x.top = element_blank()
          )
       ),
  nrow = 2) +
  plot_layout(
    guides = "collect", 
    axes = "collect",
    heights = c(2/9, 7/9)
    ) &
  theme(
    legend.position = "bottom",    
    legend.direction = "horizontal",
    legend.title.align = 0.5,
    legend.box.margin =  margin(l = -50)
    )

joint_fp_scfa_hm
```





# 3) PCA fecal SCFA
```{r}
if (!params$is_mb_means){

  top_k <- length(fecal %>% select(-Sample_ID))
  loadings_factor <- 10
  
  loadings_offset_y_pos <- 0.01
  loadings_offset_y_neg <- -0.01
  
  loadings_offset_x_pos <- 0.01
  loadings_offset_x_neg <- -0.01

  
  fecal_pca_data <- fecal %>% rename_with(~ str_remove(., "f_")) %>% na.omit()
  
  pca <- prcomp(fecal_pca_data %>% select(-Sample_ID), scale = T)
  
  scores <- fecal_pca_data %>%
    select(Sample_ID) %>% 
    left_join(meta %>% select(Sample_ID, Subject_ID, params$methane_group),
              by = "Sample_ID") %>% 
    bind_cols(data.frame(pca$x[ , 1:top_k]))
  
  
  loadings <- pca$rotation[ , 1:top_k] %>% 
    as_tibble(rownames = "var") %>% 
    mutate(
      y_offset = ifelse(PC2 > 0, loadings_offset_y_pos, loadings_offset_y_neg),
      x_offset = ifelse(PC1 > 0, loadings_offset_x_pos, loadings_offset_x_neg),
      label_x = PC1*loadings_factor + x_offset,
      label_y = PC2*loadings_factor + y_offset
    )
  
  importances <- summary(pca)
  
  ylim <- c(paulaR::ceiling_signif(min(scores$PC2), 0)-1,
          paulaR::ceiling_signif(max(scores$PC2), 0)+2)
  
  xlim <- c(paulaR::ceiling_signif(min(scores$PC1), 0),
          paulaR::ceiling_signif(max(scores$PC1), 0)+1)
  
  ybreaks <- seq(ylim[1], ylim[2], 2)
  xbreaks <- seq(xlim[1], xlim[2], 2)
  
  scores_plot <- 
    scores %>% 
  ggplot() + 
  geom_point(aes(PC1,
                 PC2,
                 color = !!sym(params$methane_group),
                 fill = !!sym(params$methane_group)
                 ),
             shape = 21, alpha = 0.5, size = 1) +
  geom_hline(yintercept = 0,
             color = "gray80",
             linetype = "dashed") +
  geom_vline(xintercept = 0,
             color = "gray80",
             linetype = "dashed")  +
  stat_ellipse(data = scores, 
               aes(x = PC1, y = PC2,
                   group = !!sym(params$methane_group),
                   fill = !!sym(params$methane_group)),
               geom = "polygon",
               alpha = 0.1,
              # linetype = "dashed"
               ) +
  scale_fill_manual(values = study_colors$methane) +
  scale_color_manual(values = study_colors$methane) +
  labs(x = glue("PC1 ({as.character(signif(importances$importance[2,1]*100, 3))}%)"),
       y = glue("PC2 ({as.character(signif(importances$importance[2,2]*100, 3))}%)"),
       fill = "",
       color = "") +
  scale_y_continuous(limits = ylim,
                     breaks = ybreaks) +
    scale_x_continuous(limits = xlim,
                       breaks = xbreaks) +
  geom_segment(data = loadings %>% select(-var),
               aes(x = 0,
                   y = 0,
                   xend = PC1*loadings_factor, # to make the arrows more far from the center
                   yend = PC2*loadings_factor),
             color = "gray30",
             arrow = arrow(angle = 25, 
                           length = unit(4, "pt"))
             ) +
  ggrepel::geom_text_repel(data = loadings,
                  aes(x = label_x,
                      y = label_y,
                      label = var),
            segment.color = "black",
            size = fontsize/3,
            vjust = 1,
            hjust = 1,
            # nudge_x = 0.5,
            # nudge_y = 0.5,
            max.overlaps = Inf,
            segment.linetype = "dashed"
            ) +
  paulaR::white_theme(font_size = fontsize) +
  theme(legend.position = "bottom",
        plot.margin = margin(5,5,5,5))



test_pc1 <- lmerTest::lmer(
  as.formula(paste("PC1 ~", params$methane_group, "+ (1|Subject_ID)")),
  data = scores) %>%
  broom.mixed::tidy() %>%
  rename("p" = `p.value`) %>% 
  filter(str_detect(term, params$methane_group)) %>% 
  mutate(group1 = "HMP", 
         group2 = "LMP",
    label = case_when(
    p < 0.001 ~ "p < 0.001",
    p > 0.1 ~ "p > 0.1",
    TRUE ~ glue("p = {round(p, 3)}")
    ),
    y.position = min(scores$PC1)*1.05)

test_pc2 <- lmerTest::lmer(
  as.formula(paste("PC2 ~", params$methane_group, "+ (1|Subject_ID)")),
  data = scores) %>%
  broom.mixed::tidy() %>%
  rename("p" = `p.value`) %>% 
  filter(str_detect(term, params$methane_group)) %>% 
  mutate(group1 = "HMP", 
         group2 = "LMP",
    label = case_when(
    p < 0.001 ~ "p < 0.001",
    p > 0.1 ~ "p > 0.1",
    TRUE ~ glue("p = {round(p, 3)}")
    ),
    y.position = max(scores$PC2)*1.05)

boxplot_pc1 <- 
  scores %>% 
  ggplot(aes(x = !!sym(params$methane_group),
             y = PC1,
             group = !!sym(params$methane_group))) +
  geom_boxplot(aes(fill = !!sym(params$methane_group)),
               alpha = 0.5, 
               outlier.shape = NA) +
  geom_jitter(aes(fill = !!sym(params$methane_group),
                  color = !!sym(params$methane_group)),
              shape = 21,
              alpha = 0.5,
              height = 0,
              size = 1)+
  scale_fill_manual(values = study_colors$methane) +
  scale_color_manual(values = study_colors$methane) +
  # yes its on purpose we scale Y with X params (below there's a coord flip)
  scale_y_continuous(breaks = xbreaks,
                     limits = xlim) +
  guides(color = "none") +
  stat_pvalue_manual(
        test_pc1,
        label = "label",
        angle = 90,
        hjust = 1,
        vjust = -1,
        tip.length = -0.01,
        size = fontsize/3
      ) +
  coord_flip() +
  labs(x = "",
       y = "",
       fill = "") +
  paulaR::white_theme(font_size = fontsize) +
  theme(legend.position = "none",
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_blank(),
        axis.line = element_blank(),
        panel.border = element_blank(),
        plot.margin = margin(0,0,0,0),
        panel.grid = element_blank())



boxplot_pc2 <- 
  scores %>% 
  ggplot(aes(x = !!sym(params$methane_group),
             y = PC2,
             group = !!sym(params$methane_group))) +
  geom_boxplot(aes(fill = !!sym(params$methane_group)),
               alpha = 0.5, 
               outlier.shape = NA) +
  geom_jitter(aes(fill = !!sym(params$methane_group),
                  color = !!sym(params$methane_group)),
              shape = 21,
              alpha = 0.5,
              height = 0,
              size = 1)+
  scale_fill_manual(values = study_colors$methane) +
  scale_color_manual(values = study_colors$methane) +
  scale_y_continuous(breaks = ybreaks,
                     limits = ylim) +
  guides(color = "none") +
  stat_pvalue_manual(
        test_pc2,
        label = "label",
        tip.length = 0.01,
        size = fontsize/3
      ) +
  labs(x = "",
       y = "",
       fill = "") +
  paulaR::white_theme(font_size = fontsize) +
  theme(legend.position = "none",
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_blank(),
        axis.line = element_blank(),
        panel.border = element_blank(),
        plot.margin = margin(0,0,0,0),
        panel.grid = element_blank())



thin_pc2 <- ggplotGrob(boxplot_pc2)
thin_pc2$widths <- unit(rep(3, length(thin_pc2$widths)), "cm")

joint_pca <- (boxplot_pc1 + plot_spacer()) /
  (scores_plot + wrap_elements(thin_pc2)) +
  plot_layout(heights = c(1, 3),
              widths = c(3, 1),
              axes = "collect",
              ) &
  theme(plot.margin = margin(5,5,5,5),
          plot.tag.position = c(0, 1),    
    plot.tag = element_text(size = fontsize))  

joint_pca

}
```

# 4) Individual SCFA plots
## 4.1) Fecal propionate
```{r}
x <- params$methane_group
y <- "f_Propionate"
ylab <- "Propionate (μmol/g dry feces)"
ylim <- c(
  paulaR::trunc_signif(min(data[y], na.rm = T), 0),
  paulaR::ceiling_signif(max(data[y], na.rm = T), 0)
  )
ppos <- ylim[1] + ((ylim[2]-ylim[1]) * 0.99)
breaks <- seq(ylim[1], 100, 20)

test <- tibble(
  .y. = y,
  group1 = "LMP",
  group2 = "HMP",
  p = gc_res %>% filter(response == y, adjusted_by == ", (1|Subject_ID)") %>% pull(p),
  y.position = ppos,
) %>% 
  mutate(label = case_when(
    p < 0.001 ~ "p < 0.001",
    p > 0.1 ~ "p > 0.1",
    TRUE ~ glue("p = {round(p, 3)}")
  ))

f_prop_plot <- 
  data %>% 
     select(!!sym(x), !!sym(y)) %>% 
     filter(!is.na(!!sym(x)), !is.na(!!sym(y))) %>% 
     ggplot(aes(x = !!sym(x), 
                y = !!sym(y),
                fill = !!sym(x))) + 
     geom_boxplot(outlier.shape = NA,
                  color = "black") +
     geom_jitter(aes(fill = !!sym(x)), 
                 color = "black",
                 shape = 21,
                 size = 1,
                 height = 0,
                 stroke = 1.2,
                 alpha = 0.3) +
     scale_fill_manual(values = study_colors$methane) +
    stat_pvalue_manual(
        test,
        label = "label",
        tip.length = 0.01,
        size = fontsize/3
      ) +
     labs(y = ylab,
          x = "",
          ) +
    scale_y_continuous(limits = ylim,
                       breaks = breaks,
                       expand = expansion(mult = c(0.05, 0.15))) +
    paulaR::white_theme(font_size = fontsize) +
    theme(legend.position = "none",
          plot.title = element_text(size = fontsize - 2))

f_prop_plot
```

## 4.2) Fecal butyrate
```{r}
x <- params$methane_group
y <- "f_Butyrate"
ylab <- "Butyrate (μmol/g dry feces)"
ylim <- c(
  paulaR::trunc_signif(min(data[y], na.rm = T), 0),
  paulaR::ceiling_signif(max(data[y], na.rm = T), 0)
  )
ppos <- ylim[1] + ((ylim[2]-ylim[1]) * 0.99)
breaks <- seq(ylim[1], 30, 5)

test <- tibble(
  .y. = y,
  group1 = "LMP",
  group2 = "HMP",
  p = gc_res %>% filter(response == y, adjusted_by == ", (1|Subject_ID)") %>% pull(p),
  y.position = ppos,
) %>% 
  mutate(label = case_when(
    p < 0.001 ~ "p < 0.001",
    p > 0.1 ~ "p > 0.1",
    TRUE ~ glue("p = {round(p, 3)}")
  ))

f_but_plot <- 
  data %>% 
     select(!!sym(x), !!sym(y)) %>% 
     filter(!is.na(!!sym(x)), !is.na(!!sym(y))) %>% 
     ggplot(aes(x = !!sym(x), 
                y = !!sym(y),
                fill = !!sym(x))) + 
     geom_boxplot(outlier.shape = NA,
                  color = "black") +
     geom_jitter(aes(fill = !!sym(x)), 
                 color = "black",
                 shape = 21,
                 size = 1,
                 height = 0,
                 stroke = 1.2,
                 alpha = 0.3) +
     scale_fill_manual(values = study_colors$methane) +
    stat_pvalue_manual(
        test,
        label = "label",
        tip.length = 0.01,
        size = fontsize/3
      ) +
     labs(y = ylab,
          x = "",
          ) +
    scale_y_continuous(limits = ylim,
                       breaks = breaks,
                       expand = expansion(mult = c(0.05, 0.15))) +
    paulaR::white_theme(font_size = fontsize) +
    theme(legend.position = "none",
          plot.title = element_text(size = fontsize - 2))

f_but_plot
```


## 4.3) Fecal butyrate - distal colonic pH
```{r}
x <- "ph_distal_colon"
y <- "f_Butyrate"
ylab <- "Butyrate (μmol/g dry feces)"


model_data <- data %>%
   select(!!sym(x), !!sym(y), Subject_ID, !!sym(params$methane_group)) %>%
   group_by(Subject_ID) %>%
   summarize(!!sym(x) := mean(!!sym(x), na.rm = TRUE),
             !!sym(y) := mean(!!sym(y), na.rm = TRUE),
             !!sym(params$methane_group) := unique(!!sym(params$methane_group)))

ylim <- c(
  paulaR::trunc_signif(min(model_data[y], na.rm = T), 0),
  paulaR::ceiling_signif(max(model_data[y], na.rm = T), 0)+5
  )
breaks <- seq(ylim[1], ylim[2], 10)

xlim <- c(paulaR::trunc_signif(min(model_data[[x]], na.rm = T), 0)-0.3,
          paulaR::ceiling_signif(max(model_data[[x]], na.rm = T), -1)+0.3)

test <- tibble(
  estimate = cor_res %>% filter(response == y, predictor == x, adjusted_by == "") %>% pull(estimate),
  p = cor_res %>% filter(response == y, predictor == x, adjusted_by == "") %>% pull(p)
) %>% 
  mutate(p_lab = ifelse(
    p < 0.001,
    "p < 0.001",
    glue("p = {round(p, 3)}")
    ),
    r_lab = glue("β = {round(estimate, 3)}"),
    label = glue("{r_lab}, {p_lab}")
  )
  
    

f_but_dist_ph_plot <-
  model_data %>% 
  select(!!sym(x), !!sym(y), !!sym(params$methane_group)) %>% 
  na.omit() %>% 
    ggplot(aes(x = !!sym(x),
               y = !!sym(y))
           ) +
    geom_point(aes(fill = !!sym(params$methane_group)), 
               shape = 21,
               stroke = 1.2,
               color = "black",
               alpha = 0.6,
               size = 3) +
    scale_fill_manual(values = study_colors$methane) +
    geom_smooth(method = "lm", color = "black", se = F) +
    annotate("text",
              x = 5.1,
              y = 24, 
             label = test$label,
             size = fontsize/3,
             hjust = 0) +
    scale_x_continuous(limits = xlim
                       ) +
    scale_y_continuous(limits = ylim,
                       breaks = breaks
                       ) +
    labs(title = "",
         y = ylab,
         x = "Distal colonic pH",
         fill = ""
    ) +
    paulaR::white_theme(font_size = fontsize) +
    theme(legend.position = "bottom",
          plot.margin = margin(5,5,5,5),
          axis.text.x = element_text(angle = 0, hjust = 0.5))

f_but_dist_ph_plot
```

## 4.4) Plasma butyrate - prox colonic pH
```{r}
x <- "ph_proximal_colon"
y <- "p_Butyrate"
ylab <- "Plasma butyrate (μmol/mL)"


model_data <- data %>%
  select(!!sym(x), !!sym(y), Subject_ID, !!sym(params$methane_group)) %>%
  group_by(Subject_ID) %>%
  summarize(!!sym(x) := mean(!!sym(x), na.rm = TRUE),
            !!sym(y) := mean(!!sym(y), na.rm = TRUE),
            !!sym(params$methane_group) := unique(!!sym(params$methane_group)))

ylim <- c(
  paulaR::trunc_signif(min(model_data[y], na.rm = T), 0),
  paulaR::ceiling_signif(max(model_data[y], na.rm = T), 0)+0.5
  )
breaks <- seq(ylim[1], ylim[2], 1)

xlim <- c(4.7, 8.5) # comparable limits to previous plot

test <- tibble(
  estimate = cor_res %>% filter(response == y, predictor == x, adjusted_by == "") %>% pull(estimate),
  p = cor_res %>% filter(response == y, predictor == x, adjusted_by == "") %>% pull(p)
) %>% 
  mutate(p_lab = ifelse(
    p < 0.001,
    "p < 0.001",
    glue("p = {round(p, 3)}")
    ),
    r_lab = glue("β = {round(estimate, 3)}"),
    label = glue("{r_lab}, {p_lab}")
  )
  
    

p_but_ph_plot <-
  model_data %>% 
  select(!!sym(x), !!sym(y), !!sym(params$methane_group)) %>% 
  na.omit() %>% 
    ggplot(aes(x = !!sym(x),
               y = !!sym(y))
           ) +
    geom_point(aes(fill = !!sym(params$methane_group)), 
               shape = 21,
               stroke = 1.2,
               color = "black",
               alpha = 0.6,
               size = 3) +
    scale_fill_manual(values = study_colors$methane) +
    geom_smooth(method = "lm", color = "black", se = F) +
    annotate("text",
              x = 5.1,
              y = ylim[2], 
             label = test$label,
             size = 14/3,
             hjust = 0) +
    scale_x_continuous(limits = xlim
                       ) +
    scale_y_continuous(limits = ylim,
                       breaks = breaks
                       ) +
    labs(title = "",
         y = ylab,
         x = "Proximal colonic pH",
         fill = ""
    ) +
    paulaR::white_theme(font_size = 14) +
    theme(legend.position = "bottom",
          plot.margin = margin(5,5,5,5),
          axis.text.x = element_text(angle = 0, hjust = 0.5))

p_but_ph_plot
```

# 5) Extra plots
## 5.1) Plasma GLP1
```{r}
x <- params$methane_group
y <- "glp1"
ylab <- "GLP-1 (μmol/mL)"
ylim <- c(
  0, 
  paulaR::ceiling_signif(max(data[y], na.rm = T), 0)+1
  )
ppos <- ylim[1] + ((ylim[2]-ylim[1]) * 0.99)
breaks <- seq(ylim[1], ylim[2], 2)


test <- tibble(
  .y. = y,
  group1 = "LMP",
  group2 = "HMP",
  p = gc_res %>% filter(response == y, adjusted_by == ", (1|Subject_ID)") %>% pull(p),
  y.position = ppos,
) %>% 
  mutate(label = case_when(
    p < 0.001 ~ "p < 0.001",
    p > 0.1 ~ "p > 0.1",
    TRUE ~ glue("p = {round(p, 3)}")
  ))

p_glp1_plot <- 
  data %>% 
     select(!!sym(x), !!sym(y)) %>% 
     filter(!is.na(!!sym(x)), !is.na(!!sym(y))) %>% 
     ggplot(aes(x = !!sym(x), 
                y = !!sym(y),
                fill = !!sym(x))) + 
     geom_boxplot(outlier.shape = NA,
                  color = "black") +
     geom_jitter(aes(fill = !!sym(x)), 
                 color = "black",
                 shape = 21,
                 size = 1,
                 height = 0,
                 stroke = 1.2,
                 alpha = 0.3) +
     scale_fill_manual(values = study_colors$methane) +
    stat_pvalue_manual(
        test,
        label = "label",
        tip.length = 0.01,
        size = fontsize/3
      ) +
     labs(y = ylab,
          x = "",
          ) +
    scale_y_continuous(limits = ylim,
                       breaks = breaks,
                       expand = expansion(mult = c(0.05, 0.15))) +
    paulaR::white_theme(font_size = fontsize) +
    theme(legend.position = "none",
          plot.title = element_text(size = fontsize - 2))

p_glp1_plot
```

## 5.2) BMI
```{r}
meta_means_w_bmi <- meta_raw %>% 
  select(Subject_ID,
         !!sym(params$methane_group),
         sex, bmi, weight) %>% 
  distinct()

x <- params$methane_group
y <- "bmi"
ylab <- expression(BMI~(kg/m^2))
ylim <- c(
  paulaR::trunc_signif(min(meta_means_w_bmi[y], na.rm = T), 0),
  paulaR::ceiling_signif(max(meta_means_w_bmi[y], na.rm = T), 0)+1
  )
ppos <- ylim[1] + ((ylim[2]-ylim[1]) * 0.99)
breaks <- seq(ylim[1]+1, ylim[2]-1, 2)

test <- wilcox.test(
  as.formula(paste("bmi ~", x)),
  data = meta_means_w_bmi) %>%
  tidy() %>% 
  mutate(
  .y. = y,
  group1 = "LMP",
  group2 = "HMP",
  p = p.value,
  y.position = ppos,
) %>% 
  mutate(label = case_when(
    p < 0.001 ~ "p < 0.001",
    p > 0.1 ~ "p > 0.1",
    TRUE ~ glue("p = {round(p, 3)}")
  ))

bmi_plot <- 
  meta_means_w_bmi %>% 
     select(!!sym(x), !!sym(y)) %>% 
     filter(!is.na(!!sym(x)), !is.na(!!sym(y))) %>% 
     ggplot(aes(x = !!sym(x), 
                y = !!sym(y),
                fill = !!sym(x))) + 
     geom_boxplot(outlier.shape = NA,
                  color = "black") +
     geom_jitter(aes(fill = !!sym(x)), 
                 color = "black",
                 shape = 21,
                 size = 1,
                 height = 0,
                 stroke = 1.2,
                 alpha = 0.3) +
     scale_fill_manual(values = study_colors$methane) +
    stat_pvalue_manual(
        test,
        label = "label",
        tip.length = 0.01,
        size = fontsize/3
      ) +
     labs(y = ylab,
          x = "",
          ) +
    scale_y_continuous(limits = ylim,
                       breaks = breaks,
                       expand = expansion(mult = c(0.05, 0.15))) +
    paulaR::white_theme(font_size = fontsize) +
    theme(legend.position = "none",
          plot.title = element_text(size = fontsize - 2))

bmi_plot
```

## 5.3) Body weight
```{r}
x <- params$methane_group
y <- "weight"
ylab <- "Weight (kg)"
ylim <- c(
  paulaR::trunc_signif(min(meta_means_w_bmi[y], na.rm = T), 0),
  paulaR::ceiling_signif(max(meta_means_w_bmi[y], na.rm = T), 0)+1
  )
ppos <- ylim[1] + ((ylim[2]-ylim[1]) * 0.99)
breaks <- seq(ylim[1]+5, ylim[2]-4, 10)

test <- wilcox.test(
  as.formula(paste("weight ~", x)),
  data = meta_means_w_bmi) %>%
  tidy() %>% 
  mutate(
  .y. = y,
  group1 = "LMP",
  group2 = "HMP",
  p = p.value,
  y.position = ppos,
) %>% 
  mutate(label = case_when(
    p < 0.001 ~ "p < 0.001",
    p > 0.1 ~ "p > 0.1",
    TRUE ~ glue("p = {round(p, 3)}")
  ))

weight_plot <- 
  meta_means_w_bmi %>% 
     select(!!sym(x), !!sym(y)) %>% 
     filter(!is.na(!!sym(x)), !is.na(!!sym(y))) %>% 
     ggplot(aes(x = !!sym(x), 
                y = !!sym(y),
                fill = !!sym(x))) + 
     geom_boxplot(outlier.shape = NA,
                  color = "black") +
     geom_jitter(aes(fill = !!sym(x)), 
                 color = "black",
                 shape = 21,
                 size = 1,
                 height = 0,
                 stroke = 1.2,
                 alpha = 0.3) +
     scale_fill_manual(values = study_colors$methane) +
    stat_pvalue_manual(
        test,
        label = "label",
        tip.length = 0.01,
        size = fontsize/3
      ) +
     labs(y = ylab,
          x = "",
          ) +
    scale_y_continuous(limits = ylim,
                       breaks = breaks,
                       expand = expansion(mult = c(0.05, 0.15))) +
    paulaR::white_theme(font_size = fontsize) +
    theme(legend.position = "none",
          plot.title = element_text(size = fontsize - 2))

weight_plot
```

### 5.4) Join all (supl fig)
```{r}
glp1_w_bmi_plot <- (p_glp1_plot + labs(title = "GLP-1")) +
  (weight_plot + labs(title = "Weight")) + 
  (bmi_plot + labs(title = "Body mass index")) +
  plot_annotation(tag_levels = "A")

glp1_w_bmi_plot
```


# 6) Extra numbers for manuscript
```{r}
# fecal butyrate and propionate
gc_res %>%
  filter(type_response == "fecal", adjusted_by == ", (1|Subject_ID)") %>%
  select(response, estimate, p, FDR) %>%
  flextable()

# individual correlations - faecal butyrate and ph
cor_res %>% 
  filter(type_response == "fecal",
         type_predictor == "pH", 
         adjusted_by == "",
         predictor %in% c("ph_proximal_colon", "ph_distal_colon", "ph_sigmoid_colon")) %>%
  mutate(fdr = p.adjust(p, "BH")) %>%
  select(response, predictor, test, estimate, p, fdr) %>%
  mutate(estimate = round(estimate, 4),
         p = round(p, 4),
         fdr = round(fdr, 4)) %>%
  filter(str_detect(response, "Butyrate"))  %>%
  flextable()

# individual correlations - plasma butyrate and ph
 cor_res %>% 
   filter(type_response == "plasma",
          type_predictor == "pH", 
          adjusted_by == "",
          predictor %in% c("ph_proximal_colon", "ph_distal_colon", "ph_sigmoid_colon")) %>%
   mutate(fdr = p.adjust(p, "BH")) %>%
   select(response, predictor, test, estimate, p, fdr) %>%
   mutate(estimate = round(estimate, 4),
          p = round(p, 4),
          fdr = round(fdr, 4)) %>%
   filter(str_detect(response, "Butyrate"))  %>%
   flextable()
 
 # individual correlations - plasma butyrate and smithii
 cor_res %>% 
   filter(response == "p_Butyrate",
          adjusted_by == "",
          predictor %in% c("smithii_inv_ct")) %>%
   select(response, predictor, test, estimate, p, FDR) %>%
   mutate(estimate = round(estimate, 4),
          p = round(p, 4),
          FDR = round(FDR, 4)) %>%
   flextable()
```




# Save results

## Plasma + fecal SCFA heatmap
```{r}
format <- "png"
joint_fp_scfa_hm 
ggsave(glue("{fig_dir}/{params$markdown_id}_plasma_fecal_SCFA_GC_corr_heatmap.{format}"),
       bg = "white",
       width = 8,
       height = 6.5,
       dpi = 400
       )
saveRDS(joint_fp_scfa_hm,
        glue("{rds_dir}/{params$markdown_id}_plasma_fecal_SCFA_GC_corr_heatmap.rds"))
```

## Boxplots SCFA
```{r}
format <- "png" # png # pdf
f_prop_plot
ggsave(glue("{fig_dir}/{params$markdown_id}_fecal_propionate_boxplot.{format}"),
       bg = "white",
       width = 3,
       height = 3,
       dpi = 300)
saveRDS(f_prop_plot,
        glue("{rds_dir}/{params$markdown_id}_fecal_propionate_boxplot.rds"))

f_but_plot
ggsave(glue("{fig_dir}/{params$markdown_id}_fecal_butyrate_boxplot.{format}"),
       bg = "white",
       width = 3,
       height = 3,
       dpi = 300)
saveRDS(f_but_plot,
        glue("{rds_dir}/{params$markdown_id}_fecal_butyrate_boxplot.rds"))


p_glp1_plot
ggsave(glue("{fig_dir}/{params$markdown_id}_plasma_glp1_boxplot.{format}"),
       bg = "white",
       width = 3,
       height = 3,
       dpi = 300)
```

## Individual correlations
```{r}
format <- "png" # png # pdf
p_but_ph_plot
ggsave(glue("{fig_dir}/{params$markdown_id}_plasma_butyrate_ph_proximal_colon.{format}"),
       bg = "white",
       width = 4,
       height = 4,
       dpi = 300)
saveRDS(p_but_ph_plot,
        glue("{rds_dir}/{params$markdown_id}_plasma_butyrate_ph_proximal_colon.rds"))

f_but_dist_ph_plot
ggsave(glue("{fig_dir}/{params$markdown_id}_fecal_butyrate_ph_distal_colon.{format}"),
       bg = "white",
       width = 4.5,
       height = 4,
       dpi = 300)
saveRDS(f_but_dist_ph_plot,
        glue("{rds_dir}/{params$markdown_id}_fecal_butyrate_ph_distal_colon.rds"))
```

## Fecal SCFA PCA
```{r}
if (! params$is_mb_means){
  joint_pca + plot_annotation(tag_levels = "A")
  
  ggsave(glue("{fig_dir}/{params$markdown_id}_fecal_SCFA_PCA.{format}"),
       bg = "white",
       width = 8,
       height = 6,
       dpi = 400
       )
  saveRDS(joint_pca, glue("{rds_dir}/{params$markdown_id}_fecal_SCFA_PCA.rds"))

}
```

## GLP-1 - BMI - Weight (Supplementary)
```{r}
ggsave(glue("{fig_dir}/{params$markdown_id}_Sup_Fig_GLP1_Weight_BMI.{format}"),
       bg = "white",
       width = 8,
       height = 3,
       dpi = 400
       )
  saveRDS(glp1_w_bmi_plot, glue("{rds_dir}/{params$markdown_id}_boxplots_glp1_weight_bmi.rds"))
```


## Results tables - correlations and methane group comparison
```{r}
saveRDS(cor_res,
      glue("{rds_dir}/{params$markdown_id}_scfa_correlation.rds"))

write.xlsx(cor_res,
         glue("{tab_dir}/{params$markdown_id}_scfa_correlation.xlsx"))


saveRDS(gc_res,
      glue("{rds_dir}/{params$markdown_id}_scfa_methane_group_comparison.rds"))

write.xlsx(gc_res,
         glue("{tab_dir}/{params$markdown_id}_scfa_methane_group_comparison.xlsx"))
```



# Finalise
```{r}
sessionInfo()
```

